---
title: "Structure selection with an estimated penalty: selectEB"
author: "Simon Beyer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: false
    number_sections: true
    keep_tex: false
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath,amssymb}
vignette: >
  %\VignetteIndexEntry{Structure selection with an estimated penalty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, echo = TRUE)
```

# Problem

A model has $J$ candidate terms, each of which is either present or exactly
absent. Two cases share the same mathematics:

* **Mechanism.** A rate constant $k_j$ enters through a gate,
  $k_j = 10^{\ell_j} s_j$ (see `gateL1()`); absent means $s_j = 0$.
* **Differences between individuals.** A parameter of cell line $i$ is the
  reference value plus $\delta_i$; absent means $\delta_i = 0$.

Fitting all $2^J$ structures with a multistart each is out of reach beyond a
handful of candidates. An L1 penalty avoids the enumeration, but the problem
is not convex, and every penalty strength needs a multistart of its own.
`selectEB()` estimates the penalty with the structure, so a single multistart
of a single problem remains.

# Score of a structure

Let $z \in \{0,1\}^J$ be the structure, $D_z$ the minimal $-2\log L$ with the
absent candidates fixed at 0, $p_z$ the number of parameters of the present
candidates, and $D_0$ the value of the structure without candidates. All other
parameters (initial values, error model, rates that are certainly present) are
fitted in every structure.

**Slab.** The effects $\beta$ of the present candidates (the rate $k_j$, the
difference $\delta_i$) follow Zellner's g-prior,
$$\beta \mid z, g \sim N\!\left(0,\; g\, I_z^{-1}\right),$$
with $I_z$ the Fisher information of $\beta$. The prior is $g$ times as wide as
the uncertainty of the estimate, so $g$ is the squared effect size in units of
its standard error. It is dimensionless and does not change when a rate is
given in other units or fitted on another scale, which a slab on $\log k$ or
on $k$ with a fixed width cannot offer. A slab on $\log k$ also places mass
on rates too small to have an effect, so a present candidate at such a rate
costs nothing, and the selection no longer separates present from absent.

**Marginal likelihood.** For a linear model with known noise, integrating
$\beta$ gives exactly
$$-2\log m(y \mid z, g) = \frac{g\,D_z + D_0}{1+g} + p_z \log(1+g) + \text{const}.$$
For a nonlinear model the same expression follows from the Gaussian
approximation of the likelihood in $\beta$, with the likelihood ratio
$D_0 - D_z$ in place of the Wald statistic.

**Priors.** The structure prior integrates the share $\theta_c$ of present
candidates of family $c$ under a $\mathrm{Beta}(a, b)$ prior, which gives the
beta-binomial probability $p(z) = \prod_c B(n_c + a, J_c - n_c + b)/B(a,b)$.
With $a = 1$ and $b = J_c$ it favours sparse structures and corrects for the
number of candidates: a further null candidate does not raise the chance of a
false inclusion. The hyper-g prior $p(g) \propto (1+g)^{-a_g/2}$, $a_g = 3$,
keeps $g$ from growing with few present candidates. The score is
$$G(z, g) = \frac{g\,D_z + D_0}{1+g} + (p_z + a_g)\log(1+g) - 2\log p(z).$$

# Estimating the penalty

At fixed $z$, $G$ is minimal at
$$1 + \hat g = \frac{D_0 - D_z}{p_z + a_g},$$
and $\hat g = 0$ when the present candidates gain less than their number. This
is the M-step. The search over $z$ at fixed $g$ is the E-step in its mode
form: starting from a structure, every candidate is switched in random order,
each switch a fit warm-started from the current optimum, and a switch is kept
when it lowers $G$. Both steps alternate until neither lowers $G$; every
accepted step lowers it, so a start ends after finitely many fits.

The fits do not depend on $g$: one fit per structure and start suffices, and
the update of $g$ costs nothing. The best fit of a structure over all starts
enters its score, so starts profit from each other.

**Penalty per candidate.** A candidate with $p_j$ parameters is present iff its
gain $\Delta_j = D_{z,\text{off}} - D_{z,\text{on}}$ exceeds
$$\lambda_j = \frac{1+\hat g}{\hat g}\left(p_j \log(1+\hat g) -
  2\log\frac{p(z_\text{on})}{p(z_\text{off})}\right),$$
an L0 penalty whose strength is estimated: $\log(1+g)$ per parameter and the
correction for the number of candidates. For $g = n$ the first part is the
penalty of the BIC. The conditional inclusion probability is
$\mathrm{plogis}((G_\text{off} - G_\text{on})/2)$, and $\lambda_j$ is exceeded
iff it is above 0.5.

**Rule alpha.** With `rule = "alpha"` the score is $D_z + q\,p_z$ with
$q$ the $1-\alpha$ quantile of $\chi^2_1$. Each candidate then enters iff its
likelihood ratio test at level $\alpha$ rejects, so the error of the first
kind is $\alpha$ per candidate and up to $J\alpha$ over all of them.

# Multistart

Starts draw the structure (each candidate present with probability one half)
and the parameters around `center`; the first start has all candidates present
from the full optimum, the second none. The final scores are comparable across
starts because model, data and priors are the same. Batches of starts follow
until `control$hits` starts end within `control$tolHits` of the best.

```{r}
fit <- selectEB(obj, center, zero = attr(trafo, "gates"), fits = 10,
                control = list(hits = 3))
fit
plot(fit, type = "terms")
plot(fit, type = "waterfall")
```

# Final decision by refits

The candidates are ordered by their gain above the threshold, present ones
first, which for `rule = "pip"` is the order of the inclusion probabilities. The
structures along this order are refitted without priors, starting at the size
of the best structure, and tested against the full model by a likelihood ratio
test. The smallest structure the test does not reject at level `alpha` is
selected. With `lrt = "data"` the test uses the attribute `"data"` of the
objective value, so a prior term in the objective does not enter.

# Candidates that replace each other

In ODE models two candidates often have the same effect on the observables,
for example two parallel paths. The score handles this without special
treatment: both present gain as much as one, but cost twice, so one of them is
selected. Which one is arbitrary, and the result reports it in two ways:

* `alternatives` lists every evaluated structure within `control$window` of
  the best score. With `control$swaps = TRUE` every exchange of a present
  against an absent candidate is evaluated, so a substitute appears there.
* `nonidentifiable` lists groups of present candidates whose parameters are
  close to collinear at the best structure: eigenvectors of the correlation
  form of the profile Hessian with eigenvalue below `control$kappa`.

The score needs no determinant of the Hessian, so a singular direction does
not distort it.

# Limits

* $D_0$ is the structure without any candidate. If the candidates include
  terms that are certainly present, $D_0$ is far from all reasonable
  structures, $\hat g$ becomes large and the threshold for weak terms rises
  with $\log \hat g$. Terms that are certainly present belong in the model,
  not among the candidates.
* One $g$ applies to all candidates: strong and weak effects share one slab.
  A few strong effects raise the threshold for weak ones.
* The marginal likelihood is exact for linear models; for nonlinear ones it
  uses the likelihood ratio in a Gaussian approximation.
* The final refits test against the full model with many degrees of freedom
  and have less power than the score for a single weak candidate.

# Comparison with the L1 path

`scanL1()` minimises $D + \lambda \sum_j |\beta_j|$ over a grid of $\lambda$, with
a multistart per grid point or a continuation along the grid, and selects by
refits. `selectEB()` replaces the grid by the estimate of $g$, the soft
threshold by an exact zero, and the multistart per $\lambda$ by one multistart
over structures. The cost is about $J$ fits per sweep, a few sweeps per start.
