---
title: "Symmetries in dMod: structural identifiability, the observability-identifiability matrix, and symmetry reduction"
author: "Simon Beyer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: false
    number_sections: true
    keep_tex: false
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath,amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \usepackage{fvextra}
  # a Unicode-complete monospace font so the summary()/print() output renders its
  # generators verbatim: the sigma, eta and partial-derivative signs and the digit
  # subscripts (X_1, eta_i, d_i) are absent from the default Latin Modern Mono.
  - \usepackage{fontspec}
  - \setmonofont{DejaVu Sans Mono}[Scale=0.82]
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
vignette: >
  %\VignetteIndexEntry{Symmetries and structural identifiability in dMod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  echo     = TRUE
)
# symmetryDetection() is purely symbolic: it drives a Python module (sympy,
# numpy, scipy) through reticulate and needs no compiled model. Every
# computational chunk runs only if sympy is already present, so building the
# vignette never triggers a Python install.
haveSympy <- isTRUE(tryCatch(
  reticulate::py_module_available("sympy"),
  error = function(e) FALSE))
```

\setcounter{tocdepth}{3}
\tableofcontents
\newpage

# Introduction

A parameter is *structurally non-identifiable* when the model produces exactly the
same output for a whole family of parameter values, so that no amount of data and
no reduction of noise can tell those values apart. This is a property of the model
equations and the observation map alone. It is distinct from *practical*
non-identifiability, which depends on the data at hand; `symmetryDetection()`
addresses the structural question only.

For an ODE model with states $x \in \mathbb{R}^n$ and parameters
$\theta \in \mathbb{R}^q$,
\begin{equation}
\dot x(t) = f(x(t), \theta), \qquad x(0) = x_0(\theta),
\label{eq:ode}
\end{equation}
with observation function
\begin{equation}
g : (x, \theta) \longmapsto g(x, \theta)
\label{eq:obs}
\end{equation}
collecting the measured quantities, the question is whether two different unknown
vectors produce the same $g\big(x(t), \theta\big)$ for all $t$. If they do, the
data constrains only certain combinations of the unknowns.

dMod answers this through the `method` argument of `symmetryDetection()`.

* `method = "observability"` (the default) builds the *observability-identifiability
  matrix* and reads identifiability off its rank. It uses no ansatz and, for a
  rational model, is evaluated entirely in exact finite-field arithmetic, so it
  scales to large systems. With `reconstruct = TRUE` each non-identifiable
  direction comes back as an exact rational function of the parameters
  (Section \ref{sec-closedform}).
* `method = "polynomial"` searches for *continuous symmetries* via the polynomial
  Lie-symmetry ansatz and returns the explicit transformation that leaves the
  output invariant.
* `method = "scaling"` returns the scaling (toric) symmetries
  $z_i \mapsto \lambda^{c_i} z_i$ exactly, from an integer kernel. Pure integer
  linear algebra, so it scales, but it sees only the scaling part of the symmetry
  algebra (Section \ref{sec-scaling}).

Detection is half the job: a direction the data cannot constrain has to be removed
before fitting, and `symmetryReduction()` does that constructively
(Section \ref{sec-reduce}).

Sections \ref{sec-sym} and \ref{sec-oim} develop the mathematics of the two
methods, assuming only multivariable calculus. Section \ref{sec-supply} shows how
they consume the dMod pipeline objects, and Section \ref{sec-examples} works
through examples. Section \ref{sec-cycle} carries one model through the whole
route, from the determining equations by hand to the reduced parametrisation.

## Structural identifiability, local and global

Collect the unknowns into $z = (x_0, \theta)$ and write
$g(t\,; z) := g\big(x(t; z), \theta\big)$. Two unknown vectors are
*output-equivalent*, $z \sim z'$, when $g(t\,; z) = g(t\,; z')$ for all admissible
times and inputs; output-equivalent vectors are indistinguishable from data.

A component $z_i$ is *structurally globally identifiable* if, for generic $z$,
\begin{equation}
z \sim z' \;\Longrightarrow\; z_i = z_i' ,
\label{eq:sgi}
\end{equation}
and *structurally locally identifiable* if \eqref{eq:sgi} holds after restricting
$z'$ to a neighbourhood $U \ni z$,
\begin{equation}
z' \in U \ \text{and}\ z \sim z' \;\Longrightarrow\; z_i = z_i' .
\label{eq:sli}
\end{equation}
The model is identifiable when every component is. The two failure modes map onto
the two methods:

* A *continuous* family of output-equivalent vectors through $z$ is a local
  non-identifiability. Its tangent at $z$ is a null vector of the
  observability-identifiability matrix (Section \ref{sec-oim}); its explicit form
  along the family is the symmetry generator (Section \ref{sec-sym}).
* An *isolated* output-equivalent $z' \neq z$ (a sign flip, say) is purely global
  and has no infinitesimal signature. Neither method targets it: a full-rank
  observability matrix certifies *local* identifiability only.

# Symmetries and the determining equations {#sec-sym}

## The infinitesimal generator

The unknowns $z = (x_0, \theta) \in \mathbb{R}^{n+q}$ are treated as one augmented
system carrying $\dot\theta = 0$ alongside \eqref{eq:ode}. A *symmetry* is a
smooth, invertible map $z \mapsto \Phi_\varepsilon(z)$ with $\Phi_0 = \mathrm{id}$
that maps every solution of the dynamics to a solution and leaves the output
\eqref{eq:obs} unchanged. The canonical example trades a state against a
calibration parameter: if only $\alpha\,x_1$ is observed, doubling $\alpha$ and
halving $x_1$ reproduces the readout.

Working with the finite map is awkward; one differentiates at $\varepsilon = 0$
and keeps the *infinitesimals*
\begin{equation}
\eta(z) = \left.\frac{\partial}{\partial \varepsilon}
            \Phi_\varepsilon(z)\right|_{\varepsilon = 0},
\qquad
X = \sum_{i=1}^{n+q} \eta_i(z)\, \frac{\partial}{\partial z_i},
\end{equation}
the *infinitesimal generator*. For any smooth $h$,
\begin{equation}
X(h) = \sum_i \eta_i(z)\, \frac{\partial h}{\partial z_i}
\label{eq:lie-h}
\end{equation}
is the rate of change of $h$ along the family, the Lie derivative of $h$ along
$X$. The notation follows Merkt et al. (2015), whose algorithm is the
`"polynomial"` method.

## Recovering the transformation by integrating the generator {#sec-flow}

Both methods return $\eta$, not the finite map. The transformation is its flow,
\begin{equation}
\frac{d}{d\varepsilon}\,\Phi_\varepsilon(z) = \eta\big(\Phi_\varepsilon(z)\big),
\qquad \Phi_0(z) = z,
\label{eq:flow}
\end{equation}
that is $\Phi_\varepsilon = \exp[\varepsilon X]$ (Merkt et al., 2015, Theorem 1).

For a *scaling*, $\eta_i = c_i z_i$, \eqref{eq:flow} decouples and integrates at
once; with $\lambda = e^\varepsilon$,
\begin{equation}
z_i \;\longmapsto\; \lambda^{c_i}\, z_i ,
\label{eq:scaling-flow}
\end{equation}
which is why a scaling is reported by its integer weight vector $c$: the weights
*are* the generator. For a *general* direction the $\eta_i$ are rational and
\eqref{eq:flow} is a nonlinear coupled flow: still a one-parameter family through
$z$, a curved orbit rather than a ray, but its integral need not be elementary.
What the method delivers is always $\eta$, exactly.

## The determining equations

*Output invariance.* With $h = g$ in \eqref{eq:lie-h},
\begin{equation}
\sum_{i=1}^{n+q} \eta_i\, \frac{\partial g_{k'}}{\partial z_i} \;=\; 0 .
\label{eq:obs-cond}
\end{equation}

*Compatibility with the dynamics.* Transforming $z$ also transforms $\dot z$, so
$X$ is *prolonged*,
\begin{equation}
X' = \sum_i \eta_i\,\frac{\partial}{\partial z_i}
   + \sum_i \eta'_i\,\frac{\partial}{\partial \dot z_i},
\qquad
\eta'_i = \sum_j \dot z_j\, \frac{\partial \eta_i}{\partial z_j},
\label{eq:prolong}
\end{equation}
and the model admits $X$ iff $X'\cdot(\dot z_k - f_k) = 0$ on solutions
(Merkt et al., 2015, Theorem 2), that is
\begin{equation}
\sum_{j=1}^{n} f_j\, \frac{\partial \eta_k}{\partial z_j}
\;-\;
\sum_{i=1}^{n+q} \eta_i\, \frac{\partial f_k}{\partial z_i} \;=\; 0,
\qquad k = 1, \dots, n .
\label{eq:flow-cond}
\end{equation}
The first sum is the change of $\eta_k$ along $f$, the second the change of $f_k$
along $\eta$; their cancellation says $[f, X] = 0$, transforming and flowing
commute. It runs over the states alone because the parameters do not move under
the dynamics, while the second sum runs over all unknowns because $f$ depends on
the parameters.

Equations \eqref{eq:obs-cond} and \eqref{eq:flow-cond} are the *determining
equations*. They are linear in $\eta$, and the Lie bracket of two admitted
generators is admitted, so the solutions form a Lie algebra: a reported set of
directions is a *basis*, not a canonical list.

## Solving them: the polynomial ansatz

$\eta$ is an unknown *function*, so the system is closed by an ansatz. Merkt et
al. (2015) take a polynomial one, motivated by the elementary one-parameter groups
\begin{equation}
\begin{aligned}
z_i^\ast &= z_i + \varepsilon,
  &\quad X &= \partial_{z_i} &&\text{(translation)},\\
z_i^\ast &= e^{\varepsilon} z_i,
  &\quad X &= z_i\,\partial_{z_i} &&\text{(scaling)},\\
z_i^\ast &= \frac{z_i}{1 - \varepsilon\, z_i},
  &\quad X &= z_i^2\,\partial_{z_i} &&\text{(M\"obius)},
\end{aligned}
\label{eq:elementary}
\end{equation}
whose infinitesimals are all of the form $z_i^{\,d}$, hence
\begin{equation}
\eta_i(z_i) \;=\; \sum_{d = 0}^{d_{\max}} r_{i,d}\, z_i^{\,d} .
\label{eq:ansatz}
\end{equation}
Degree $d_{\max} = 1$ covers translations and scalings; higher degrees bring in
Möbius transformations and beyond. Letting $\eta_i$ depend on the other
coordinates as well replaces the single power by monomials of total degree
$\le d_{\max}$. The two are `polynomialControl(ansatz = "uni")` and `"multi"`,
with `"par"` the intermediate choice, and `pMax` $= d_{\max}$.

For rational $f$ and $g$, substituting the ansatz and clearing denominators gives
polynomial identities in $z$; each monomial coefficient is linear in the
$r_{i,d}$, so the system collapses to
\begin{equation}
C\,r \;=\; 0
\label{eq:linsys}
\end{equation}
in $N_r = (n+q)(d_{\max}+1)$ unknowns, solved in exact rational arithmetic
(modular elimination with rational reconstruction), so the rank decision is exact
rather than a floating-point threshold, and every surviving generator is verified
symbolically. Two sets of rows are added beyond the paper: the Lie chain
$X(L_f^{\,k} g) = 0$ up to `lieOrder`, and the resting field
$[f_{\mathrm{ss}}, X] = 0$ when the model is equilibrated.

# The observability-identifiability matrix {#sec-oim}

The default method takes a different route to the same information and needs no
ansatz. It is the local algebraic observability test of Sedoglavic (2002),
extended so that the unknowns include parameters and initial states, the FISPO
setting of Villaverde et al. (2016).

## Output derivatives and the rank criterion

At $t = 0$ the output is $g$ itself; its time derivative follows by the chain rule
and \eqref{eq:ode}, $\frac{d}{dt} g|_{t=0} = \sum_i \partial_{x_i} g\, f_i = L_F g$,
and iterating, the $k$-th derivative is $L_F^{\,k} g$. The sequence
\begin{equation}
\big(\, g,\; L_F g,\; L_F^2 g,\; \dots,\; L_F^N g \,\big)
\label{eq:summary}
\end{equation}
is the list of Taylor coefficients of the output at $t = 0$, everything the
output curve can reveal about the unknowns, and each entry is an explicit
function of $z$ (the classical power-series method, Pohjanpalo, 1978; used
numerically in SIAN, Hong et al., 2020).

Collect \eqref{eq:summary} into $\Phi: z \mapsto (g, L_F g, \dots, L_F^N g)$.
Output-equivalent vectors give the same $\Phi(z)$, so the unknowns are locally
determined exactly when $\Phi$ is injective near a generic point, which holds when
its Jacobian
\begin{equation}
\mathcal{O}(z) \;=\; \frac{\partial \Phi}{\partial z}
\;=\;
\begin{pmatrix}
\partial g / \partial z \\[2pt]
\partial (L_F g) / \partial z \\[2pt]
\vdots \\[2pt]
\partial (L_F^N g) / \partial z
\end{pmatrix}
\label{eq:oim}
\end{equation}
has full column rank $n + q$. That is the observability-identifiability matrix
(OIM) and the rank condition of Sedoglavic (2002). When the rank is deficient, the
null space of $\mathcal{O}$ collects the directions along which the output and all
its derivatives are unchanged. These are exactly the values $\eta(z)$ of the generators of
Section \ref{sec-sym} at $z$ (Massonis and Villaverde, 2020): the two methods
describe the same object, one as a tangent vector, the other as a field.

## The matrix on a small model

Take one species with a lumped influx and linear decay, measured directly,
\begin{equation}
\dot A \;=\; k_1 + u\, k_2 - k_d\, A, \qquad y \;=\; A,
\label{eq:toy-oim}
\end{equation}
with $u$ a known constant input and unknowns $z = (A_0, k_1, k_2, k_d)$. The
output and its derivatives are read straight off \eqref{eq:toy-oim},
\begin{align}
g          &= A, \nonumber\\
L_F g      &= k_1 + u\, k_2 - k_d\, A, \nonumber\\
L_F^2 g    &= -k_d\,(k_1 + u\, k_2) + k_d^2\, A,
\label{eq:toy-lie}
\end{align}
and differentiating in $z$ at $A = A_0$ gives the first three rows of the OIM,
\begin{equation}
\mathcal{O} \;=\;
\begin{pmatrix}
1            & 0     & 0     & 0 \\
-k_d         & 1     & u     & -A_0 \\
k_d^2        & -k_d  & -u\,k_d & -(k_1 + u\,k_2) + 2 k_d A_0
\end{pmatrix}.
\label{eq:toy-matrix}
\end{equation}
The two rates enter every derivative only through $K = k_1 + u\, k_2$, so
\eqref{eq:toy-matrix} has a one-dimensional null space spanned by
\begin{equation}
\eta_{k_1} = -u, \quad \eta_{k_2} = 1, \quad \eta_{A_0} = \eta_{k_d} = 0 ,
\label{eq:toy-null}
\end{equation}
which one checks row by row ($-u + u = 0$, $u k_d - u k_d = 0$). The higher
derivatives add no constraint because they too depend on the rates only through
$K$, which is why the Lie order saturates. $A_0$, $K$ and $k_d$ are identifiable.
Note that $\eta$ depends on the input value $u$: holding $u$ at one generic value
reports this trade-off, whereas measuring at two distinct values pins $k_1$ and
$k_2$ separately (Section \ref{sec-conditions}).

## Evaluation over a finite field {#sec-finitefield}

Carrying $\mathcal{O}$ symbolically is what makes the naive test intractable: the
entries $L_F^{\,k} g$ are rational functions whose numerators and denominators
grow with every Lie derivative, and the expression swell dominates the cost. The
remedy (Sedoglavic, 2002) is to evaluate $\mathcal{O}$ at a single generic point,
because the rank there equals the symbolic rank away from a measure-zero set.

dMod evaluates in exact arithmetic over $\mathrm{GF}(p)$. Two reasons: every
intermediate number stays in $\{0,\dots,p-1\}$, so there is no swell yet the
result is exact; and $p$ prime makes $\mathrm{GF}(p)$ a *field*, so every nonzero
residue is invertible, which is what row reduction needs. dMod uses primes just
below $2^{31}$, so products stay within a 63-bit integer.

For a rational model $\mathcal{O}$ is never written symbolically at all. Expanding
the solution as $x(t) = \sum_k x_k t^k$ and the right-hand side along it as
$f = \sum_k \phi_k t^k$, matching $\dot x = f(x)$ order by order gives
\begin{equation}
x_{k+1} = \tfrac{1}{k+1}\, \phi_k ,
\label{eq:taylor-rec}
\end{equation}
where $\phi_k$ depends only on $x_0,\dots,x_k$ because the series of a product is
a convolution $c_k = \sum_i a_i b_{k-i}$ and of a reciprocal the same convolution
rearranged. Running the operations that define $f$ on truncated series is
automatic differentiation in *Taylor mode*: no symbolic $L_F^{\,k} g$ is ever
formed, yet the coefficient of $t^k$ in $g(x(t))$ equals $L_F^{\,k} g / k!$. The
columns come from the same pass by carrying, next to each coefficient, its
gradient with respect to $z$ (forward-mode dual numbers nested inside the Taylor
recurrence); after the recurrence the gradient parts are the rows of
$\mathcal{O}$. The work is a few convolutions per operation and grows only
polynomially with model size, so chains that defeat a symbolic construction take a
fraction of a second.

Because the construction uses $+, -, \times, \div$ only, it is defined for
*rational* right-hand sides and observables; the one exception is a power or Hill
term $\mathrm{base}^{\,n}$ with a parameter exponent, recast to a rational
coordinate (Section \ref{sec-hill}). A logarithmic readout
$\log_{10}(h) + \mathrm{offset}$ carries the same information as the rational
$\mathrm{scale}\cdot h$, since a monotone transformation has the same level sets
and the additive offset plays the role of a multiplicative scale, and is supplied
in that form. An observable whose *gradient* is not rational either is outside both
methods.

One prime and one point already give the rank with probability one, but dMod
repeats the evaluation at several points and primes and takes the maximal rank;
agreement certifies it, since a non-generic coincidence at every point and prime
has probability zero. The points are deterministic, so runs are reproducible. The
support of each null vector names the unknowns taking part in the
non-identifiability; `reconstruct = TRUE` returns the directions as exact rational
functions instead (Section \ref{sec-closedform}). The price is that the verdict is
*local*, and a null vector gives the tangent rather than the finite
transformation; for that, use `"polynomial"`.

## Scaling symmetries and large models {#sec-scaling}

The dominant class of non-identifiability is scaling,
$z_i \mapsto \lambda^{c_i} z_i$, whose generator has the constant integer tangent
$c$. Both conditions of Section \ref{sec-sym} then become integer-linear: output
invariance forces every monomial exponent of the observation to be orthogonal to
$c$, and compatibility forces every monomial of $f_i$ to carry the weight of
$x_i$, that is $(b - e_i) \cdot c = 0$ for each exponent $b$ of $f_i$. Stacking
these and taking the integer kernel returns all scaling symmetries exactly, with
no point evaluation and no swell, hence `method = "scaling"` scales to large
models, at the price of seeing the scaling part of the algebra only, which in
practice is the bulk of identifiability problems (units, calibration, absolute
levels).

## Closed-form directions by multi-point reconstruction {#sec-closedform}

The rank verdict needs $\mathcal{O}$ at one point; the *functional* form of a
direction needs sampling. Over $\mathbb{Q}(z)$ the null space has a basis whose
entries are rational functions of $z$, and evaluating at one point yields one
*sample* of each. Reconstructing a function from samples is interpolation, done
here in exact arithmetic.

**A canonical sample.** At $z^\ast$ the reduced row echelon form of
$\mathcal{O}(z^\ast)$ over $\mathrm{GF}(p)$ splits the coordinates into pivot and
free columns. Each free column $f$ gives one null vector $v$ with $v_f = 1$ and
$v_c = -R_{cf}$ on the pivot coordinates. As long as the pivot pattern is generic,
the same free column yields the *same* rational functions $r_c(z) := v_c$ at every
point, so values from different points are samples of one function.

**Two reductions.** The scaling directions are peeled off first by the integer
kernel of Section \ref{sec-scaling}, so only the non-scaling residue is sampled.
Then each *entry* is probed for its relevant variables by perturbing each
coordinate in turn, and reconstructed over that small set only. Because the probe
is per entry, a direction may couple many coordinates while each entry stays
narrow, and the cost is set by the widest single entry.

**The per-entry fit.** Write $r_c(u) = \mathrm{num}(u)/\mathrm{den}(u)$ with total
degree at most $D$. Clearing the denominator gives, at every sample $u^{(s)}$,
\begin{equation}
\mathrm{num}\!\left(u^{(s)}\right)
\;-\; r_c\!\left(u^{(s)}\right)\,\mathrm{den}\!\left(u^{(s)}\right) \;=\; 0 ,
\label{eq:fit}
\end{equation}
*linear and homogeneous* in the unknown coefficients. Stacking over sample points
gives a matrix whose kernel is the coefficient vector, up to one common scale that
cancels in the ratio. Solving homogeneously, rather than fixing the constant term
of $\mathrm{den}$, is what lets the fit recover a denominator without constant
term. $D$ is raised from zero: below the true degree the fit is inconsistent, at
it the kernel is one-dimensional, and that minimal $D$ is taken; a few extra
sample points make a spurious low-degree fit fail \eqref{eq:fit}.

**Wide entries.** The dense fit costs
$\binom{n_{\mathrm{rel}}+D}{n_{\mathrm{rel}}}$ samples, prohibitive for an entry
coupling many variables, in practice the conserved-moiety cofactors
$-(x_1+\dots+x_k)/r$. Such an entry is a Laurent polynomial, so on a geometric
schedule its values form a linear-recurrence sequence and it is recovered by
Ben-Or-Tiwari interpolation (Ben-Or and Tiwari, 1988) from a handful of samples,
the work set by the number of terms. An entry with a genuine multi-term
denominator is first split by a univariate Cauchy fit along a ray, after which
numerator and denominator are ordinary sparse polynomials. Anything beyond the
degree caps falls back to its support rather than to a guess.

**Lifting $\mathrm{GF}(p)$ back to $\mathbb{Q}$.** A single prime returns the
coefficients only as residues. The Chinese Remainder Theorem combines residues
modulo $p_1, \dots, p_r$ into one value modulo $M = p_1\cdots p_r$, and *rational
reconstruction*, the extended Euclidean algorithm on $x$ and $M$, turns that
integer back into the unique fraction with
\begin{equation}
x \,\equiv\, n\, d^{-1} \pmod{M},
\qquad |n|,\ |d| \,\le\, \sqrt{M/2} .
\label{eq:ratrec}
\end{equation}
The bound is the point of using several primes: identifiability coefficients are
small, so once $\sqrt{M/2}$ exceeds them the lift is exact. dMod uses four primes
near $2^{31}$, giving $M \approx 2^{124}$ and a range up to $|n|,|d| \approx
2^{61}$.

**Certification.** Every reconstructed direction is re-evaluated at a fifth prime,
disjoint from the four used to build it, and checked against the null vector
there; one that fails is reported by its support with `explicit = FALSE` rather
than as a wrong closed form. No floating-point number enters anywhere. The caps
are the fields of `reconstControl()`, passed as
`symmetryDetection(..., control = reconstControl(...))`.

## Several experimental conditions {#sec-conditions}

Each condition $c$ fixes the externally controlled quantities (switches, dose) to
known values and may reparametrise a few rates, so it has its own $f^{(c)}$,
$g^{(c)}$ and $\mathcal{O}^{(c)}(z)$ over a shared set of unknowns. A direction is
invisible in the whole experiment exactly when it is invisible in every condition,
\begin{equation}
\mathcal{O}(z) \;=\;
\begin{bmatrix} \mathcal{O}^{(1)}(z) \\ \vdots \\ \mathcal{O}^{(K)}(z) \end{bmatrix},
\qquad
\ker \mathcal{O} \;=\; \bigcap_{c=1}^{K} \ker \mathcal{O}^{(c)} .
\label{eq:stacked}
\end{equation}
The intersection can be strictly smaller than any single kernel: a combination
hidden under one switch setting is often exposed by another.

This is also the correct treatment of a known switch. Holding a switch at one
generic symbol finds the directions valid for *that* generic value and can produce
regime-specific artefacts; setting it to its actual value per condition and
stacking removes them, and since the switch is then a constant rather than a
symbol of $z$, it cannot appear in any reconstructed direction. Conditions are
supplied as an [eventlist] with a condition grid.

## Several stimuli in time: piecewise segments {#sec-segments}

An event after $t_0$ breaks the analytic continuation of the output, so the
timeline is split at the distinct event times into segments $S_0, S_1, \dots$, one
local jet per segment, stacked exactly as in \eqref{eq:stacked}.

The subtlety is the seed of a later segment: building $S_j$'s jet needs the state
*and* its $z$-Jacobian at the boundary, i.e. the previous segment's state flowed
across the gap. That finite-time nonlinear flow is transcendental in the gap
length and is not a finite-field number. The exact device is to keep the gap
length a **formal power-series variable**: the flow of $S_{j-1}$ over its gap is
the state's Taylor series re-read as a series in $\Delta t_{j-1}$,
\begin{equation}
x(\tau_j^-) \;=\; \sum_{m\ge 0} x_m\,\Delta t_{j-1}^{\,m},
\label{eq:gapseries}
\end{equation}
whose coefficients the kernel already computes (with their $z$-duals). Every entry
of $\mathcal{O}$ becomes a power series in the gaps, and a constant direction $v$
is unobservable for *generic* timing exactly when $R_m v = 0$ for every
coefficient layer $R_m$. Those layers are simply stacked as extra rows: nothing
symbolic enters the linear algebra, $\Delta t$ only labels which layer a row came
from. The gap order is raised until the rank stops growing
(`info$gapOrderUsed`). This recovers parameters that enter only through a
transient between events, an inhibitor pre-incubation say, whose relaxed state
at the stimulus is propagated exactly across the gap.

## Verifying the rank: the saturation guard {#sec-verify}

The verdict rests on one heuristic: the Lie order is raised only until the rank
stops growing for a few orders. A rank can plateau and then climb again, a free
Hill exponent being the typical case, and stopping on a false plateau drops rows,
which only enlarges $\ker\mathcal{O}$: the failure mode is a silent
*over*-reporting.

`verify = TRUE`, the default, closes that gap. It re-uses the kernel and the
cached base point and simply extends the Lie order by `DMOD_SYM_VERIFY_MARGIN`
further orders (default six). Only the jet grows; the expensive
$f_{\mathrm{ss}} = 0$ solve is not repeated. If the rank is still climbing, the
guard FAILS and warns; if not, saturation is confirmed. `summary()` prints it:

```
saturation guard: PASSED (rank 233 stable through Lie order 18
  (6 orders beyond the reported saturation))
```

That rank is *not* the identifiability rank of the header line. The header's
`rank r / dim d` is the rank after projecting onto the $d$ unknowns, so $d - r$ is
the number of reported directions; the guard's rank is that of the raw kernel,
whose columns span the full jet across all stacked conditions and can be many
times $d$. Only its *constancy* certifies anything. For small models
`symEngine = "symbolic"` gives an independent cross-check over the exact
rational-function field, with no guard needed.

The guard is complementary to the per-direction certification of
Section \ref{sec-closedform}: that one rejects a wrong functional *form*, this one
checks the *number* of directions.

# Supplying the model {#sec-supply}

`symmetryDetection()` needs no compiled model; it takes the equations directly.

\begin{center}
\begin{tabular}{l p{0.66\textwidth}}
\toprule
argument & meaning \\
\midrule
\texttt{f} & the right-hand sides: an \texttt{eqnlist}, an \texttt{eqnvec}, or a named character vector \\
\texttt{g} & the observation functions, as an \texttt{eqnvec} or named character vector \\
\texttt{trafo} & an optional parameter transformation (parameter-named entries) and initial conditions (state-named entries) \\
\texttt{method} & \texttt{"observability"} (default), \texttt{"polynomial"} or \texttt{"scaling"} \\
\bottomrule
\end{tabular}
\end{center}

Anything `as.eqnvec` accepts is a valid `f`; with an `eqnlist` the conserved
moieties are reduced first (`reduceCQ`). Known constants are declared through
`fixed`, externally driven inputs through `forcings`, several conditions through
`events` and `conditions`, and the pre-stimulus steady state through a
steady-state `trafo` or `equilibrate`.

## Parameter transformations {#sec-trafo}

`trafo` is an [eqnvec] mapping each inner symbol to an expression in the outer
parameters, substituted symbolically into $f$ and $g$. Three uses occur
constantly: reparametrisation (a rate through its half-life); fixing, $k = 1$,
which deletes $k$ from the unknowns; and lumping, $k_{\mathrm{deg},m} =
k_{\mathrm{deg},p} = k_{\mathrm{deg}}$. A state-named entry is that state's
initial condition rather than a substitution, so a solved steady state from
[steadyStates] can be passed whole.

Structural identifiability is invariant under an invertible, smooth
reparametrisation: it neither creates nor removes a non-identifiability, only
changes its form. A scaling in natural parameters is a translation in
logarithmic ones. Both engines require rationality, so a `trafo` making $f$ or $g$
non-rational, such as $k = \exp(\log k)$, is inadmissible; a log-parametrised
model is analysed in its natural rational parameters, with identical conclusions.

## Steady-state constraints

Before the experiment a model often rests at a steady state, so $x_0$ is not free
but solves
\begin{equation}
f_{\mathrm{ss}}(x_0, \theta) = 0 ,
\label{eq:ssfield}
\end{equation}
which binds $x_0$ to $\theta$ and can only make the model more identifiable
(Saccomani et al., 2003). `"polynomial"` imposes this as one more commutation
condition $[F_{\mathrm{ss}}, X] = 0$. `"observability"` binds it through the
initial condition, in one of two ways: pass a closed form through state-named
`trafo` entries (`trafo = steadyStates(reactions)`), or set `equilibrate = TRUE`,
which puts the states on a numeric point of \eqref{eq:ssfield}, recomputed at each
evaluation point and prime.

Four things make the latter exact and cheap. The resting system is solved for the
*states only*, the rates held as symbolic coefficients, because solving for a rate would
express a globally shared rate through condition-specific quantities. States
occurring linearly are eliminated by Gaussian elimination and substituted, leaving
only the nonlinear core, which over $\mathrm{GF}(p)$ is triangulated by a Groebner
basis (grevlex plus FGLM, since a lexicographic basis is ruinous on a coupled
core). Degenerate branches where a species vanishes are removed by the
Rabinowitsch trick, and the back-substitution accepts a point only when every core
equation vanishes, so branches cannot be mixed. Finally, when the resting state is
algebraic of degree $d$ its coordinates lie in $\mathrm{GF}(p)$ only at some
points and primes, so generic points are drawn until one works for every
condition. The parameter sensitivities come from the implicit function theorem,
$J_x\,(\partial x_0/\partial\theta) = -J_\theta$, solved at that point.

The steady state is event-free; perturbations are applied on top of it. A switch
is a `forcing`, zero at rest; a dose is a `t0` event composed onto the resting
initial condition with its sensitivities. State initial conditions in `trafo` are
ignored under `equilibrate`.

## Free Hill and power exponents {#sec-hill}

A rate with a free exponent is not rational, and is handled by a recast (Savageau
and Voit, 1987): $\mathrm{base}^{\,n}$ becomes a coordinate $E$ with companion
$L = \log(\mathrm{base})$, carrying
\begin{equation}
\dot E = n\,E\,\frac{\dot{\mathrm{base}}}{\mathrm{base}}, \qquad
\dot L = \frac{\dot{\mathrm{base}}}{\mathrm{base}}, \qquad
\frac{\partial E}{\partial\,\mathrm{base}} = n\,\frac{E}{\mathrm{base}}, \qquad
\frac{\partial E}{\partial n} = E\,L .
\end{equation}
With $\mathrm{base}^{\,n}\to E$ the field is rational and $n$ enters the OIM
through $\partial E/\partial n = E\,L$, so it is assessed like any other
parameter. The logarithm is never evaluated: $\mathrm{base}$,
$\log(\mathrm{base})$ and $\mathrm{base}^{\,n}$ are algebraically independent, so
giving $L$ an independent generic residue yields the structural rank by the same
argument as Section \ref{sec-finitefield}.

*Without* `equilibrate` the base carries a free initial value, so $E$ and $L$ are
two further free-initial-value coordinates and the transient itself resolves the
exponent. *With* `equilibrate` the base is a solved steady-state coordinate,
\eqref{eq:ssfield} is solved with $E$ held generic, and the chain rule of $E$
folds into the steady-state Jacobian. (If the base has no linear turnover term in
its own balance the roles invert: $E$ is solved rationally and
$\mathrm{base} = E^{1/n}$, $L$ are held generic, with duals from the inverse chain
rule.)

A self-inhibited producer with a free Hill exponent, observed on an unknown scale
and analysed *without* a steady state, has its exponent identified:

```{r hill-transient, eval = haveSympy}
f <- eqnvec(p = "kpr / (1 + kinh*p^nhill) - dp*p")     # free Hill exponent nhill
res <- symmetryDetection(f, eqnvec(y = "s*p"), method = "observability",
                         reconstruct = TRUE, reduceCQ = FALSE)
c(rank = res$rank, dim = res$dim,
  nhill_identifiable = !("nhill" %in% res$symmetries[[1]]$support))
res$symmetries[[1]]$generator
```

Rank $5$ of $6$: `nhill` is identifiable and the lone direction is the weighted
scaling $\eta_{kinh} = nhill\cdot kinh$, $\eta_{kpr} = -kpr$, $\eta_p = -p$,
$\eta_s = s$, rational in `nhill` with no $\log$ term, since here the exponent
is pinned rather than freed.

With `reconstruct`, $E$ and $L$ are sampled as coordinates and back-substituted,
so a direction that involves the exponent carries an explicit $\log$ factor. A
feedback species `FB` throttling the production of `x` through
$K^{\,n}/(K^{\,n}+\mathrm{FB}^{\,n})$, with only a scaled `x` measured, shows both
regimes at once: the transient resolves the exponent, the resting state does not.

```{r hill-free, eval = haveSympy}
hill <- eqnlist() |>
  addReaction("0",  "FB", "k_pr_FB")                     |>
  addReaction("FB", "0",  "d_FB * FB")                   |>
  addReaction("0",  "x",  "k_pr_x * K^n / (K^n + FB^n)") |>
  addReaction("x",  "0",  "d_x * x")

# the modular engine assesses the free exponent n through the recast, like any
# parameter. Started from a free initial state, the transient resolves it: n is
# identifiable and the two directions left are ordinary scalings
res <- symmetryDetection(hill, eqnvec(xobs = "scale * x"),
                         method = "observability", equilibrate = FALSE,
                         reconstruct = TRUE, reduceCQ = FALSE)

c(identifiable = res$identifiable, dim = res$dim,
  n_confounded = "n" %in% unlist(lapply(res$symmetries, function(d) d$support)))

# started at rest, that information is gone and n trades against the Michaelis
# constant K with a log factor. The symbolic engine carries base^n as an atom, so
# a free exponent is native to it; it needs the steady state supplied.
ss <- eqnvec(FB = "k_pr_FB/d_FB",
             x  = "k_pr_x*K^n/(d_x*(K^n + (k_pr_FB/d_FB)^n))")
sres <- symmetryDetection(hill, eqnvec(xobs = "scale * x"),
                          method = "observability", symEngine = "symbolic", trafo = ss)
Filter(function(d) "n" %in% d$support, sres$symmetries)[[1]]$generator
```

The exponent direction has $\eta_n = 1$ and
\begin{equation*}
\eta_K = \frac{K}{n}\,\log\!\frac{k_{pr\_FB}/d_{FB}}{K},
\end{equation*}
so at rest the free exponent and the Michaelis constant cannot be told apart (the
resting feedback level $k_{pr\_FB}/d_{FB}$ is the other base), and the predicted
$\log(\mathrm{base})$ factor appears explicitly. `equilibrate = TRUE` returns the
same verdict from the recast alone. Entered linearly, the same
confounder would be an ordinary integer scaling; entered through a Hill term it
keeps the exponent in its weight.

# Removing a symmetry: reduction to invariants {#sec-reduce}

Detection names the free directions. Removing one is a change of parameters, and
`symmetryReduction()` constructs it.

It is the construction Lagrangian mechanics performs with generalised
coordinates: the configuration space is foliated, by constraint surfaces or by the
orbits of a symmetry, and one keeps the coordinates transverse to the leaves and
drops the directions along them. Here the space is the parameter space
$\mathbb{R}^m_{>0}$, the leaves are the symmetry orbits, and the transverse
coordinates are the *invariants*. Three consequences carry over. The count is
$\#q = \dim - \dim(\text{leaf})$, here the corank of the observability matrix,
whose kernel is tangent to the orbit. The choice is free up to any
reparametrisation $q \to Q(q)$, so the report offers an admissible family instead
of one pick. And a generalised coordinate need not inherit an old meaning:
$\theta$ replaces $(x, y)$ for the pendulum, while a free particle keeps its $x$.
Two things do not carry over. There is no dynamics on the $q$; an invariant is a
first integral of the flow of $X$, whose time is the group parameter $\varepsilon$
and not the model time $t$. And the section that picks one point per orbit is a
gauge fixing, so it must meet every orbit exactly once, which is what
Section \ref{sec-chart} has to establish.

## Invariants, and how many are needed

For a block of generators $X_1, \dots, X_r$ on $m$ coordinates, $I(z)$ is an
*invariant* when
\begin{equation}
X_a(I) \;=\; \sum_i \eta^{(a)}_i\, \frac{\partial I}{\partial z_i} \;=\; 0,
\qquad a = 1, \dots, r,
\label{eq:invariant}
\end{equation}
that is when $I$ is constant along every orbit. Output invariance
\eqref{eq:obs-cond} is \eqref{eq:invariant} for $I = g$: the observed quantities
are themselves invariants. That is why these are the right coordinates. They are
what the data pins down, and the orbit directions are what it cannot.

If the orbits through a generic point have dimension $k$, the rank of the
generator matrix there, then $m - k$ functionally independent invariants
coordinatise the orbit space. The reduction is that statement made constructive:

1. find a complete set $I_1, \dots, I_{m-k}$ (Sections \ref{sec-lattice} and
   \ref{sec-search});
2. make them the new parameters;
3. add $k$ equations that single out one point on each orbit, a *section*, and
   solve for the old coordinates (Section \ref{sec-chart}).

The composite is the emitted `trafo`, ready for [P()].

The naming follows one rule. A scaling leaves each surviving coordinate carrying
an invariant monomial, so it keeps its name. A curved carrier is an invariant
*value* and not the coordinate it was solved from, so every invariant of a curved
block gets a fresh parameter `q_<k>`: reusing a coordinate's name would suggest
substitutions that do not exist, a pinned entry `z = 1` fixing the inner
coordinate where `q_<k>` is free. The `Invariants:` block of the report states
which invariant each `q_<k>` holds.

## Scalings: an integer kernel {#sec-lattice}

For a scaling block everything is integer linear algebra. Stack the weights as a
matrix $W$, one row per direction, one column per coordinate. A Laurent monomial
$z^a = \prod_i z_i^{a_i}$ satisfies \eqref{eq:invariant} exactly when $W a = 0$,
so the invariant monomials are the integer kernel of $W$.

The section is as explicit. A *transversal* is a set $T$ of $k$ coordinates whose
block $A = W[\,\cdot\,, T]$ is invertible and for which $M = A^{-1} W$ is an
integer matrix, the integrality being what keeps each survivor's monomial Laurent.
Pinning $T$ leaves $z_j$ carrying $z_j \prod_{t \in T} z_t^{-M_{tj}}$. Since a
scaling orbit is a ray through the positive orthant, the pins may be *any*
positive values; $1$ is a convention and a known calibration value does as well.
The admissible $T$ are the bases of the matroid of the columns of $W$, and which
one to hold is a modelling decision, so the report lists the whole family.

## Curved directions: the search {#sec-search}

Beyond scalings there is no lattice. Every stage of the search rests on one
observation: \eqref{eq:invariant} is *linear in $I$*. Each is therefore the same
three steps over a richer class of candidates.

1. **Ansatz.** Write $I$ with undetermined coefficients over the class.
2. **Linear system.** Demand $X_a(I) = 0$ identically. By linearity that is a
   homogeneous system $C\,c = 0$, whose rows come from *evaluating* at generic
   points rather than from symbolic expansion, over $\mathrm{GF}(p)$ with the lift
   of Section \ref{sec-closedform}.
3. **Keep what is new.** Products and powers of invariants are invariants, so the
   null space is far larger than the number of independent invariants. A Jacobian
   rank test at a generic point keeps an independent subset, which a certificate
   such as `2 functionally independent of 9` reports.

The cheapest class is *Laurent monomials*: for $I = \prod_i z_i^{a_i}$, dividing
\eqref{eq:invariant} by $I$ gives
\begin{equation}
\sum_i a_i\, \frac{\eta^{(a)}_i}{z_i} \;=\; 0 ,
\label{eq:monomial-stage}
\end{equation}
linear in the exponent vector $a$, whose integer null space is the exponent
lattice. From there the ladder escalates:

* **Polynomials** of total degree up to `dPoly`. Every monomial coefficient of
  $X_a(I)$ is a linear form in the unknown coefficients: one row per monomial and
  generator. Generators with linear components preserve the total degree, so the
  degrees can be searched one at a time.
* **Separable characteristics.** If every $\eta_i$ involves no moved coordinate
  but $z_i$, the characteristic system
  $dz_1/\eta_1 = \cdots = dz_m/\eta_m$ (Olver, 1993) separates: each ratio
  integrates on its own as $G_i = \int dz_i/\eta_i(z_i)$, and the differences
  $G_i - G_j$ are invariants. No degree cap enters and the antiderivative may
  leave the Darboux language, $\eta = 1 + z^2$ giving $\arctan z$. It runs after
  the two ansatz stages, which answer in the form one wants to read,
  $k_1 + u\,k_2$ rather than $\exp(-(k_1 + u\,k_2)/u)$.
* **Rational invariants,** through a *Darboux polynomial*: a $P$ with
  $X(P) = \lambda P$ for a polynomial cofactor $\lambda$, which says the surface
  $\{P = 0\}$ is invariant. Cofactors add under multiplication, so a product of
  Darboux factors is invariant exactly when its cofactors cancel, and the
  Darboux-type rational invariants are the integer kernel of the cofactor matrix.
  Linear once more. The candidates come from the extactic determinant: every
  Darboux polynomial of degree at most $d$ divides the extactic determinant of the
  degree-$d$ monomial basis (Christopher, Llibre and Pereira, 2007), so one
  determinant and one factorisation deliver all of them.
* **Exponential factors** $\exp(g/h)$, with $h$ a product of Darboux factors and
  $\deg g \le$ `dExp`, each carrying its own cofactor into the same kernel. Here
  the escalation ends: by Prelle and Singer (1983) a vector field with an
  elementary first integral has one of Darboux form, so within the caps nothing
  expressible in closed form is missed.
* **Integrating factor,** for a planar block only: an $M$ with
  $\operatorname{div}(M X) = 0$, then a quadrature. Its value may fall outside the
  language of a parameter transformation, an $\arctan$ in the damped-oscillator
  case, and is then reported as an invariant with no reparametrisation claimed.

Before the ladder runs, the block is *module reduced*: over the field of rational
functions its generators may be combined with *function* coefficients, which can
collapse the support every later stage works in. After it, the count decides.
Pinning a coordinate with invariants still missing would discard identifiable
structure, so a block that ends short is reported `invariantOnly` with what it
found and the caps it exhausted.

## The chart: the old parameters as functions of the new {#sec-chart}

The invariants say what the new parameters *mean*. The `trafo` has to say the
opposite, what each old coordinate *is* as a function of them. That is one square
system, in four steps.

**1. Name.** Each invariant becomes one outer parameter,
\begin{equation}
I_1(z) = q_1, \qquad \dots, \qquad I_{m-k}(z) = q_{m-k},
\label{eq:chart-names}
\end{equation}
which is $m - k$ equations for the $m$ moved coordinates.

**2. Section.** The missing $k$ equations single out one point per orbit. A scaling
orbit is a ray and meets $\{z_i = 1\}$ once, so a constant pin suffices. A curved
orbit may never reach the pinned value, and may cross it twice and leave the solve
without a single branch. `symmetryReduction()` therefore searches for a *balance
section*, an equation $m_1 = m_2$ between two low-degree monomials in the moved
coordinates, and admits it only if the ratio $r = m_1/m_2$ is strictly monotone
along every orbit. That is a sign condition on
\begin{equation}
X(\log r) \;=\; \frac{X(m_1)\,m_2 - m_1\,X(m_2)}{m_1\,m_2},
\label{eq:section}
\end{equation}
checked symbolically: if the numerator's coefficients have one sign for every
generator then $r$ cannot return to a value it has taken, so the section is met
exactly once, inside the positive orthant. Mixed signs admit tangencies and are
dropped before any solve. With several gauge directions the *set* of sections must
be transversal to all of them, which a rank test on the values of
\eqref{eq:section} decides.

**3. Eliminate.** Equations \eqref{eq:chart-names} and the section are $m$
equations in $m$ unknowns, and the elimination is triangular. Each naming equation
is solved for one coordinate of its own, its *carrier*, no two invariants sharing
one. A carrier $v_l$ is admissible only if that single solve stays inside the
language a `trafo` may use: either $v_l$ enters the rational part of $I_l$ in the
degrees $\{0, d\}$ only, so that
$\mathrm{num} - q_l\,\mathrm{den} = \alpha\,v_l^{\,d} + \beta$ and $v_l$ follows
by a $d$-th root, with $d = 1$ giving a rational entry, or $v_l$ sits only inside
a single exponential factor, linearly in its argument, and follows by a logarithm.
Substituting the carriers into the section leaves the $k$ gauge coordinates as the
only unknowns; solving for them and back-substituting gives one entry per moved
coordinate. Where a carrier's solution still references another carrier, the
substitution is iterated until only outer symbols remain. The degree that survives
the substitution fixes the shape of the chart: linear leaves rational entries, a
quadratic produces a square root, and a radical denominator is cleared by
conjugate multiplication.

**4. Certify positivity.** Rates and concentrations are estimated on
$(0, \infty)$, so the chart is usable only if every entry is positive *for every
admissible value of the outer parameters*; otherwise a fit started in the wrong
corner leaves the model's domain with no sign that anything is wrong. Admissible is
the carrier's own domain: $(0, \infty)$ where the invariant is certified
sign-definite on the orthant, $\mathbb{R}$ otherwise. Each entry
is therefore proven sign-pure, numerator and denominator of one sign, and an entry
$a + c\sqrt{r}$ by the exact comparison of $a^2$ with $c^2 r$, which also pins the
branch of the root. The constant pin $z = 1$ goes through the same test instead of
sitting below it as a fallback, so it is emitted exactly where the orbit
demonstrably reaches it. If no candidate survives, nothing is emitted.

The four steps decide one geometric object, worth naming because it is easy to
confuse with its complement. The image of the emitted trafo is not an orbit; it is
the *section*. An orbit has dimension $k$ and carries points the data cannot tell
apart, while the image has dimension $m - k$ and carries one representative of
every orbit, so on it everything is distinguishable. The two are transverse and
their dimensions add up to $m$. Read that way the certificates are statements
about the image: monotonicity in step 2 says each orbit is met at most once,
positivity and coverage in step 4 say each is met at least once and inside the
positive orthant. Together the image is exactly the reachable part of the section
surface, a global cross-section of the map onto the orbit space.

**Offsets.** Positivity can fail for a reason no section repairs: the invariant
itself does not range over all of $(0, \infty)$, so no parameter free there can
carry it. The fix is a triangular change of the invariant basis, the outer
parameter carrying $I - \varphi$ with the offset $\varphi$ read off the offending
factor of a solved entry, either a constant or a function of the *other*
invariants. Two things are proven: $\varphi \ge 0$ for all positive outer values,
so the shifted entries stay positive, and *coverage*, $I - \varphi > 0$ on the
whole positive orthant with $\varphi$ composed back into inner coordinates, so the
chart still reaches every positive orbit. Coverage is what separates a sharp bound
from a lossy one: an offset read off a section that meets only part of the orbit
space also yields positive entries, and only coverage rejects it.
Section \ref{sec-egf} carries one through by hand.

**Real carriers.** An offset needs a finite bound. Where the invariant's value set
on the orthant is all of $\mathbb{R}$ there is none, and no reparametrisation
$\psi(I)$ inside the language repairs it: a rational $\psi$ that is defined on all of
$\mathbb{R}$, injective (hence strictly monotone, so that $\psi(I)$ still separates
orbits) and positive does not exist, since $\deg p \le \deg q$ forces equal limits at
$\pm\infty$ while $\deg p > \deg q$ forces $-\infty$ at one end or $+\infty$ at
both. The carrier is declared *real* instead, and the gauge follows, a constant pin
being reached only by the orbits that happen to cross it. With each entry written as a
polynomial in the pin coordinate $\tau$ over the invariants,
\begin{equation}
E_i(\iota, \tau) \;=\; \sum_{k \le n_i} a_{i,k}(\iota)\,\tau^{k},
\qquad a_{i,n_i} > 0 ,
\label{eq:pin-poly}
\end{equation}
every real root obeys $|\tau| \le 1 + \sum_k |a_{i,k}/a_{i,n_i}|$, and
$|x| \le (1 + x^2)/2$ turns that into a rational pin
\begin{equation}
\sigma(\iota) \;=\; 1 + \sum_i \sum_{k < n_i}
  \tfrac{1}{2}\Big(1 + \big(a_{i,k}/a_{i,n_i}\big)^{2}\Big)
\label{eq:sos-pin}
\end{equation}
above every root of every entry, where each $E_i$ carries the sign of its leading
coefficient. For an entry affine in $\tau$ this is $\sigma = 1 + \sum_i L_i^2$ with
$L_i$ its lower bound, since $1 + L^2 - L = (L - \tfrac12)^2 + \tfrac34 > 0$. The
chart then meets every orbit that meets the orthant, which `coverage` records.

A block can still come back with its invariants and no chart, and that is a
statement about the model. An invariant whose *sign* is not constant on the
positive orthant, a difference of rates that is a Darboux polynomial and so has a
fixed sign along every orbit, is a genuine boundary between components of the
orbit space; no chart over free *positive* parameters covers both sides, which is
what \eqref{eq:sos-pin} trades for a real one.

**The four steps on the phosphorylation cycle.** With the gain $s$ known, the
block has the two invariants \eqref{eq:cycle-I1} and \eqref{eq:cycle-I2} on the
three moved coordinates $(k_p, k_d, P_0)$, so one section is needed, and the
balance $k_d\,P = k_p\,pP$ is monotone along the orbit. Steps 1 and 2 give
\begin{equation}
k_p + k_d = \Lambda, \qquad
k_p\,(P_0 + pP_0) = C, \qquad
k_d\,P_0 = k_p\,pP_0 ,
\label{eq:chart-cycle}
\end{equation}
with $p_0 = pP_0$ untouched. The carriers are $k_p$ from the first equation and
$P_0$ from the second, which leaves $k_d$ as the gauge coordinate:
\begin{equation}
k_p = \Lambda - k_d, \qquad P_0 = \frac{C}{k_p} - p_0 .
\label{eq:chart-cycle-carriers}
\end{equation}
Substituting both into the section leaves one equation in $k_d$,
\begin{equation}
k_d\Big(\frac{C}{\Lambda - k_d} - p_0\Big) = (\Lambda - k_d)\,p_0
\qquad\Longleftrightarrow\qquad
k_d\,C = \Lambda\,p_0\,(\Lambda - k_d) ,
\label{eq:chart-cycle-gauge}
\end{equation}
linear in $k_d$ because the quadratic terms cancel, with the single root
$k_d = \Lambda^2 p_0/(\Lambda\,p_0 + C)$, the `transversal` line of the report.
Back-substituting into \eqref{eq:chart-cycle-carriers} gives
$k_p = \Lambda - k_d = \Lambda C/(\Lambda p_0 + C)$ and then
$P_0 = C/k_p - p_0 = C/\Lambda$, which is the chart \eqref{eq:cycle-trafo}. Every
entry is a ratio of sums of products of $\Lambda, C, p_0 > 0$, so step 4 passes
with no offset and the chart covers the whole reachable interval
\eqref{eq:cycle-reach}. Its image satisfies $k_d P_0 = k_p\,pP_0$ identically, one
codimension inside $(k_p, k_d, P_0, pP_0)$: the section surface, met by each orbit
once, and not an orbit itself.

Nothing in the chain is assumed. Every emitted invariant is verified
symbolically, $X(I) = 0$ for every generator, before the transformation is
assembled, and every entry is proven positive as in step 4. A stage that comes up
empty records instead what it excluded, so an unreduced block still says how many
invariants the caps produced and therefore which cap to raise; the full per-stage
record stays on the object in `$blocks[[i]]$certificates`.

# Examples {#sec-examples}

```{r lib, eval = haveSympy, message = FALSE}
library(dMod2)
```

## A gene-expression cascade at steady state

A stimulus `TF` drives transcription, the mRNA is translated into protein, both
species degrade at a shared rate, and only a scaled protein readout is measured.
Before stimulation the cell rests at the basal steady state.

```{r genex-model, eval = haveSympy}
eq <- eqnlist() |>
  addReaction("",     "mRNA", "ksyn * TF")   |>
  addReaction("mRNA", "",     "kdeg * mRNA") |>
  addReaction("",     "Prot", "ktrl * mRNA") |>
  addReaction("Prot", "",     "kdeg * Prot")

g <- eqnvec(yProt = "scale * Prot")          # readout of unknown gain
```

```{r genex-oim, eval = haveSympy}
res <- symmetryDetection(eq, g, method = "observability",
                         equilibrate = TRUE, reconstruct = TRUE,
                         verbose = FALSE)
summary(res)          # the report; printing `res` alone shows the directions only
```

At rest the protein sits at $\mathrm{scale}\cdot TF\,ksyn\,ktrl/kdeg^2$, so only
that product is observed: rank $1$ of $5$, four free directions, all exact integer
scalings. Three trade the numerator factors of the product; the fourth couples
$TF$, $ktrl$ and $kdeg$ with equal weight and is induced by the steady state,
where the degradation enters squared. Handing the steady state in explicitly with
`trafo = steadyStates(eq)` gives the identical verdict.

## Acting on a result: reparametrisation

A non-identifiable direction almost always means that a product or ratio of
parameters is identifiable while the factors are not. When a factor is
independently known, fix it:

```{r genex-fix, eval = haveSympy}
res2 <- symmetryDetection(eq, g, method = "observability",
                          equilibrate = TRUE, fixed = "scale",
                          reconstruct = TRUE)
c(rank = res2$rank, dim = res2$dim)
```

More often the product itself is the meaningful quantity. Carry it as one
parameter by fixing the redundant factor to one in a `trafo`:

```{r reparam, eval = haveSympy}
f <- eqnvec(x = "k1*k2 - d*x")            # only the product k1*k2 is identifiable
symmetryDetection(f, eqnvec(y = "x"), method = "observability")$identifiable

symmetryDetection(f, eqnvec(y = "x"), method = "observability",
                  trafo = eqnvec(k1 = "1", k2 = "k"))$identifiable   # now TRUE
```

The same recipe resolves any scaling direction: fix all but one of the involved
coordinates and let the survivor carry the product.

### Automated reduction

`symmetryReduction()` mechanises that and extends it beyond scalings:

```{r reduce, eval = haveSympy}
res <- symmetryDetection(eqnvec(m = "ktx - dm*m", p = "ktl*m - dp*p"),
                         eqnvec(y = "p"), method = "observability",
                         reconstruct = TRUE, verbose = FALSE)
red <- symmetryReduction(res)
red
symmetryDetection(eqnvec(m = "ktx - dm*m", p = "ktl*m - dp*p"), eqnvec(y = "p"),
                  method = "observability", trafo = red$trafo,
                  verbose = FALSE)$identifiable
```

`print()` shows only the verdict, the trafo and the invariants it is built from
(`q_1 = k_p + k_d` reads "the invariant $k_p + k_d$ is now the parameter `q_1`");
`summary()` adds one line per block, naming its kind, status, stage and how it
was gauged, and `summary(verbose = TRUE)` adds the admissible gauges. A
candidate fixing of your own goes in through `fixed = `. A transcendental
invariant may solve into `log`/`exp` entries, which `P()` accepts but the rational
round trip `symmetryDetection(trafo = )` does not. Directions the search cannot
resolve within its degree caps are reported with the caps they exhausted and the
remaining options: a structural assumption, a separating experiment, or prediction
profiles.

### A real-valued carrier

Only $a - b$ enters, and it takes both signs on the orthant, so the carrier is
declared real and the pin picks up \eqref{eq:sos-pin}:

```{r realcarrier, eval = haveSympy}
res <- symmetryDetection(eqnvec(x = "-(a - b)*x"), eqnvec(y = "x"),
                         method = "observability", reconstruct = TRUE,
                         verbose = FALSE)
symmetryReduction(res)
```

Both entries are positive for every real `q_1` (discriminant $1 - 4 < 0$), so every
positive $(a, b)$ is reached; the constant pin $b = 1$ would have reached only
$a - b > -1$.

## A phosphorylation cycle, worked through in full {#sec-cycle}

This model is small enough to solve in closed form yet carries both kinds of
direction: a scaling that trades amounts against the readout gain, and a curved
direction that splits the relaxation rate into its two halves. Everything is
derived by hand first and then confirmed. The numerical counterpart, what the
directions do to a profile likelihood, is `bench/symmetry_profile_flatness.R` in
the package sources.

A protein is phosphorylated at rate $k_p$ and dephosphorylated at rate $k_d$, and
only the phospho-form is measured, on an unknown gain $s$:
\begin{equation}
\dot P = -k_p\,P + k_d\,pP, \qquad
\dot{pP} = k_p\,P - k_d\,pP, \qquad
y = s\,pP .
\label{eq:cycle}
\end{equation}

```{r cycle-model, eval = haveSympy}
reactions <- eqnlist() |>
  addReaction("P",  "pP", "k_p*P",  "phosphorylation") |>
  addReaction("pP", "P",  "k_d*pP", "dephosphorylation")
observables <- eqnvec(y = "s*pP")
```

The unknowns are $z = (P, pP, k_p, k_d, s)$, $\dim z = 5$; an initial value is
carried under the name of its state, so `P` and `pP` *are* $P_0, pP_0$.

### What the data can see

The cycle conserves $T = P_0 + pP_0$, so the phospho-equation closes on its own,
$\dot{pP} = k_p T - \lambda\,pP$ with $\lambda = k_p + k_d$, and
\begin{equation}
pP(t) = pP_\infty + (pP_0 - pP_\infty)\,e^{-\lambda t},
\qquad
pP_\infty = \frac{k_p\,T}{\lambda} .
\label{eq:cycle-sol}
\end{equation}
The measured curve is a single exponential,
\begin{equation}
y(t) = s\,pP_\infty \;+\; s\,(pP_0 - pP_\infty)\,e^{-\lambda t},
\label{eq:cycle-y}
\end{equation}
which carries three coefficients. No experiment on this model can deliver more
than
\begin{equation}
\lambda = k_p + k_d,
\qquad
s\,pP_0,
\qquad
s\,pP_\infty = \frac{s\,k_p\,(P_0 + pP_0)}{k_p + k_d} .
\label{eq:cycle-obs}
\end{equation}
Five unknowns, three identifiable functions: the OIM must have rank $3$ and two
directions must be free. The next section recovers the same verdict from the
matrix, which applies equally when no closed form exists.

### The generators, from the observability matrix

Write $u = k_p P - k_d\,pP$ for the net flux, so $\dot P = -u$, $\dot{pP} = u$.
Differentiating $g = s\,pP$ along the flow,
\begin{equation}
g = s\,pP, \qquad L_F\,g = s\,u, \qquad L_F^2\,g = -\lambda\,s\,u ,
\label{eq:cycle-lie}
\end{equation}
because $\dot u = -\lambda\,u$: past the first derivative the flow only multiplies
by $-\lambda$, so $L_F^{\,k} g = (-\lambda)^{k-1} s\,u$ for $k \ge 1$. The rows of
$\mathcal{O}$ are the gradients of \eqref{eq:cycle-lie} in $z$,
\begin{equation}
\mathcal{O} \;=\;
\begin{pmatrix}
0 & s & 0 & 0 & pP\\[2pt]
s\,k_p & -s\,k_d & s\,P & -s\,pP & u\\[2pt]
-\lambda\,s\,k_p & \lambda\,s\,k_d & -s\,(u + \lambda P) & -s\,(u - \lambda\,pP)
  & -\lambda\,u
\end{pmatrix},
\label{eq:cycle-oim}
\end{equation}
and every further row is a combination of $d(s\,u)$ and $d\lambda$. The row space
is spanned by $d(s\,pP)$, $d(s\,u)$ and $d\lambda$ whatever the Lie order, the
rank saturates at $3$, and the null space is two-dimensional, the count read off
\eqref{eq:cycle-obs}, now without the solution. A vector $\eta$ lies in it exactly
when it annihilates the three differentials,
\begin{equation}
\eta_{k_p} + \eta_{k_d} = 0,
\qquad
s\,\eta_{pP} + pP\,\eta_s = 0,
\qquad
s\big(k_p\,\eta_P - k_d\,\eta_{pP} + P\,\eta_{k_p} - pP\,\eta_{k_d}\big)
  + u\,\eta_s = 0 .
\label{eq:cycle-null}
\end{equation}
Any basis of that solution space is an equally valid answer; the natural one
splits on whether the gain moves. Taking $\eta_s = -s$ gives $\eta_{pP} = pP$ and,
with $\eta_{k_p} = \eta_{k_d} = 0$, $\eta_P = P$:
\begin{equation}
X_1 = P\,\partial_P + pP\,\partial_{pP} - s\,\partial_s .
\label{eq:cycle-X1}
\end{equation}
Taking $\eta_s = 0$ forces $\eta_{pP} = 0$, leaves $\eta_{k_d} = -\eta_{k_p}$, and
with $\eta_{k_p} = -k_p$ gives $\eta_P = P + pP$:
\begin{equation}
X_2 = (P + pP)\,\partial_P + k_p\,\partial_{k_d} - k_p\,\partial_{k_p} .
\label{eq:cycle-X2}
\end{equation}

$X_1$ is a *scaling* with integer weights $(P\!:1,\; pP\!:1,\; s\!:-1)$: the
amounts enter \eqref{eq:cycle-obs} only through gain-times-amount products. The
$P$-component of $X_2$ is $P + pP$, not a multiple of $P$, so $X_2$ is *general*.
It holds $s$ and $pP_0$ and moves $(k_p, k_d, P_0)$, which carry only the two
constraints $\lambda$ and $k_p(P_0 + pP_0)$: a fast forward rate on a small
unphosphorylated pool cannot be told from a slow one on a large pool.

That split is a choice of reduction, not a taxonomy. $X_1$ is gauged by holding
any one of $P$, $pP$, $s$ at any value: its orbit is the ray
$(\mu P_0,\, \mu\,pP_0,\, \mu^{-1}s)$, which meets each of $\{P = c\}$,
$\{pP = c\}$, $\{s = c\}$ exactly once for every $c > 0$. $X_2$ has to be
reparametrised instead.

```{r cycle-detect, eval = haveSympy}
res <- symmetryDetection(reactions, observables, method = "observability",
                         reconstruct = TRUE)
```

Rank $3$ of $5$, and the two directions come back as derived: `X₁` under
`Scalings` with the weights of \eqref{eq:cycle-X1}, `X₂` under `General` with the
components of \eqref{eq:cycle-X2}. The `Reduction` block of `summary(res)` names
the two strategies.

### The finite transformations

For $X_1$ the flow is the power law \eqref{eq:scaling-flow}, with
$\mu = e^{\varepsilon}$:
\begin{equation}
P_0 \to \mu\,P_0, \qquad pP_0 \to \mu\,pP_0, \qquad s \to \mu^{-1}s .
\label{eq:cycle-flow1}
\end{equation}
For $X_2$ the flow decouples in the right order: $dk_p/d\varepsilon = -k_p$ gives
$k_p(\varepsilon) = e^{-\varepsilon}k_p$, which drives
$dk_d/d\varepsilon = k_p(\varepsilon)$, and $dP/d\varepsilon = P + pP$ integrates
with $pP$ fixed. With $\mu = e^{-\varepsilon}$,
\begin{equation}
k_p \to \mu\,k_p, \qquad
k_d \to k_d + (1-\mu)\,k_p, \qquad
P_0 \to \frac{P_0 + pP_0}{\mu} - pP_0 ,
\label{eq:cycle-flow2}
\end{equation}
with $pP_0$ and $s$ unchanged: $k_d$ absorbs what $k_p$ gives up so $\lambda$ is
held, and the total moves inversely to $k_p$ so $k_p(P_0 + pP_0)$ is held too.
Those are exactly the two constraints of \eqref{eq:cycle-obs} involving the moved
unknowns.

Since \eqref{eq:cycle-y} is available in closed form, the invariance can be
checked on the output directly, with no solver and no symbolics:

```{r cycle-flow}
ysol <- function(z, t) {                              # y(t) from (eq:cycle-sol)
  lambda <- z[["k_p"]] + z[["k_d"]]
  pPinf  <- z[["k_p"]] * (z[["P"]] + z[["pP"]]) / lambda
  z[["s"]] * (pPinf + (z[["pP"]] - pPinf) * exp(-lambda * t))
}
flowX1 <- function(z, mu) {                           # readout units
  z[["P"]] <- mu * z[["P"]]; z[["pP"]] <- mu * z[["pP"]]
  z[["s"]] <- z[["s"]] / mu
  z
}
flowX2 <- function(z, mu) {                           # split of the relaxation rate
  total      <- z[["P"]] + z[["pP"]]
  z[["k_d"]] <- z[["k_d"]] + (1 - mu) * z[["k_p"]]
  z[["k_p"]] <- mu * z[["k_p"]]
  z[["P"]]   <- total / mu - z[["pP"]]
  z
}

z <- c(P = 1, pP = 0.2, k_p = 0.3, k_d = 0.05, s = 1.7)
t <- seq(0, 20, by = 0.5)

max(abs(ysol(flowX1(z, 2.5), t) - ysol(z, t)))
max(abs(ysol(flowX2(z, 0.4), t) - ysol(z, t)))
```

`flowX2(z, 0.4)` turns $(k_p, k_d, P_0) = (0.3, 0.05, 1)$ into $(0.12, 0.23,
2.8)$, reversing which rate dominates, and the predicted curve does not move.

### Finding the invariants

Suppose the readout is calibrated, so $s$ is known. The fixing removes `X₁` and
leaves the curved `X₂`, whose moved coordinates are $P$, $k_p$, $k_d$ and whose
orbit is one-dimensional: $3 - 1 = 2$ invariants to find, each stage small enough
to do by hand.

*Laurent monomials.* For $I = P^{\alpha}\,pP^{\beta}\,k_p^{\gamma}\,k_d^{\delta}$,
\eqref{eq:monomial-stage} reads
\begin{equation}
\alpha\,\frac{P + pP}{P} - \gamma + \delta\,\frac{k_p}{k_d}
\;=\;
(\alpha - \gamma) + \alpha\,\frac{pP}{P} + \delta\,\frac{k_p}{k_d}
\;=\; 0 .
\end{equation}
The functions $1$, $pP/P$, $k_p/k_d$ are linearly independent, so
$\alpha = \gamma = \delta = 0$: no monomial invariant. That is the negative
result recorded before the search escalates.

*Degree one.* With $I = a\,P + b\,k_p + c\,k_d$ (coefficients in the untouched
$pP$, $s$ allowed), $X_2(I) = a\,P + a\,pP + (c - b)\,k_p$, so $a = 0$, $b = c$:
\begin{equation}
I_1 \;=\; k_p + k_d .
\label{eq:cycle-I1}
\end{equation}

*Degree two.* One invariant is still missing. $X_2$ maps the nine degree-two
monomials involving a moved coordinate to
\begin{equation}
\begin{aligned}
P^2      &\mapsto 2P^2 + 2P\,pP, &\quad
P\,pP    &\mapsto P\,pP + pP^2,  &\quad
P k_p    &\mapsto pP\,k_p, \\
P k_d    &\mapsto P k_d + pP\,k_d + P k_p, &\quad
pP\,k_p  &\mapsto -\,pP\,k_p, &\quad
pP\,k_d  &\mapsto pP\,k_p, \\
k_p^2    &\mapsto -2\,k_p^2, &\quad
k_p k_d  &\mapsto k_p^2 - k_p k_d, &\quad
k_d^2    &\mapsto 2\,k_p k_d ,
\end{aligned}
\end{equation}
and collecting $X_2(I)$ by monomial gives a three-dimensional solution space
spanned by $k_p\,(P + pP)$, $pP\,(k_p + k_d)$ and $(k_p + k_d)^2$. The last two
are a product and a square of quantities already known, so the independence test
keeps one new invariant,
\begin{equation}
I_2 \;=\; k_p\,(P_0 + pP_0) .
\label{eq:cycle-I2}
\end{equation}
With $I_1$ that is the corank $m - k = 2$, so the search stops at degree two and
the Darboux and exponential stages are never reached.

### The reduced model

```{r cycle-reduce, eval = haveSympy}
red <- symmetryReduction(res, fixed = "s")
summary(red)
```

The report has one block per coupled set of directions. `X₁` and `X₂` share the
coordinate `P`, so the scaling is not gauged on its own; it is carried along with
the curved direction, and fixing `s` removes it outright: that is the first
block, `fixed`. The second block is `X₂` alone, and its stages are the ones
computed above. No monomial invariant exists, and the polynomial stage returns
\eqref{eq:cycle-I1} and \eqref{eq:cycle-I2}, the two values `q_1` and `q_2`
carry under `Invariants:`. Write them $\lambda$ and $C$.

```{r cycle-roundtrip, eval = haveSympy}
red$trafo
symmetryDetection(reactions, observables, method = "observability",
                  trafo = red$trafo, fixed = "s", verbose = FALSE)$identifiable
```

The `trafo` maps the *outer* (estimated) coordinates to the *inner* (model) ones,
and the `now carries` lines say what the outer names mean: `q_1` holds
$\lambda = k_p + k_d$, `q_2` the flux $C = k_p(P_0 + pP_0)$, while `pP` keeps its
own name and its meaning. Those are the three quantities of \eqref{eq:cycle-obs}
with $s$ known, so the reduced model is parametrised by exactly what the data
delivers: three outer parameters where the model had four free coordinates
$(P_0, pP_0, k_p, k_d)$. The redundant direction is gone, not merely constrained.
Writing the outer values as $(\Lambda, C, p_0)$,
\begin{equation}
k_p = \frac{\Lambda\,C}{\Lambda\,p_0 + C},
\qquad
k_d = \frac{\Lambda^2\,p_0}{\Lambda\,p_0 + C},
\qquad
P_0 = \frac{C}{\Lambda},
\qquad
pP_0 = p_0 ,
\label{eq:cycle-trafo}
\end{equation}
and indeed $k_p + k_d = \Lambda$ and $k_p (P_0 + pP_0) = C$ identically.

### How far a curved orbit reaches

Parametrise the orbit by $k_p$; the invariants fix the rest,
$k_d = \lambda - k_p$ and $P_0 = C/k_p - pP_0$, and positivity of the two rates
and two amounts bounds
\begin{equation}
0 \;<\; k_p \;<\; \min\!\Big(\lambda,\ \frac{C}{pP_0}\Big)
 \;=\; \lambda\cdot\min\!\Big(1,\ \frac{pP_\infty}{pP_0}\Big) .
\label{eq:cycle-reach}
\end{equation}
Both walls mean something: at $k_p = \lambda$ the dephosphorylation has vanished,
at $k_p = C/pP_0$ the unphosphorylated pool has. Which binds is decided by the
direction of the measured curve: the rate wall for a rising readout
($pP_0 < pP_\infty$), the amount wall for a falling one.

```{r cycle-reach}
mu   <- c(0.05, 0.5, 1, 1.15, 1.2)         # k_p is multiplied by mu along the orbit
orbit <- t(vapply(mu, function(m) unlist(flowX2(z, m)), numeric(5)))
cbind(orbit[, c("k_p", "k_d", "P")],
      maxdev = apply(orbit, 1, function(r) max(abs(ysol(r, t) - ysol(z, t)))))
```

The last row has left the model's domain. The likelihood profiles follow: along
the orbit the output does not change at all, so the profile of any coordinate is
exactly flat over the interval its orbit reaches and can only rise where the walk
must leave the orbit. With $\lambda = 0.35$ here, $k_p$ and $k_d$ are flat below
$\lambda$, $P_0$ is flat above $pP_\infty - pP_0 \approx 0.83$, and $pP_0$, which
no direction moves, is the only coordinate with a proper parabola. Reparametrising
with `red$trafo` makes every profile a parabola.

The two walls are different objects, and the distinction is exactly the Darboux
bookkeeping of Section \ref{sec-reduce}. The surface $\{k_p = 0\}$ is *invariant*:
$X_2(k_p) = -k_p$ vanishes on it, so $k_p$ is a Darboux polynomial with cofactor
$-1$ and no orbit crosses it, so the profile of $k_p$ stays flat all the way down.
$\{k_d = 0\}$ is not invariant, since $X_2(k_d) = k_p \neq 0$ there; the orbit
crosses at finite $\varepsilon$ into $k_d < 0$, outside the model's domain. That
crossing, not the symmetry, is the wall: the flatness ends where the model does.

`symmetryReduction()` carries both walls, on request: `reportZeroCompatibility`
asks which coordinates the orbit can drive to $0$ with nothing running off to
infinity -- several at once allowed, since a set of rates switched off is still a
model. It is off by default: it costs one exact face solve per coordinate of each
block.

```{r cycle-zero, eval = haveSympy}
red <- symmetryReduction(res, fixed = "s", reportZeroCompatibility = TRUE)
red$zeroCompatibility[, c("coordinates", "verdict", "limit", "condition")]
```

`k_p` is excluded outright -- $C = k_p(P_0 + pP_0)$ is invariant and strictly
positive, so no orbit carries $k_p$ to zero at finite amounts, which is the
Darboux statement again. The other two are the two walls of
\eqref{eq:cycle-reach}, and `limit = FALSE` says both are met at finite
$\varepsilon$. Neither holds everywhere: `condition` is the comparison
\eqref{eq:cycle-reach} decides, $P_0 k_p > k_d\,pP_0$ for the rate wall and its
reverse for the amount wall -- a rising readout switches the dephosphorylation
off, a falling one empties the unphosphorylated pool. That is a property of the
point, not of the model, so it is stated as a condition and never as a verdict.
The reduction itself takes no parameter values; the condition is an R expression
over the model's own names, and one `eval()` settles it wherever it matters:

```{r cycle-zero-point, eval = haveSympy}
cond <- setNames(red$zeroCompatibility$condition, red$zeroCompatibility$coordinates)
sapply(cond[c("P", "k_d")], function(cc) eval(parse(text = cc), as.list(z)))
```

### The gauge section

The `Family` block closes the report with the gauge section `k_d*P = k_p*pP`. The
naive pin would have failed: by \eqref{eq:cycle-reach} $k_d$ never exceeds
$\lambda$, so for a slow cycle setting $k_d = 1$ is not a different chart but an
unreachable one. Substituting the orbit parametrisation into the section gives a
linear equation with the single root
\begin{equation}
k_p^{\ast} = \frac{\lambda\,C}{\lambda\,pP_0 + C} ,
\end{equation}
the first entry of \eqref{eq:cycle-trafo}, satisfying both $k_p^\ast < \lambda$
and $k_p^\ast < C/pP_0$: the representative lies strictly inside the reachable
interval whatever the invariants are. That is the monotonicity condition
\eqref{eq:section} in this instance.

```{r cycle-section}
lambda <- z[["k_p"]] + z[["k_d"]]
C      <- z[["k_p"]] * (z[["P"]] + z[["pP"]])
out    <- c(k_p = lambda, P = C, pP = z[["pP"]], s = z[["s"]])        # invariants

inner <- with(as.list(out), c(
  k_p = k_p * P / (k_p * pP + P),
  k_d = k_p^2 * pP / (k_p * pP + P),
  P   = P / k_p,
  pP  = pP,
  s   = s))
signif(inner, 4)
max(abs(ysol(inner, t) - ysol(z, t)))
```

Without the calibration, `symmetryReduction(res)` gauges both directions at once:
the block is then `{X₁, X₂}`, the invariant set gains $s\,pP$, and the section
becomes `k_d = pP, k_p = P`. The reduced model has three free parameters, matching
the three numbers of \eqref{eq:cycle-obs}.

## Closed-form directions and a known input

The components of a direction may be rational functions of a known input. Model
\eqref{eq:toy-oim} is the minimal case: declaring `u` with `fixed` keeps it out of
the unknowns yet lets it appear as a coefficient.

```{r toy-input, eval = haveSympy}
f   <- eqnvec(A = "k1 + u*k2 - kdeg*A")
toy <- symmetryDetection(f, eqnvec(y = "A"), method = "observability",
                         reconstruct = TRUE, fixed = "u")
Filter(function(d) d$type != "scaling", toy$symmetries)
```

The direction is the constant `k1 : u`, `k2 : -1`, the input-weighted slope
\eqref{eq:toy-null} up to sign. When such an input takes several known values
across conditions, stacking them separates `k1` and `k2` outright, and `u`, now a
constant of each condition, drops out of every reported direction.

## A pre-equilibrated model with a dose

Most experiments do not start from an arbitrary state: the system rests, a
stimulus is applied, and the relaxation is recorded. Let
\begin{equation}
\dot x \;=\; b - a\,x, \qquad y \;=\; s\,x,
\label{eq:ss-model}
\end{equation}
with basal production $b$, turnover $a$ and readout scale $s$, resting at
\begin{equation}
x_\ast \;=\; b/a
\label{eq:ss-fixed}
\end{equation}
until a dose sets the level to $D$ at $t = 0$ (a `replace` event), after which
\begin{equation}
x(t) \;=\; \frac{b}{a} + \Bigl(D - \frac{b}{a}\Bigr) e^{-a t},
\qquad y(t) = s\,x(t).
\label{eq:ss-traj}
\end{equation}
The Taylor coefficients at $t = 0$ are $sD$, $s(b - aD)$, $sa(aD - b)$, each a
multiple of the single relaxation mode with $a$ fixing the ratio between
successive coefficients. So $a$ is identifiable, and beyond it the data carry only
the two amplitude products $sD$ and $sb$. With $D$ unknown the OIM has rank $3$
and the null space is spanned by $b \mapsto -b$, $s \mapsto s$, $D \mapsto -D$:
the products are held but $s$ alone is not. With $D$ set by the experiment, $sD$
pins $s$, then $sb$ gives $b$, and the model is identifiable. The
pre-equilibration is what makes this clean, since a free initial value would add a
second non-identifiability.

```{r ss-dose, eval = haveSympy}
f <- eqnvec(x = "b - a*x")     # turnover; steady state b/a
g <- eqnvec(y = "s*x")         # scaled readout
dose <- addEvent(eventlist(), var = "x", time = 0, value = "dose",
                 method = "replace")

# a known dose: the relaxation identifies the rate, the scale and the production
symmetryDetection(f, g, method = "observability", events = dose,
                  conditions = data.frame(dose = 2,
                                          row.names = "stim"))$identifiable

# an unknown dose magnitude: the readout scale confounds with the amounts
res <- symmetryDetection(f, g, method = "observability", reconstruct = TRUE,
                         events = dose,
                         conditions = data.frame(row.names = "stim"))
res$symmetries[[1]]$generator
```

The direction is the exact integer scaling $b:1,\ \mathrm{dose}:1,\ s:-1$, peeled
from the integer kernel, so no reconstruction is needed. The same resting state
can be imposed without writing it down:

```{r ss-constraint, eval = haveSympy}
# rank 1 of 3: only the resting product s*b/a is seen, exactly as with the
# explicit steady state trafo = eqnvec(x = "b/a")
res <- symmetryDetection(f, g, method = "observability",
                         equilibrate = TRUE, reconstruct = TRUE)
c(identifiable = res$identifiable, rank = res$rank, dim = res$dim)

# a genuinely irrational steady state x* = sqrt(b/a): no rational form exists,
# yet the modular interior point still delivers the verdict
symmetryDetection(eqnvec(x = "b - a*x^2"), g, method = "observability",
                  equilibrate = TRUE)$rank
```

## A transient channel made identifiable by exact propagation

The gap propagation of Section \ref{sec-segments} identifies parameters that enter
only through a transient between events. A species $x$ is produced under a
feedback inhibition of strength $k_{\mathrm{inh}}$, switched on by an inhibitor
input, and read out on an unknown scale; a stimulus follows later. The inhibition
strength affects only the inhibitor-relaxed state at the stimulus onset.

```{r segments, eval = haveSympy}
f <- eqnvec(x = "kpr/(1 + kinh*inh) - kdeg*x + kstim*stim",
            inh = "0", stim = "0")
g <- eqnvec(y = "s*x")

# inhibitor and stimulus at the same time (no gap): kinh not identifiable
ev0 <- eventlist() |>
  addEvent(var = "inh",  time = 0, value = "1", method = "replace") |>
  addEvent(var = "stim", time = 0, value = "1", method = "replace")
r0 <- symmetryDetection(f, g, method = "observability", equilibrate = TRUE,
                        events = ev0, forcings = c("inh", "stim"))
"kinh" %in% unlist(lapply(r0$symmetries, function(d) d$support))   # TRUE

# inhibitor applied 30 units before the stimulus: the gap is propagated exactly
ev30 <- eventlist() |>
  addEvent(var = "inh",  time = -30, value = "1", method = "replace") |>
  addEvent(var = "stim", time = 0,   value = "1", method = "replace")
r30 <- symmetryDetection(f, g, method = "observability", equilibrate = TRUE,
                         events = ev30, forcings = c("inh", "stim"))
c(gapOrder = r30$info$gapOrderUsed,
  kinhIdentifiable = !("kinh" %in% unlist(lapply(r30$symmetries,
                                                 function(d) d$support))))
```

The second run reports a non-zero `info$gapOrderUsed`, the gap series carrying
$k_{\mathrm{inh}}$, and a higher rank than the first.

## An enzyme assay and its hidden symmetries

In the classical enzyme-kinetics assay the non-identifiabilities are not visible
by inspection yet can still be derived by hand:
\begin{equation}
\dot S = -\,\frac{k_{\mathrm{cat}}\,E_{\mathrm{tot}}\,S}{K_m + S},
\qquad y = s\,S,
\label{eq:mm}
\end{equation}
with unknowns $(S_0, E_{\mathrm{tot}}, K_m, s, k_{\mathrm{cat}})$, $\dim z = 5$.
Two substitutions leave the whole trajectory unchanged.

*The turnover-enzyme confounder.* The pair enters \eqref{eq:mm} only through
$V_{\max} = k_{\mathrm{cat}} E_{\mathrm{tot}}$, so
$k_{\mathrm{cat}} \to \lambda k_{\mathrm{cat}}$,
$E_{\mathrm{tot}} \to \lambda^{-1} E_{\mathrm{tot}}$ fixes everything observed:
weights $(k_{\mathrm{cat}}\!:1,\; E_{\mathrm{tot}}\!:-1)$.

*The readout-units confounder.* The right-hand side is jointly homogeneous of
degree one in the amounts $(S, K_m, E_{\mathrm{tot}})$, so rescaling them by $\mu$
and compensating $s \to \mu^{-1}s$ keeps $y = sS$: weights
$(S\!:1,\; E_{\mathrm{tot}}\!:1,\; K_m\!:1,\; s\!:-1)$.

Both are scalings, so the data pin exactly $sS$, $sK_m$ and $sV_{\max}$: three
functions, matching a rank of three out of five.

```{r mm, eval = haveSympy}
f <- eqnvec(S = "-kcat*Etot*S/(Km + S)")   # quasi-steady-state turnover
g <- eqnvec(y = "s*S")                      # readout of unknown gain s

res <- symmetryDetection(f, g, method = "observability", reconstruct = TRUE)
```

The engine returns two integer scalings spanning the same rank-$2$ lattice. The
printed basis may combine the two hand-derived confounders, since a null space
has no canonical basis, but the invariants it fixes are the same.

## Two rates on a curve

A non-identifiability need not be a straight ray. Gene expression observed only at
the protein level,
\begin{equation}
\dot m = k_{tx} - d_m\,m, \qquad
\dot p = k_{tl}\,m - d_p\,p, \qquad y = p,
\label{eq:gene}
\end{equation}
cannot separate transcription from translation: more mRNA poorly translated and
less mRNA well translated give the same protein. Scaling the hidden mRNA by any
$\lambda > 0$,
\begin{equation}
m \to \lambda\,m, \quad m_0 \to \lambda\,m_0, \quad
k_{tx} \to \lambda\,k_{tx}, \quad k_{tl} \to \lambda^{-1} k_{tl},
\label{eq:gene-sym}
\end{equation}
leaves $y = p$ invariant. Eliminating $\lambda$ confines the two rates to
$k_{tx}\,k_{tl} = \text{const}$, a hyperbola rather than a line: only the
product, which sets $p_\infty = k_{tx} k_{tl}/(d_m d_p)$, is identifiable. The
orbit is curved, but its generator is a plain scaling with weights
$(k_{tx}\!:1,\; k_{tl}\!:-1,\; m\!:1)$:

```{r gene, eval = haveSympy}
f <- eqnvec(m = "ktx - dm*m", p = "ktl*m - dp*p")   # mRNA hidden, protein seen
res <- symmetryDetection(f, eqnvec(y = "p"), method = "observability",
                         reconstruct = TRUE)
res$rank                                  # 5 of 6: one direction
res$symmetries[[1]]$generator
```

The $m\!:1$ weight records that the hidden mRNA scales with $k_{tx}$ along the
hyperbola. Degradation rates and the protein do not appear: the data fix them.

## EGF/EGFR feeding a MEK/ERK cascade {#sec-egf}

A larger but still hand-checkable network shows how to *read* a whole set of
directions at once. EGF binds its receptor to form an active complex that
phosphorylates MEK, phospho-MEK phosphorylates ERK, and each kinase is returned by
a constitutive phosphatase. Only the two phospho-forms are measured, each on an
unknown gain, so the ligand, the receptor and the unphosphorylated pools are
hidden. Every rate is mass-action, so every direction is rational.

```{r egf-model, eval = haveSympy}
reactions <- eqnlist() |>
  addReaction("EGF + EGFR", "EGF_EGFR", "k_bind * EGF * EGFR")   |>
  addReaction("EGF_EGFR", "EGF + EGFR", "k_unbind * EGF_EGFR")   |>
  addReaction("MEK", "pMEK", "k_phos_MEK * EGF_EGFR * MEK")      |>
  addReaction("pMEK", "MEK", "k_dephos_MEK * pMEK")              |>
  addReaction("ERK", "pERK", "k_phos_ERK * pMEK * ERK")          |>
  addReaction("pERK", "ERK", "k_dephos_ERK * pERK")
reactions <- customTotals(reactions, list(
  totalEGF  = "EGF + EGF_EGFR", totalEGFR = "EGFR + EGF_EGFR",
  totalMEK  = "MEK + pMEK",     totalERK  = "ERK + pERK"))

observables <- eqnvec(pMEK_obs = "scale_pMEK * pMEK",
                      pERK_obs = "scale_pERK * pERK")
```

`reduceCQ = FALSE` reports the directions in the species themselves rather than
through the moiety totals, which makes them easier to read:

```{r egf-run, eval = haveSympy}
egf <- symmetryDetection(reactions, observables, method = "observability",
                         reduceCQ = FALSE, reconstruct = TRUE)
```

Rank $11$ of $15$: four directions, three scalings and one rational confounder,
each forced by the partial observation.

* **Receptor-module units**
  $(EGF\!:\!1,\, EGFR\!:\!1,\, EGF\_EGFR\!:\!1,\, k_{bind}\!:\!-1,\, k_{phosMEK}\!:\!-1)$.
  The receptor module is never seen directly, so its concentration scale is free:
  multiplying the three species by $\lambda$ scales the bimolecular flux by
  $\lambda^2$, undone by $k_{bind}\!\to\!k_{bind}/\lambda$, and the MEK activation
  is held by $k_{phosMEK}\!\to\!k_{phosMEK}/\lambda$.
* **MEK amount against its readout gain**
  $(MEK\!:\!1,\, pMEK\!:\!1,\, k_{phosERK}\!:\!-1,\, scale_{pMEK}\!:\!-1)$.
* **ERK amount against its readout gain**
  $(ERK\!:\!1,\, pERK\!:\!1,\, scale_{pERK}\!:\!-1)$, the same trade-off at the
  foot of the cascade; nothing lies downstream, so no rate compensates.
* **The binding equilibrium** (general). Only $EGF$, $EGFR$ and $k_{unbind}$ move,
  with the complex and $k_{bind}$ held. Because the cascade reads the receptor
  module only through that complex, the dissociation rate and the split between
  free ligand and free receptor cannot be told apart.

The scalings integrate by \eqref{eq:scaling-flow} into power laws. The general
direction is the same construction with a non-constant $\eta$: writing
$C = EGF\_EGFR$ and $D = k_{bind}(EGF - EGFR)$, the flow \eqref{eq:flow} is
\begin{equation}
\frac{d\,EGF}{d\varepsilon} = \frac{-(EGF + C)}{D},\qquad
\frac{d\,EGFR}{d\varepsilon} = \frac{EGFR + C}{D},\qquad
\frac{d\,k_{unbind}}{d\varepsilon} = 1,
\end{equation}
with $C$, $k_{bind}$ and everything downstream constant. It has no elementary
closed form, but integrating it is unnecessary to see what it does: along the flow
\begin{equation}
\frac{d}{d\varepsilon}\big(k_{bind}\,EGF\,EGFR\big) = C,\qquad
\frac{d}{d\varepsilon}\big(k_{unbind}\,EGF\_EGFR\big) = C,
\end{equation}
so binding and unbinding fluxes move in lockstep and their difference, the net
rate of complex formation, is preserved. The orbit re-partitions free ligand and
receptor and shifts the dissociation rate while holding the active complex and
hence the entire measured cascade. That is exactly the non-identifiability the
support announced.

### Reducing it: the four steps in one report

The receptor-module direction is a scaling, so one of its coordinates may be
pinned at any positive value; doing that by hand with `trafo = eqnvec(EGF = "1")`
leaves the two readout scalings and the binding equilibrium, and the curved one
exercises all of Section \ref{sec-chart}.

```{r egf-reduce, eval = haveSympy}
egf1 <- symmetryDetection(reactions, observables, method = "observability",
                          reduceCQ = FALSE, reconstruct = TRUE,
                          trafo = eqnvec(EGF = "1"), verbose = FALSE)
red <- symmetryReduction(egf1)
summary(red)
```

The general block moves five coordinates, `EGFR`, `EGF_EGFR`, `k_bind`,
`k_phos_MEK` and `k_unbind`, along a one-dimensional orbit, so the search looks for
$5 - 1 = 4$ invariants. The polynomial stage already meets that count, but one of
its members changes sign on the positive orthant and no positive parameter can
carry it; the escalation therefore continues and the exponential-factor stage
supplies a sign-definite rational replacement, which is why the block reports
stage `exp` although the final set is entirely rational:
\begin{equation*}
I_1 = C\,k_{bind}, \quad
I_2 = \frac{k_{phosMEK}}{k_{bind}}, \quad
I_3 = k_{unbind} + k_{bind}(1 + EGFR), \quad
I_4 = \frac{EGFR\,C + EGFR + C}{C^2},
\end{equation*}
with $C = EGF\_EGFR$. Step 2 finds the balance $EGFR\cdot k_{bind} = k_{bind}$ between two
summands of $I_3$, reported in lowest terms as $EGFR = 1$ and certified monotone
along every orbit, as the fifth equation. Step 3 solves the
five jointly; the section makes one of them quadratic, hence the
$\sqrt{q_4 + 1}$, and step 4 certifies every entry positive for all positive
outer values, which is also what pins the branch of that root. $I_3$ needs an
offset, worked out below, after which its entry is the bare parameter,
`k_unbind = q_3`. Four outer parameters for five moved coordinates, one fewer than
the model started with.

**The elimination, by hand.** Write $C = EGF\_EGFR$. On the section $EGFR = 1$ the
fourth invariant collapses to $I_4 = (2C+1)/C^2$, so naming it $q_4$ gives a
genuine quadratic in its carrier,
\begin{equation*}
q_4\,C^2 - 2\,C - 1 = 0
\qquad\Longrightarrow\qquad
C = \frac{1 \pm \sqrt{1 + q_4}}{q_4} ,
\end{equation*}
and since $\sqrt{1+q_4} > 1$ only the $+$ branch is positive: step 4 does the
choosing, and the entry is `EGF_EGFR = (sqrt(q_4 + 1) + 1)/q_4`. The first
invariant then gives $k_{bind}$ by division, $k_{bind} = q_1/C$, whose radical
denominator clears by conjugate multiplication, the one rewrite `cancel()` cannot
do on its own:
\begin{equation*}
k_{bind} \;=\; \frac{q_1\,q_4}{1 + \sqrt{1+q_4}}
        \;=\; \frac{q_1\,q_4\,\big(\sqrt{1+q_4} - 1\big)}{(1+q_4) - 1}
        \;=\; q_1\big(\sqrt{1+q_4} - 1\big) ,
\end{equation*}
and $I_2 = k_{phosMEK}/k_{bind}$ multiplies it up to $k_{phosMEK} = q_2\,k_{bind}$.

The third invariant is where an offset becomes unavoidable. On the section
$I_3 = k_{unbind} + 2\,k_{bind}$, so naming it directly would give
$k_{unbind} = q - 2\,k_{bind}$, negative wherever $q$ is small: no parameter free
on $(0,\infty)$ can carry $I_3$. The offset is the AM--GM bound. With
$I_4 + 1 = (1+C)(EGFR+C)/C^2$,
\begin{equation*}
\varphi \;=\; 2\,I_1\big(\sqrt{I_4+1} - 1\big)
        \;=\; 2\,k_{bind}\Big[\sqrt{(1+C)(EGFR+C)} - C\Big] ,
\end{equation*}
so $I_3 - \varphi > 0$ reads
\begin{equation*}
k_{unbind} + k_{bind}\big[(1 + C) + (EGFR + C)\big]
\;>\; 2\,k_{bind}\sqrt{(1+C)(EGFR+C)} ,
\end{equation*}
which is AM--GM on the two bracketed factors, strict because $k_{unbind} > 0$ --
the coverage proof. AM--GM holds with equality exactly when the two factors agree,
that is at $EGFR = 1$, which is the section itself; there $\varphi = 2\,k_{bind}$
and $I_3 - \varphi$ is *exactly* $k_{unbind}$. The offset is therefore not merely
sufficient but sharp, which is why that entry comes out as the bare parameter,
`k_unbind = q_3`.

### The same network at rest, with a dose {#sec-egf-dose}

Michaelis--Menten rates, the states started at the resting state `steadyStates()`
solves for, and a unit dose of EGF at $t = 0$:

```{r egf-dose, eval = haveSympy}
reactionsMM <- eqnlist() |>
  addReaction("EGF + EGFR", "EGF_EGFR", "k_bind * EGF * EGFR")                |>
  addReaction("EGF_EGFR", "EGF + EGFR", "k_unbind * EGF_EGFR")                |>
  addReaction("MEK", "pMEK", "k_phos_MEK * EGF_EGFR * MEK / (Km_MEK + MEK)")  |>
  addReaction("pMEK", "MEK", "k_dephos_MEK * pMEK / (Km_pMEK + pMEK)")        |>
  addReaction("ERK", "pERK", "k_phos_ERK * pMEK * ERK / (Km_ERK + ERK)")      |>
  addReaction("pERK", "ERK", "k_dephos_ERK * pERK / (Km_pERK + pERK)")
rest <- eqnvec(
  k_phos_MEK = "k_dephos_MEK*pMEK*(Km_MEK + MEK)/(EGF_EGFR*MEK*(Km_pMEK + pMEK))",
  k_phos_ERK = "k_dephos_ERK*pERK*(ERK + Km_ERK)/(ERK*pMEK*(Km_pERK + pERK))",
  EGF        = "EGF_EGFR*k_unbind/(EGFR*k_bind)")
dose <- eventlist(var = "EGF", time = 0, value = "1", method = "add")

egf2 <- symmetryDetection(reactionsMM, eqnvec(pMEK_obs = "pMEK", pERK_obs = "pERK"),
                          method = "observability", events = dose, trafo = rest,
                          reduceCQ = FALSE, reconstruct = TRUE)
```

A `trafo` right-hand side names *initial values*, never the running states
(Section \ref{sec-trafo}), and here that distinction is load-bearing: the resting
state is written in the state symbols themselves, so reading them as states and
substituting $k_{phosMEK}$ back into its own equation gives, with
$C = EGF\_EGFR$,
\begin{equation}
\dot{pMEK} \;=\; k_{phosMEK}\,\frac{C\,MEK}{K_{MEK}+MEK}
              \;-\; k_{dephosMEK}\,\frac{pMEK}{K_{pMEK}+pMEK} \;\equiv\; 0 ,
\label{eq:ss-collapse}
\end{equation}
the cascade frozen and every Michaelis constant and phosphatase rate out of the
analysis with it. The dose is what makes the resting state informative: without it
the trajectory never leaves the steady state and the observation says no more than
$pMEK$ and $pERK$ themselves.

Rank $13$ of $14$, one direction, on the receptor module alone:
\begin{equation}
\eta_{EGFR} = 2\,EGFR\,(EGFR + C), \quad
\eta_{C} = C\,(EGFR + C), \quad
\eta_{k_{bind}} = -\,k_{bind}\,(EGFR + C),
\label{eq:egf-dose-gen}
\end{equation}
with $\eta_{k_{unbind}} = k_{bind}\,EGFR\,(1 - EGFR) + C\,k_{unbind}$. Two invariants
come straight off \eqref{eq:egf-dose-gen}, since $\eta_{k_{bind}}/k_{bind}$ is both
$-\tfrac12\,\eta_{EGFR}/EGFR$ and $-\,\eta_{C}/C$:
\begin{equation}
I_1 = EGFR\,k_{bind}^2, \qquad I_2 = C\,k_{bind} .
\label{eq:egf-dose-inv}
\end{equation}
The third moves $k_{unbind}$ and comes from the rational stage,
\begin{equation}
I_3 \;=\; \frac{EGFR^2 k_{bind} + EGFR\,k_{bind} + EGFR\,k_{unbind} + C\,k_{unbind}}
               {EGFR}
    \;=\; k_{bind}\,(EGFR + 1) \;+\; k_{unbind}\,\frac{EGFR + C}{EGFR} .
\label{eq:egf-dose-I3}
\end{equation}

Step 2 balances two summands of \eqref{eq:egf-dose-I3}; in lowest terms the section
is $EGFR = 1$, monotone since
$X(\log r) = -\,X(EGFR)/EGFR = -2\,(EGFR + C) < 0$. Step 4 needs an offset again,
and it is AM--GM once more, now between $EGFR$ and $1$:
$k_{bind}(EGFR + 1) \ge 2\,k_{bind}\sqrt{EGFR} = 2\sqrt{I_1}$, so the carrier holds
\begin{equation}
q_3 \;=\; I_3 - 2\sqrt{I_1}
    \;=\; k_{bind}\big(\sqrt{EGFR} - 1\big)^2 + k_{unbind}\,\frac{EGFR + C}{EGFR} ,
\label{eq:egf-dose-offset}
\end{equation}
a completed square plus a positive term, with infimum $0$ -- the range a parameter
free on $(0,\infty)$ needs. Equality in AM--GM holds exactly at $EGFR = 1$, the
section itself, so the offset is sharp there too. It is a function of $I_1$ alone,
which makes $X(q_3) = X(I_3) - 2\,X(\sqrt{I_1}) = 0$ immediate. The chart then
follows from \eqref{eq:egf-dose-inv} on the section,
\begin{equation}
EGFR = 1, \quad k_{bind} = \sqrt{q_1}, \quad C = \frac{q_2}{\sqrt{q_1}}, \quad
k_{unbind} = \frac{\sqrt{q_1}\;q_3}{\sqrt{q_1} + q_2} ,
\label{eq:egf-dose-chart}
\end{equation}
the last entry because $q_3 = k_{unbind}(1 + C)$ there while
$\sqrt{q_1} + q_2 = k_{bind}(1 + C)$.

```{r egf-dose-reduce, eval = haveSympy}
summary(symmetryReduction(egf2))
```

# References

Ben-Or, M., and Tiwari, P. (1988). A deterministic algorithm for sparse
multivariate polynomial interpolation. In *Proceedings of the 20th Annual ACM
Symposium on Theory of Computing (STOC)*, 301--309.

Christopher, C., Llibre, J., and Pereira, J. V. (2007). Multiplicity of invariant
algebraic curves in polynomial vector fields. *Pacific Journal of Mathematics*
229, 63--117.

Hong, H., Ovchinnikov, A., Pogudin, G., and Yap, C. (2020). Global
identifiability of differential models. *Communications on Pure and Applied
Mathematics* 73, 1831--1879.

Massonis, G., and Villaverde, A. F. (2020). Finding and breaking Lie symmetries:
implications for structural identifiability and observability in biological
modelling. *Symmetry* 12, 469.

Merkt, B., Timmer, J., and Kaschek, D. (2015). Higher-order Lie symmetries
in identifiability and predictability analysis of dynamic models.
*Physical Review E* 92, 012920.

Olver, P. J. (1993). *Applications of Lie Groups to Differential Equations*,
2nd edition. Springer, New York.

Pohjanpalo, H. (1978). System identifiability based on the power series
expansion of the solution. *Mathematical Biosciences* 41, 21--33.

Prelle, M. J., and Singer, M. F. (1983). Elementary first integrals of
differential equations. *Transactions of the American Mathematical Society* 279,
215--229.

Saccomani, M. P., Audoly, S., and D'Angiò, L. (2003). Parameter identifiability
of nonlinear systems: the role of initial conditions. *Automatica* 39, 619--632.

Savageau, M. A., and Voit, E. O. (1987). Recasting nonlinear differential
equations as S-systems: a canonical nonlinear form. *Mathematical Biosciences*
87, 83--115.

Sedoglavic, A. (2002). A probabilistic algorithm to test local algebraic
observability in polynomial time. *Journal of Symbolic Computation* 33,
735--755.

Villaverde, A. F., Barreiro, A., and Papachristodoulou, A. (2016).
Structural identifiability of dynamic systems biology models. *PLoS
Computational Biology* 12, e1005153.
