---
title: "PEtab problems in dMod2"
author: "Simon Beyer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
    keep_tex: false
    latex_engine: lualatex
bibliography: refs.bib
csl: ieee.csl
link-citations: true
header-includes:
  - \usepackage{fontspec}
  - \usepackage{unicode-math}
  - \setmainfont{CMU Serif}
  - \setsansfont{CMU Sans Serif}
  - \setmonofont{CMU Typewriter Text}
  - \setmathfont{NewCMMath-Regular.otf}
  - \usepackage{mathtools}
  - \usepackage{booktabs}
  - \usepackage{fancyvrb}
  - \fvset{fontsize=\small}
  - \usepackage{tikz}
  - \usetikzlibrary{arrows.meta, positioning, calc}
  - \usepackage{graphicx}
  - \usepackage{microtype}
  - \usepackage{etoolbox}
  - \AtBeginEnvironment{longtable}{\small}
  - \AtBeginDocument{\let\OldTOC\tableofcontents\renewcommand{\tableofcontents}{\OldTOC\clearpage}}
  - \setcounter{secnumdepth}{2}
fontsize: 10pt
---

```{r setup, include=FALSE}
options(width = 70)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      collapse = TRUE, comment = "#>",
                      dpi = 400, fig.width = 10, fig.height = 3)
library(dMod2)
have_libsbml <- isTRUE(tryCatch(
  reticulate::py_module_available("libsbml"), error = function(e) FALSE))
knitr::opts_chunk$set(eval = have_libsbml)
```

# Overview

PEtab [@schmiester2021petab] is a file format for parameter estimation
problems in systems biology: a YAML manifest, an SBML model
[@hucka2003sbml] and a set of TSV tables. dMod2 imports such a problem as a
fully assembled fit problem in one call and exports dMod problems in the
same format.

PEtab refers to v2 unless stated otherwise. Version 1 problems are read as
well and normalised onto the same internal shape.

# PEtab objects and their dMod2 counterparts

dMod2 composes a fit problem from function objects, `g * x * p`, closed by
an objective [@kaschek2019dmod]. The importer fills these slots:

| PEtab | dMod2 |
|---|---|
| SBML reactions, rules, events | `reactions` (an `eqnlist`), `odemodel`, `x = Xs(odemodel)` |
| `observables.tsv` formulas | `g = Y(observables)` |
| `observables.tsv` noise formulas | `e`, the error model, or `NULL` when every noise formula is constant |
| `parameters.tsv`, `conditions.tsv` | `p = P(trafo)`, one branch per condition |
| `measurements.tsv` | `dataList`, a `datalist` keyed by condition |
| `experiments.tsv` | simulation start times and mid-run events |
| `parameters.tsv` priors | a prior term added to the objective |

PEtab declares a scale per parameter (`lin`, `log`, `log10`). dMod2 wraps
`p` so that the optimiser sees each parameter on its scale and the chain
rule carries through the composition; `bestfit` is therefore on the PEtab
scale, not the linear one.

# Importing a problem

```{r import-recipe, eval=FALSE}
petab <- importPEtab("problem.yaml", backend = "cppDE")
```

`importPEtab()` parses the manifest, reads the SBML through libsbml,
generates and compiles the right-hand side with its sensitivity equations,
and returns an object of class `petabproblem`.

| argument | meaning |
|---|---|
| `yamlPath` | path to the YAML manifest |
| `backend` | `"cppDE"` (default, handles events and second-order sensitivities) or `"deSolve"` |
| `compile` | `FALSE` reuses an already compiled model of the same name |
| `cores` | conditions evaluated in parallel inside one objective call |
| `modelname` | base name for the generated sources |
| `outdir` | where the generated sources and the shared object are written |
| `optionsOde`, `optionsSens` | solver settings, forwarded to `Xs()` |

`outdir` defaults to the working directory. A problem with many conditions
writes one source per condition, so a scratch folder is preferable. Solver
settings default to the backend's; reproducing reference values computed at
tighter tolerances requires passing those tolerances.

The returned object is a plain list whose slots are ordinary dMod objects:

| slot | what it is |
|---|---|
| `dataList` | the measurements as a `datalist` |
| `reactions` | the reaction network as an `eqnlist` |
| `odemodel`, `x` | compiled model and prediction function |
| `g`, `e` | observation function and error model |
| `p` | parameter transformation, one branch per condition |
| `prd` | the composite `g * x * p` |
| `obj` | `normL2(dataList, prd, errmodel = e)`, fixed parameters baked in |
| `bestfit` | `nominalValue` per estimated parameter, on the PEtab scale |
| `parlower`, `parupper` | bounds, likewise on the PEtab scale |

With the fixed parameters inside `obj`, the likelihood at the published
optimum is

```{r import-obj, eval=FALSE}
petab$obj(petab$bestfit)$value
```

and a fit is the usual dMod call, with the bounds passed to the optimiser,
not to the objective:

```{r import-fit, eval=FALSE}
fit <- trust(petab$obj, petab$bestfit, rinit = 1, rmax = 10,
             parlower = petab$parlower, parupper = petab$parupper)
```

## Prerequisites

Reading the SBML model requires `python-libsbml`, which dMod2 provisions on
first use in a Python environment managed by `reticulate`. Setting
`DMOD_LIBSBML_PYTHON` to an existing interpreter uses that one instead. The
YAML and TSV readers, `readPetabYaml()` and `readPetabTables()`, work
without libsbml.

# Exporting a problem

**Imported problems.** `exportPEtabObject()` writes the manifest, the SBML
and every table. The tables it read are reproduced, including prior
declarations, per-row noise parameters and the period schedule.

```{r export-recipe, eval=FALSE}
exportPEtabObject(petab, dir = "out", formatVersion = "2.0.0")
```

| argument | meaning |
|---|---|
| `petab` | an imported problem, or a list with the same slots |
| `dir` | output directory, created if missing |
| `modelID` | model id in the manifest, defaults to the imported one |
| `formatVersion` | `"2.0.0"` or `"1"` |
| `overwrite` | overwrite an existing directory |

**Problems built by hand.** `exportPEtab()` takes the dMod objects directly
rather than a `petabproblem`:

```{r export-native, eval=FALSE}
exportPEtab(data = myData, reactions = myEqnlist,
            observables = myObservables,
            p = myTrafo, pouter = myPars,
            errors = myErrorModel,
            lower = myLower, upper = myUpper,
            parameterScale = "log10",
            dir = "out", formatVersion = "2.0.0")
```

`conditions.tsv` is recovered from the per-condition structure of the trafo
`p`. A symbol that `p` references but that is neither estimated nor
declared fixed raises an error.

# A worked example: Boehm

The package ships the JAK/STAT5 problem of Boehm et al.
[@boehm2014identification], a PEtab benchmark problem, so the example runs
as written.

```{r example-files}
petab_dir <- system.file("extdata/petab_boehm", package = "dMod2")
list.files(petab_dir)
```

The manifest and tables can be inspected before compilation:

```{r example-tables}
yamlPath <- file.path(petab_dir, "Boehm.yaml")
tables   <- readPetabTables(yamlPath)

obs <- tables$observables
data.frame(id = obs$observableId, noise = obs$noiseFormula)
```

The observable formulas:

```{r example-formulas}
writeLines(strwrap(paste0(obs$observableId, " = ",
                          obs$observableFormula),
                   width = 68, exdent = 4))
```

The model has eight STAT5 species, six estimated rate constants, three
estimated noise standard deviations and two fixed scaling parameters; all
estimated parameters are on `log10` scale.

## Import and evaluate

```{r example-import, results = "hide"}
wd <- tempfile("petab_vignette_"); dir.create(wd)
petab <- importPEtab(yamlPath, backend = "cppDE",
                     modelname = "boehm_vig", outdir = wd)
```

The objective, $-2\log L$, at the published maximum-likelihood estimate:

```{r example-obj}
petab$obj(petab$bestfit)$value
```

## Predict and plot

`prd` returns one trace per observable, and `plot()` draws them against the
data. The band spans one standard deviation either side, taken from the
error model: `as.data.frame(errfn = )` evaluates it alongside the prediction
and returns `sigma` per row.

```{r example-plot}
library(ggplot2)

fixed <- attr(petab, "petab_meta")$fixed
times <- seq(0, 240, length.out = 200)
pred  <- petab$prd(times, c(petab$bestfit, fixed), deriv = FALSE)

prdout <- as.data.frame(pred, errfn = petab$e)

plot(pred, petab$dataList) +
  geom_ribbon(data = prdout,
              aes(x = time, ymin = value - sigma, ymax = value + sigma),
              linetype = "dashed", alpha = 0.3)
```

## Write it back out

Boehm is a v1 problem, so exporting it as v2 linearises its
`parameterScale`, with a warning.

```{r example-export}
exportDir <- file.path(wd, "export"); dir.create(exportDir)
exported <- exportPEtabObject(petab, dir = exportDir,
                              formatVersion = "2.0.0",
                              overwrite = TRUE)
list.files(exportDir)
```

```{r example-cleanup, include=FALSE}
unlink(wd, recursive = TRUE)
```

# How the importer works

\begin{center}
\begin{tikzpicture}[
  font=\footnotesize,
  nbase/.style   = {draw, rounded corners=2pt, align=center,
                  minimum height=7mm, inner sep=3pt, line width=0.4pt},
  nfile/.style  = {nbase, fill=black!4,  text width=32mm},
  nstep/.style  = {nbase, fill=black!8,  text width=42mm},
  ntab/.style = {nbase, fill=black!3,  text width=24mm},
  nobj/.style   = {nbase, fill=black!8,  text width=30mm},
  nout/.style   = {nbase, fill=black!12, text width=52mm},
  narr/.style    = {-{Stealth[length=4pt]}, line width=0.4pt}
]
  \node[nfile] (in)   at (0, 0)    {YAML, TSV, SBML};
  \node[nstep] (norm) at (0, -1.3) {parse and normalise\\[-1pt]
                                   \tiny v1 and v2 onto one shape};

  \node[ntab] (tp) at (-5.1, -3.0) {parameters};
  \node[ntab] (tr) at (-1.7, -3.0) {reactions};
  \node[ntab] (to) at ( 1.7, -3.0) {observables};
  \node[ntab] (tm) at ( 5.1, -3.0) {measurements};

  \node[nobj] (op) at (-5.1, -4.6) {\texttt{p = P(trafo)}};
  \node[nobj] (or) at (-1.7, -4.6) {\texttt{x = Xs(odemodel)}};
  \node[nobj] (oo) at ( 1.7, -4.6) {\texttt{g = Y(...)}\\ \texttt{e}, error model};
  \node[nobj] (om) at ( 5.1, -4.6) {\texttt{dataList}};

  \node[nout] (prd) at (0, -6.2) {\texttt{prd = g * x * p}};
  \node[nout] (obj) at (0, -7.6) {\texttt{obj = normL2(dataList, prd, e)}};

  \draw[narr] (in)   -- (norm);
  \foreach \t in {tp, tr, to, tm} { \draw[narr] (norm) -- (\t); }
  \foreach \a/\b in {tp/op, tr/or, to/oo, tm/om} { \draw[narr] (\a) -- (\b); }
  \foreach \a in {op, or, oo, om} { \draw[narr] (\a) -- (prd); }
  \draw[narr] (prd) -- (obj);
\end{tikzpicture}
\end{center}

**Extensive against intensive flux.** SBML kinetic laws are in amount per
time, dMod2 rates in concentration per time. On import, each kinetic law is
divided by the volume of its home compartment,
$\text{rate}_{\text{dMod}} = K_{\text{SBML}} / V_{\text{home}}$, and on
export multiplied by it again. The home compartment is the shared
compartment of the educts, or of the products for a pure synthesis.
Assignment rules are inlined into rates and initial expressions before the
`eqnlist` is built.

**Experiments as period sequences.** An experiment is an ordered list of
`(time, conditionId)` pairs. A leading `-inf` period is the
preequilibration and becomes a `Pequil` stage. The first finite period
sets the simulation start, passed to the objective as `normL2(t0 = )`, so
initial values apply there rather than at zero. Each later period becomes
an event that applies its condition overrides at its start time. The number
of periods is unlimited.

**Event targets that are not states.** Solvers assign only to states, but
PEtab and SBML events may also assign to a parameter or a compartment
volume. Such a target is promoted to a state with rate zero and its SBML
default as initial value; export reverses the promotion.

**Placeholders per observable.** `measurements.tsv` may override
`observableParameter${k}_${id}` and `noiseParameter${k}_${id}` per row. The
observable id suffix keeps substitutions from colliding across observables.
The rows of one condition merge into one sub-condition whose trafo carries
one substitution per observable; a further sub-condition is created only
when an observable appears with conflicting values. Numeric noise is folded
into the data column and leaves the error model `NULL`, `normL2`'s fast
path; symbolic noise becomes an inner parameter of the trafo.

**Steady-state measurements.** A measurement at `time = inf` is evaluated
at the equilibration horizon that `Pequil` integrates to and exported as
`inf`.

# Priors

A parameter row may declare a `priorDistribution` with `priorParameters`.
With $\pi_i$ the prior density of parameter $i$, truncated to its bounds,
`obj()` returns the $-2\log$ posterior
$$
-2\log L - 2\sum_i \log\pi_i
$$
rather than $-2\log L$.

**Separating prior and likelihood.** The prior term $-2\sum_i \log\pi_i$ is
returned as the attribute `prior` of the result; subtracting it leaves
$-2\log L$, that is $-2$ times the `llh` PEtab publishes:

```{r prior-split}
res <- petab$obj(petab$bestfit)
prior <- attr(res, "prior")

c(total      = unname(res$value),
  prior      = if (is.null(prior)) 0 else unname(prior),
  likelihood = unname(res$value) - if (is.null(prior)) 0 else unname(prior))
```

Boehm declares no prior, so the attribute is absent and `total` equals
`likelihood`.

**Bounds are not a prior.** `lowerBound` and `upperBound` reach the fit as
`parlower` and `parupper` and never enter the objective, so a parameter
without a declared distribution contributes nothing. Implementations that
read bounds as an implicit uniform prior report a posterior that carries a
constant dMod2 leaves out.

**Truncation.** PEtab truncates every prior to the parameter bounds and
renormalises it to the mass between them; dMod2 does the same. The
renormalisation is constant in the parameter: it shifts the objective value
but not gradient or Hessian, so the fit is unaffected.

**Families and their parameterisation.**

| `priorDistribution` | parameters | note |
|---|---|---|
| `uniform` | a, b | density $1/(b-a)$ on $[a,b]$ |
| `normal` | mean, sd | |
| `laplace` | mu, b | |
| `cauchy` | x0, gamma | |
| `chisquare` | df | |
| `exponential` | **scale** | not a rate |
| `gamma` | shape, **scale** | not a rate |
| `rayleigh` | sigma | |
| `log-normal` | meanlog, sdlog | density in the parameter, so it carries the change of variables |
| `log-laplace` | mu, b | likewise |
| `log-uniform` | a, b | likewise |

Each name also exists with a `parameterScale` prefix. Prefixed priors act on
the parameter as the optimiser sees it, after the scale wrap; plain ones act
on the linear value. A plain `normal` or `laplace` on a non-`lin`
`parameterScale` is rejected: the two readings differ, and choosing one
silently would move the optimum.

# How the exporter works

The exporter reads the same slots and writes one table each. Besides undoing
the compartment scaling and the promotion of event targets, it reconstructs
the period schedule from the simulation start times and the mid-run events.

\begin{center}
\begin{tikzpicture}[
  font=\footnotesize,
  nbase/.style = {draw, rounded corners=2pt, align=center,
                  minimum height=7mm, inner sep=3pt, line width=0.4pt},
  nsrc/.style  = {nbase, fill=black!8,  text width=22mm},
  nband/.style = {nbase, fill=black!12, text width=118mm},
  nfil/.style  = {nbase, fill=black!3,  text width=22mm},
  nman/.style  = {nbase, fill=black!4,  text width=40mm},
  narr/.style  = {-{Stealth[length=4pt]}, line width=0.4pt}
]
  \node[nsrc] (sp) at (-5.6, 0) {\texttt{p}, bounds};
  \node[nsrc] (se) at (-2.8, 0) {start times,\\ events};
  \node[nsrc] (so) at ( 0.0, 0) {\texttt{g}, \texttt{e}};
  \node[nsrc] (sd) at ( 2.8, 0) {\texttt{dataList}};
  \node[nsrc] (sr) at ( 5.6, 0) {\texttt{reactions},\\ inits};

  \node[nband] (ex) at (0, -1.6) {\texttt{exportPEtabObject()}};

  \node[nfil] (fp) at (-5.6, -3.2) {parameters\\ conditions};
  \node[nfil] (fe) at (-2.8, -3.2) {experiments};
  \node[nfil] (fo) at ( 0.0, -3.2) {observables};
  \node[nfil] (fd) at ( 2.8, -3.2) {measurements};
  \node[nfil] (fr) at ( 5.6, -3.2) {\texttt{model.xml}};

  \node[nman] (man) at (0, -4.8) {YAML manifest};

  \foreach \n in {sp, se, so, sd, sr} { \draw[narr] (\n) -- (\n |- ex.north); }
  \foreach \n in {fp, fe, fo, fd, fr} { \draw[narr] (\n |- ex.south) -- (\n); }
  \foreach \n in {fp, fe, fo, fd, fr} { \draw[narr] (\n) -- (man); }
\end{tikzpicture}
\end{center}

`exportPEtab()` takes the same pieces as separate arguments.

# Feature coverage

All 31 v2 cases of the official PEtab test suite reproduce the published
likelihood and pass an export-to-v2 and reimport round trip.

Of the 35 problems in the
[Benchmark-Models](https://github.com/Benchmarking-Initiative/Benchmark-Models-PEtab)
collection, 33 import, evaluate their objective with derivatives, export to
v2 and reimport. Left out are `Froehlich_CellSystems2018`, whose 9570
conditions each produce their own generated transformation, and
`Laske_PLOSComputBiol2019`, where one observable divides by a sum of species
that is zero at the initial time.

## SBML model

| feature | dMod2 | notes |
|---|:---:|---|
| reactions and kinetic laws | yes | divided by the home compartment volume |
| compartments, multiple compartments | yes | first class on `eqnlist` |
| species in amounts or concentrations | yes | `hasOnlySubstanceUnits` respected |
| `<initialAssignment>` | yes | on species, parameters and compartments |
| `<assignmentRule>` | yes | inlined before the `eqnlist` is built |
| `<rateRule>` | yes | on species, and on parameters, which become states |
| `<event>`, time trigger | yes | rewritten as a dMod event |
| `<event>`, state trigger | yes | as a root condition |
| event assigning to a compartment | yes | contained species rescale, amounts conserved |
| `<functionDefinition>` | yes | inlined by libsbml |
| `piecewise` in a formula | yes | grouped into a symbolic piecewise |
| `<algebraicRule>` | no | dMod2 has no DAE solver |
| model languages other than SBML | no | only SBML is read |

## PEtab tables

| feature | dMod2 | notes |
|---|:---:|---|
| `parameterScale` | yes | `lin`, `log`, `log10`, folded into the trafo |
| `observableTransformation` | yes | `lin`, `log`, `log10` per observable |
| `conditions.tsv`, long format | yes | pivoted to wide for the trafo builder |
| condition target on a parameter or compartment | yes | promoted to a state when an event assigns to it |
| per-row `observableParameters`, `noiseParameters` | yes | numeric and symbolic |
| noise formula naming its own observable | yes | resolved from the observable equations |
| `experiments.tsv`, any number of periods | yes | preequilibration, start time, mid-run events |
| measurement at `time = inf` | yes | evaluated at the equilibration horizon |
| `mapping.tsv` | yes | textual rewrite of entity ids |
| several models in one manifest | yes | composed with `+`, condition keys prefixed |
| priors, all eleven families | yes | truncated to the bounds and renormalised |
| model without species | yes | a parameter-only problem |
| `visualisation.tsv` | no | ignored |

## Noise distributions

The supported noise distributions follow from dMod2's objective, `normL2`.

| `noiseDistribution` | dMod2 | notes |
|---|:---:|---|
| `normal` | yes | direct |
| `log-normal` | yes | through `observableTransformation = log` |
| `laplace`, `log-laplace` | no | rejected at import; an L1 residual is not implemented |

## Export caveats

| caveat | notes |
|---|---|
| `parameterScale` on a v2 export | linearised on disk with one warning; v2 has no such column |
| `log10` observable transformation on a v2 export | rewritten as `log-normal` noise; v2 has no `log10` |
| `<algebraicRule>` in the source model | absent from the dMod model and therefore from any SBML written back |
| experiment identity | experiments are keyed by their starting condition, so the exported `experimentId` need not match the imported one |

# References
