---
title: "PEtab problems in dMod2"
author: "Simon Beyer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
    keep_tex: false
    latex_engine: lualatex
bibliography: refs.bib
csl: ieee.csl
link-citations: true
header-includes:
  - \usepackage{fontspec}
  - \usepackage{unicode-math}
  - \setmainfont{CMU Serif}
  - \setsansfont{CMU Sans Serif}
  - \setmonofont{CMU Typewriter Text}
  - \setmathfont{NewCMMath-Regular.otf}
  - \usepackage{mathtools}
  - \usepackage{booktabs}
  - \usepackage{fancyvrb}
  - \fvset{fontsize=\small}
  - \usepackage{tikz}
  - \usetikzlibrary{arrows.meta, positioning, calc}
  - \usepackage{graphicx}
  - \usepackage{microtype}
  - \usepackage{etoolbox}
  - \AtBeginEnvironment{longtable}{\small}
  - \AtBeginDocument{\let\OldTOC\tableofcontents\renewcommand{\tableofcontents}{\OldTOC\clearpage}}
  - \setcounter{secnumdepth}{2}
fontsize: 10pt
---

```{r setup, include=FALSE}
options(width = 70)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      collapse = TRUE, comment = "#>",
                      dpi = 400, fig.width = 10, fig.height = 3)
library(dMod2)
have_libsbml <- isTRUE(tryCatch(
  reticulate::py_module_available("libsbml"), error = function(e) FALSE))
knitr::opts_chunk$set(eval = have_libsbml)
```

# What this vignette covers

PEtab [@schmiester2021petab] is a file format for parameter estimation
problems in systems biology: a YAML manifest, an SBML model
[@hucka2003sbml], and a handful of TSV tables. dMod2 reads such a problem
into a fully assembled fit problem in one call, and writes a dMod problem
back out in the same format. Both directions are covered here, first as a
recipe, then on a worked example, and finally as a table of what the
format offers against what dMod2 does with it.

Everything below uses PEtab v2 unless stated otherwise. Version 1 problems
are read as well; the importer normalises them onto the same internal
shape.

# What maps onto what

dMod2 composes a fit problem from function objects, `g * x * p` closed by
an objective [@kaschek2019dmod]. The importer fills exactly those slots:

| PEtab | dMod2 |
|---|---|
| SBML reactions, rules, events | `reactions` (an `eqnlist`), `odemodel`, `x = Xs(odemodel)` |
| `observables.tsv` formulas | `g = Y(observables)` |
| `observables.tsv` noise formulas | `e`, the error model, or `NULL` when every noise formula is constant |
| `parameters.tsv`, `conditions.tsv` | `p = P(trafo)`, one branch per condition |
| `measurements.tsv` | `dataList`, a `datalist` keyed by condition |
| `experiments.tsv` | simulation start times and mid-run events |
| `parameters.tsv` priors | a prior term added to the objective |

The composite is `prd = g * x * p` and the objective is
`normL2(dataList, prd, errmodel = e)`, with the fixed parameters baked in.

Two conventions are worth stating up front, because they decide what the
numbers in `bestfit` mean.

**Parameter scales live in the trafo, not in the data.** PEtab declares a
scale per parameter (`lin`, `log`, `log10`); dMod2 wraps `p` so the
optimiser sees the parameter on that scale and the chain rule carries
through the composition. `bestfit` is therefore on the PEtab scale, not
on the linear one.

**Bounds are not a prior.** `lowerBound` and `upperBound` reach the fit as
`parlower` and `parupper` and never enter the objective. A parameter that
declares no `priorDistribution` contributes nothing to the objective
value.

# Importing a problem

```{r import-recipe, eval=FALSE}
petab <- importPEtab("problem.yaml", backend = "cppDE")
```

That single call parses the manifest, reads the SBML through libsbml,
generates the right-hand side together with its sensitivity equations,
compiles them, and returns an object of class `petabproblem`. The
arguments worth knowing:

| argument | meaning |
|---|---|
| `yamlPath` | path to the YAML manifest |
| `backend` | `"cppDE"` (default, handles events and second-order sensitivities) or `"deSolve"` |
| `compile` | `FALSE` reuses an already compiled model of the same name |
| `cores` | conditions evaluated in parallel inside one objective call |
| `modelname` | base name for the generated sources |
| `outdir` | where the generated sources and the shared object are written |
| `optionsOde`, `optionsSens` | solver settings, forwarded to `Xs()` |

`outdir` defaults to the working directory, and a problem with many
conditions writes one source per condition, so a scratch folder is the
better choice. The solver defaults are the backend's; a problem whose
reference values were computed at tighter tolerances needs them passed.

The returned object is a plain list; its slots are ordinary dMod objects
that can be used, composed and plotted like any others:

| slot | what it is |
|---|---|
| `dataList` | the measurements as a `datalist` |
| `reactions` | the reaction network as an `eqnlist` |
| `odemodel`, `x` | compiled model and prediction function |
| `g`, `e` | observation function and error model |
| `p` | parameter transformation, one branch per condition |
| `prd` | the composite `g * x * p` |
| `obj` | `normL2(dataList, prd, errmodel = e)`, fixed parameters baked in |
| `bestfit` | `nominalValue` per estimated parameter, on the PEtab scale |
| `parlower`, `parupper` | bounds, likewise on the PEtab scale |

Because the fixed parameters are already inside `obj`, evaluating the
likelihood at the published optimum needs nothing else:

```{r import-obj, eval=FALSE}
petab$obj(petab$bestfit)$value
```

and fitting is the usual dMod call, with the bounds handed to the
optimiser rather than to the objective:

```{r import-fit, eval=FALSE}
fit <- trust(petab$obj, petab$bestfit, rinit = 1, rmax = 10,
             parlower = petab$parlower, parupper = petab$parupper)
```

## Prerequisites

Reading the SBML model needs `python-libsbml`. dMod2 provisions a managed
Python environment through `reticulate` on first use, so no manual setup
is required. To use an existing interpreter instead, point
`DMOD_LIBSBML_PYTHON` at it. The YAML and TSV readers, `readPetabYaml()`
and `readPetabTables()`, work without libsbml.

# Exporting a problem

There are two entry points, depending on where the problem comes from.

**A problem that was imported** goes back out with
`exportPEtabObject()`. It writes the manifest, the SBML, and every table,
and reproduces the tables it read: prior declarations, per-row noise
parameters and the period schedule all survive the round trip.

```{r export-recipe, eval=FALSE}
exportPEtabObject(petab, dir = "out", formatVersion = "2.0.0")
```

| argument | meaning |
|---|---|
| `petab` | an imported problem, or a list with the same slots |
| `dir` | output directory, created if missing |
| `modelID` | model id in the manifest, defaults to the imported one |
| `formatVersion` | `"2.0.0"` or `"1"` |
| `overwrite` | overwrite an existing directory |

**A problem built by hand** goes out with `exportPEtab()`, which takes the
dMod objects directly rather than a `petabproblem`:

```{r export-native, eval=FALSE}
exportPEtab(data = myData, reactions = myEqnlist,
            observables = myObservables,
            p = myTrafo, pouter = myPars,
            errors = myErrorModel,
            lower = myLower, upper = myUpper,
            parameterScale = "log10",
            dir = "out", formatVersion = "2.0.0")
```

The trafo `p` carries the per-condition structure, so the exporter
recovers `conditions.tsv` from it. Symbols that the trafo references but
that are neither estimated nor declared fixed raise an error rather than
being written out silently.

# A worked example

The package ships the Boehm et al. JAK/STAT5 problem
[@boehm2014identification], the canonical PEtab benchmark, so the example
below runs as written.

```{r example-files}
petab_dir <- system.file("extdata/petab_boehm", package = "dMod2")
list.files(petab_dir)
```

The manifest and the tables can be inspected before anything is compiled:

```{r example-tables}
yamlPath <- file.path(petab_dir, "Boehm.yaml")
tables   <- readPetabTables(yamlPath)

obs <- tables$observables
data.frame(id = obs$observableId, noise = obs$noiseFormula)
```

The observable formulas are long, so they are shown wrapped rather than as a
table column:

```{r example-formulas}
writeLines(strwrap(paste0(obs$observableId, " = ",
                          obs$observableFormula),
                   width = 68, exdent = 4))
```

The model has eight STAT5 species, six estimated rate constants, three
estimated noise standard deviations and two fixed scaling parameters, all
estimated parameters on `log10` scale.

## Import and evaluate

```{r example-import, results = "hide"}
wd <- tempfile("petab_vignette_"); dir.create(wd)
petab <- importPEtab(yamlPath, backend = "cppDE",
                     modelname = "boehm_vig", outdir = wd)
```

Compiling the model prints the compiler invocation, which is hidden here.
The objective is then evaluated at the published maximum-likelihood estimate:

```{r example-obj}
petab$obj(petab$bestfit)$value
```

The value is $-2\log L$ at the published maximum-likelihood estimate.

## Predict and plot

`prd` returns one trace per observable, and `plot()` draws it against the
data. The band is one standard deviation, read off the error model rather
than reconstructed by hand: `as.data.frame(errfn = )` evaluates it alongside
the prediction and returns `sigma` per row.

```{r example-plot}
library(ggplot2)

fixed <- attr(petab, "petab_meta")$fixed
times <- seq(0, 240, length.out = 200)
pred  <- petab$prd(times, c(petab$bestfit, fixed), deriv = FALSE)

prdout <- as.data.frame(pred, errfn = petab$e)

plot(pred, petab$dataList) +
  geom_ribbon(data = prdout,
              aes(x = time, ymin = value - sigma, ymax = value + sigma),
              linetype = "dashed", alpha = 0.3)
```

## Write it back out

The imported problem exports as a v2 problem in one call. Boehm arrives as
v1, so its `parameterScale` is linearised on the way out, which the exporter
warns about.

```{r example-export}
exportDir <- file.path(wd, "export"); dir.create(exportDir)
exported <- exportPEtabObject(petab, dir = exportDir,
                              formatVersion = "2.0.0",
                              overwrite = TRUE)
list.files(exportDir)
```

```{r example-cleanup, include=FALSE}
unlink(wd, recursive = TRUE)
```

# How the importer works

\begin{center}
\begin{tikzpicture}[
  font=\footnotesize,
  nbase/.style   = {draw, rounded corners=2pt, align=center,
                  minimum height=7mm, inner sep=3pt, line width=0.4pt},
  nfile/.style  = {nbase, fill=black!4,  text width=32mm},
  nstep/.style  = {nbase, fill=black!8,  text width=42mm},
  ntab/.style = {nbase, fill=black!3,  text width=24mm},
  nobj/.style   = {nbase, fill=black!8,  text width=30mm},
  nout/.style   = {nbase, fill=black!12, text width=52mm},
  narr/.style    = {-{Stealth[length=4pt]}, line width=0.4pt}
]
  \node[nfile] (in)   at (0, 0)    {YAML, TSV, SBML};
  \node[nstep] (norm) at (0, -1.3) {parse and normalise\\[-1pt]
                                   \tiny v1 and v2 onto one shape};

  \node[ntab] (tp) at (-5.1, -3.0) {parameters};
  \node[ntab] (tr) at (-1.7, -3.0) {reactions};
  \node[ntab] (to) at ( 1.7, -3.0) {observables};
  \node[ntab] (tm) at ( 5.1, -3.0) {measurements};

  \node[nobj] (op) at (-5.1, -4.6) {\texttt{p = P(trafo)}};
  \node[nobj] (or) at (-1.7, -4.6) {\texttt{x = Xs(odemodel)}};
  \node[nobj] (oo) at ( 1.7, -4.6) {\texttt{g = Y(...)}\\ \texttt{e}, error model};
  \node[nobj] (om) at ( 5.1, -4.6) {\texttt{dataList}};

  \node[nout] (prd) at (0, -6.2) {\texttt{prd = g * x * p}};
  \node[nout] (obj) at (0, -7.6) {\texttt{obj = normL2(dataList, prd, e)}};

  \draw[narr] (in)   -- (norm);
  \foreach \t in {tp, tr, to, tm} { \draw[narr] (norm) -- (\t); }
  \foreach \a/\b in {tp/op, tr/or, to/oo, tm/om} { \draw[narr] (\a) -- (\b); }
  \foreach \a in {op, or, oo, om} { \draw[narr] (\a) -- (prd); }
  \draw[narr] (prd) -- (obj);
\end{tikzpicture}
\end{center}

Six pieces are not obvious from that picture.

**Extensive against intensive flux.** SBML kinetic laws carry units of
amount per time, dMod2 stores rates as concentration per time. Each
kinetic law is divided by the volume of its home compartment on import,
$\text{rate}_{\text{dMod}} = K_{\text{SBML}} / V_{\text{home}}$, and
multiplied by it again on export. The home compartment is the shared
compartment of the educts, falling back to the products for a pure
synthesis. Assignment rules are inlined into rates and initial
expressions before the `eqnlist` is built.

**Experiments are period sequences.** An experiment is an ordered list of
`(time, conditionId)` pairs. A leading `-inf` period is the
preequilibration and becomes a `Pequil` stage. The first finite period
sets the simulation start, which reaches the objective as
`normL2(t0 = )`, so initial values take effect there rather than at zero.
Every later period becomes an event that applies its condition overrides
at that time. There is no limit on the number of periods.

**Event targets that are not states.** Solvers assign only to states, but
PEtab and SBML both allow an event to assign to a parameter or to a
compartment volume. Such a target is promoted to a state with rate zero,
keeping its SBML default as the initial value; the export takes the
promotion back.

**Placeholders are per observable.** `measurements.tsv` may override
`observableParameter${k}_${id}` and `noiseParameter${k}_${id}` per row.
Placeholders carry the observable id as a suffix, so substitutions never
collide across observables. Rows inside one condition merge into a single
sub-condition whose trafo carries one substitution per observable; a new
sub-condition appears only when the same observable turns up with
conflicting values. Numeric noise folds into the data column and leaves
the error model `NULL`, which is `normL2`'s fast path; symbolic noise
becomes an inner parameter of the trafo.

**Steady-state measurements.** A measurement at `time = inf` is evaluated
on the equilibration horizon that `Pequil` integrates to, and is written
back as `inf` on export.

# Priors, and telling them from the likelihood

A parameter row may declare a `priorDistribution` with its
`priorParameters`. dMod turns each one into a term on the objective's
$-2\log$ scale and adds it to the likelihood, so `obj()` returns
$-2\log$ posterior, not $-2\log$ likelihood.

**The two parts stay separable.** The prior contribution is reported in an
attribute of the result, so subtracting it recovers exactly the quantity
PEtab publishes as `llh`:

```{r prior-split}
res <- petab$obj(petab$bestfit)
prior <- attr(res, "prior")

c(total      = unname(res$value),
  prior      = if (is.null(prior)) 0 else unname(prior),
  likelihood = unname(res$value) - if (is.null(prior)) 0 else unname(prior))
```

Boehm declares no prior, so the attribute is absent and the two agree. For a
problem that does declare one, `total` is $-2\log$ posterior and `likelihood`
is what a published `llh` can be compared against.

**Bounds are not a prior.** `lowerBound` and `upperBound` reach the fit as
`parlower` and `parupper` and never enter the objective. A parameter with
bounds but no declared distribution contributes nothing. This differs from
implementations that read bounds as an implicit uniform prior; there the
reported posterior carries a constant that dMod leaves out.

**Truncation.** PEtab truncates every prior to the parameter bounds and
renormalises it to the mass between them. dMod does the same. The
renormalisation is a constant in the parameter, so it shifts the objective
value without touching gradient or Hessian, and the fit is unaffected.

**The families, and where their parameterisation bites.**

| `priorDistribution` | parameters | note |
|---|---|---|
| `uniform` | a, b | density $1/(b-a)$ on $[a,b]$ |
| `normal` | mean, sd | |
| `laplace` | mu, b | |
| `cauchy` | x0, gamma | |
| `chisquare` | df | |
| `exponential` | **scale** | not a rate |
| `gamma` | shape, **scale** | not a rate |
| `rayleigh` | sigma | |
| `log-normal` | meanlog, sdlog | density in the parameter, so it carries the change of variables |
| `log-laplace` | mu, b | likewise |
| `log-uniform` | a, b | likewise |

Each spelling also exists with a `parameterScale` prefix. Those act on the
parameter as the optimiser sees it, i.e. after the scale wrap, and the plain
spellings act on the linear value. A plain `normal` or `laplace` on a
non-`lin` `parameterScale` is rejected rather than guessed at: the two
readings differ, and picking one silently would move the optimum.

# How the exporter works

The reverse direction reads the same slots and writes one table each. What is
not a plain read-back is the model: the kinetic laws are multiplied by their
home compartment volume again, promoted event targets go back to being
parameters or compartments, and the period schedule is reconstructed from the
simulation start times and the mid-run events.

\begin{center}
\begin{tikzpicture}[
  font=\footnotesize,
  nbase/.style = {draw, rounded corners=2pt, align=center,
                  minimum height=7mm, inner sep=3pt, line width=0.4pt},
  nsrc/.style  = {nbase, fill=black!8,  text width=22mm},
  nband/.style = {nbase, fill=black!12, text width=118mm},
  nfil/.style  = {nbase, fill=black!3,  text width=22mm},
  nman/.style  = {nbase, fill=black!4,  text width=40mm},
  narr/.style  = {-{Stealth[length=4pt]}, line width=0.4pt}
]
  \node[nsrc] (sp) at (-5.6, 0) {\texttt{p}, bounds};
  \node[nsrc] (se) at (-2.8, 0) {start times,\\ events};
  \node[nsrc] (so) at ( 0.0, 0) {\texttt{g}, \texttt{e}};
  \node[nsrc] (sd) at ( 2.8, 0) {\texttt{dataList}};
  \node[nsrc] (sr) at ( 5.6, 0) {\texttt{reactions},\\ inits};

  \node[nband] (ex) at (0, -1.6) {\texttt{exportPEtabObject()}};

  \node[nfil] (fp) at (-5.6, -3.2) {parameters\\ conditions};
  \node[nfil] (fe) at (-2.8, -3.2) {experiments};
  \node[nfil] (fo) at ( 0.0, -3.2) {observables};
  \node[nfil] (fd) at ( 2.8, -3.2) {measurements};
  \node[nfil] (fr) at ( 5.6, -3.2) {\texttt{model.xml}};

  \node[nman] (man) at (0, -4.8) {YAML manifest};

  \foreach \n in {sp, se, so, sd, sr} { \draw[narr] (\n) -- (\n |- ex.north); }
  \foreach \n in {fp, fe, fo, fd, fr} { \draw[narr] (\n |- ex.south) -- (\n); }
  \foreach \n in {fp, fe, fo, fd, fr} { \draw[narr] (\n) -- (man); }
\end{tikzpicture}
\end{center}

A problem built by hand goes through `exportPEtab()` instead, which takes the
same pieces as separate arguments rather than reading them off a
`petabproblem`.

# Feature coverage

Measured against the official PEtab test suite: all 31 v2 cases reproduce
the published likelihood, and all 31 survive an export-to-v2 and reimport
round trip.

Beyond the test suite, the
[Benchmark-Models](https://github.com/Benchmarking-Initiative/Benchmark-Models-PEtab)
collection was run end to end: 33 of its 35 problems import, evaluate their
objective with derivatives, export to v2 and reimport.
`Froehlich_CellSystems2018` is left out because each of its 9570 conditions
produces its own generated transformation, and `Laske_PLOSComputBiol2019`
because one of its observables divides by a sum of species that is zero at the
initial time.

## SBML model

| feature | dMod2 | notes |
|---|:---:|---|
| reactions and kinetic laws | yes | divided by the home compartment volume |
| compartments, multiple compartments | yes | first class on `eqnlist` |
| species in amounts or concentrations | yes | `hasOnlySubstanceUnits` respected |
| `<initialAssignment>` | yes | on species, parameters and compartments |
| `<assignmentRule>` | yes | inlined before the `eqnlist` is built |
| `<rateRule>` | yes | on species, and on parameters, which become states |
| `<event>`, time trigger | yes | rewritten as a dMod event |
| `<event>`, state trigger | yes | as a root condition |
| event assigning to a compartment | yes | contained species rescale, amounts conserved |
| `<functionDefinition>` | yes | inlined by libsbml |
| `piecewise` in a formula | yes | grouped into a symbolic piecewise |
| `<algebraicRule>` | no | dMod2 has no DAE solver |
| model languages other than SBML | no | only SBML is read |

## PEtab tables

| feature | dMod2 | notes |
|---|:---:|---|
| `parameterScale` | yes | `lin`, `log`, `log10`, folded into the trafo |
| `observableTransformation` | yes | `lin`, `log`, `log10` per observable |
| `conditions.tsv`, long format | yes | pivoted to wide for the trafo builder |
| condition target on a parameter or compartment | yes | promoted to a state when an event assigns to it |
| per-row `observableParameters`, `noiseParameters` | yes | numeric and symbolic |
| noise formula naming its own observable | yes | resolved from the observable equations |
| `experiments.tsv`, any number of periods | yes | preequilibration, start time, mid-run events |
| measurement at `time = inf` | yes | evaluated at the equilibration horizon |
| `mapping.tsv` | yes | textual rewrite of entity ids |
| several models in one manifest | yes | composed with `+`, condition keys prefixed |
| priors, all eleven families | yes | truncated to the bounds and renormalised |
| model without species | yes | a parameter-only problem |
| `visualisation.tsv` | no | ignored |

## Noise distributions

dMod2's objective is `normL2`, so the noise model maps as follows.

| `noiseDistribution` | dMod2 | notes |
|---|:---:|---|
| `normal` | yes | direct |
| `log-normal` | yes | through `observableTransformation = log` |
| `laplace`, `log-laplace` | no | rejected at import, an L1 residual is not implemented |

## Export caveats

| caveat | notes |
|---|---|
| `parameterScale` on a v2 export | linearised on disk with one warning, v2 has no such column |
| `log10` observable transformation on a v2 export | rewritten as `log-normal` noise, v2 has no `log10` |
| `<algebraicRule>` in the source model | absent from the dMod model and therefore from any SBML written back |
| experiment identity | experiments are keyed by their starting condition, so the exported `experimentId` need not match the imported one |

# References
