---
title: "Optimisation in dMod2"
subtitle: "Hessian sources, seeding, and the economics of a multi-start"
author: "Simon Beyer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
    keep_tex: false
    latex_engine: lualatex
    fig_caption: true
bibliography: refs.bib
csl: ieee.csl
link-citations: true
header-includes:
  - \usepackage{fontspec}
  - \usepackage{unicode-math}
  - \setmainfont{CMU Serif}
  - \setsansfont{CMU Sans Serif}
  - \setmonofont{Latin Modern Mono}
  - \setmathfont{NewCMMath-Regular.otf}
  - \usepackage{mathtools}
  - \usepackage{booktabs}
  - \usepackage{fancyvrb}
  - \usepackage{fvextra}
  - \fvset{fontsize=\small,breaklines=true,breakanywhere=true}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{microtype}
  - \usepackage{etoolbox}
  - \AtBeginEnvironment{longtable}{\small}
  - \AtBeginDocument{\let\OldTOC\tableofcontents\renewcommand{\tableofcontents}{\OldTOC\clearpage}}
  - \setcounter{secnumdepth}{2}
fontsize: 10pt
---

```{r setup, include=FALSE}
options(width = 70)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      collapse = TRUE, comment = "#>",
                      dpi = 400, fig.width = 6.5, fig.height = 3.8,
                      fig.align = "center", fig.pos = "!ht")
library(dMod2)
library(ggplot2)
.outdir <- file.path(tempdir(), "vig_optim")
dir.create(.outdir, recursive = TRUE, showWarnings = FALSE)
.cores <- max(1L, min(20L, parallel::detectCores() - 2L))

# The fits are stored next to this document and reused, so a render reproduces
# the shipped PDF exactly and takes seconds. Only what the document uses is
# kept; `dev/render-optimisation.R --refit` drops the file and fits again.
.store <- "results.rds"
.saved <- if (file.exists(.store)) readRDS(.store) else list()
.fit <- function(name, expr) {
  if (is.null(.saved[[name]])) {
    .saved[[name]] <<- expr
    saveRDS(.saved, .store)
  }
  .saved[[name]]
}
```

# Introduction

Parameter estimation for ordinary differential equation models is a non-convex
problem solved by local optimisation from many starting points. dMod2
[@kaschek2019dmod] provides `trust()`, a trust-region method, and `mstrust()`,
which runs it from many starts. This vignette describes their arguments and the
behaviour those produce.

The example throughout is the STAT5 dimerisation model of Boehm et al.
[@boehm2014identification], built as in the shipped example
`example_Boehm_JProteomeRes2014.R`: nine parameters on a base-10 logarithmic
scale, a weak prior on the six dynamic ones, and bounds of $\pm 5$. Every table
and figure below rests on one shared set of 100 starting points.

```{r problem, results = "hide"}
data(boehm)
mydataL <- as.datalist(boehm)

epo <- "1.25e-7*exp(-Epo_degradation_BaF3*time)"
reactions <- eqnlist() |>
  assignCompartment(nucpApA = "nuc", nucpApB = "nuc", nucpBpB = "nuc",
                    volume = "0.45") |>
  addReaction("2*STAT5A", "pApA", paste0("k_phos*", epo, "*STAT5A^2"),
              compartment = "cyt") |>
  addReaction("STAT5A + STAT5B", "pApB", paste0("k_phos*", epo, "*STAT5A*STAT5B"),
              compartment = "cyt") |>
  addReaction("2*STAT5B", "pBpB", paste0("k_phos*", epo, "*STAT5B^2"),
              compartment = "cyt") |>
  addReaction("pApA", "nucpApA", "k_imp_homo*pApA",     compartment = "cyt") |>
  addReaction("pApB", "nucpApB", "k_imp_hetero*pApB",   compartment = "cyt") |>
  addReaction("pBpB", "nucpBpB", "k_imp_homo*pBpB",     compartment = "cyt") |>
  addReaction("nucpApA", "2*STAT5A", "k_exp_homo*nucpApA",   compartment = "nuc") |>
  addReaction("nucpApB", "STAT5A + STAT5B", "k_exp_hetero*nucpApB",
              compartment = "nuc") |>
  addReaction("nucpBpB", "2*STAT5B", "k_exp_homo*nucpBpB",   compartment = "nuc") |>
  setCompartmentVolume(cyt = "1.4")

# The optimiser can only resolve what the integrator delivers, so the tolerances
# are set explicitly rather than left at their defaults.
myOptionsODE <- list(atol = 1e-8, rtol = 1e-6, maxattemps = 100L, maxsteps = 1e6)
model <- odemodel(reactions, modelname = "boehm_ode", compile = FALSE,
                  outdir = .outdir)
x <- Xs(model, optionsOde = myOptionsODE, optionsSens = myOptionsODE)

observables <- eqnvec(
  pSTAT5A_rel = paste0("(100*pApB + 200*pApA*specC17)/",
                       "(pApB + STAT5A*specC17 + 2*pApA*specC17)"),
  pSTAT5B_rel = paste0("-(100*pApB - 200*pBpB*(specC17 - 1))/",
                       "((STAT5B*(specC17 - 1) - pApB) + 2*pBpB*(specC17 - 1))"),
  rSTAT5A_rel = paste0("(100*pApB + 100*STAT5A*specC17 + 200*pApA*specC17)/",
                       "(2*pApB + STAT5A*specC17 + 2*pApA*specC17",
                       " - STAT5B*(specC17 - 1) - 2*pBpB*(specC17 - 1))"))
errorModels <- eqnvec(pSTAT5A_rel = "sd_pSTAT5A_rel",
                      pSTAT5B_rel = "sd_pSTAT5B_rel",
                      rSTAT5A_rel = "sd_rSTAT5A_rel")

g <- Y(observables, x, modelname = "boehm_obs", attach.input = FALSE,
       compile = FALSE, outdir = .outdir)
e <- Y(errorModels, g, modelname = "boehm_err", attach.input = FALSE,
       compile = FALSE, outdir = .outdir)

innerpars <- getParameters(model, g, e)
estimated <- c("Epo_degradation_BaF3", "k_exp_hetero", "k_exp_homo",
               "k_imp_hetero", "k_imp_homo", "k_phos",
               "sd_pSTAT5A_rel", "sd_pSTAT5B_rel", "sd_rSTAT5A_rel")
p <- eqnvec() |>
  define("x~x", x = innerpars) |>
  define("x~0", x = c("pApA", "pApB", "pBpB", "nucpApA", "nucpApB", "nucpBpB")) |>
  insert("STAT5A ~ 207.6*ratio") |>
  insert("STAT5B ~ 207.6 - 207.6*ratio") |>
  insert("ratio ~ 0.693") |>
  insert("specC17 ~ 0.107") |>
  insert("x~exp10(x)", x = estimated) |>
  P(modelname = "boehm_trafo", condition = "Boehm2014", compile = FALSE,
    outdir = .outdir)

compile(g, x, p, e, output = "boehm_vig", cores = 2)

prd       <- g*x*p
outerpars <- getParameters(prd)
pouter    <- structure(rep(-1, length(outerpars)), names = outerpars)
dyn       <- setdiff(outerpars, grep("^sd_", outerpars, value = TRUE))
obj       <- normL2(mydataL, prd, e) + constraintL2(pouter[dyn], sigma = 4)

parlower <- structure(rep(-5, length(outerpars)), names = outerpars)
parupper <- structure(rep( 5, length(outerpars)), names = outerpars)
```

```{r problem-size}
length(pouter)
```

# The trust-region core

`trust()` minimises a smooth objective by repeatedly solving the constrained
quadratic subproblem

$$\min_{\lVert p \rVert \le r} \; g^{\mathsf T} p + \tfrac12 p^{\mathsf T} H p ,$$

exactly, through a full eigendecomposition of $H$ [@more1983computing]. Box
constraints are handled by the interior scheme of Coleman and Li
[@coleman1996interior], in which the distance to the bound that the gradient
pushes towards enters the metric of the subproblem. Iterates therefore remain
strictly inside the feasible box and never land on its boundary; the `atBound`
component of the result reports which parameters a bound is holding.

The subproblem is solved exactly rather than by truncated conjugate gradients
because the quasi-Newton updates described below can produce an indefinite $H$:
the Moré-Sorensen solver treats that case directly, whereas a Steihaug-type
method can only stop at the trust-region boundary.

## Termination

Six tests can end a run, and the `stopReason` component names the one that
fired. Which of them apply depends on where the Hessian comes from.

| `stopReason` | criterion | evaluated on |
|---|---|---|
| `gradient` | $\max \lvert v \cdot g \rvert \le$ `gtol` | the current iterate |
| `fvalue` | $\lvert f - f_{\mathrm{try}} \rvert <$ `ftol` | the last accepted step |
| `preddiff` | predicted decrease $<$ `mtol` | the last accepted step |
| `step` | step norm $<$ `xtol` | the last accepted step |
| `stagnation` | five rejected steps leaving $f$ flat | recent steps |
| `radius` | radius $<$ `rmin`; reports `converged = FALSE` | the radius |

Only the first tests first-order optimality. Three of the others, `fvalue`,
`preddiff` and `step`, read the prediction of the quadratic model, so a Hessian
that misrepresents the local curvature can trip them at a point that is not
stationary. The hybrid switch below exploits that asymmetry; it also accounts
for a fit reporting `converged = TRUE` at an implausibly high objective value.

Which tests may end a run follows from the same distinction. While the model is
the objective's own $J^{\mathsf T} J$, its prediction is a curvature statement
and the three apply as given. During a quasi-Newton phase the model is an
approximation still being built, so its verdict describes the approximation
rather than the iterate; the three are then suppressed and the run ends on
`gradient`, on `stagnation`, which counts rejected steps instead of reading the
model, or on the radius. The gradient norm is the termination criterion usually
recommended for quasi-Newton methods [@nocedal2006numerical].

# The source of the model Hessian

The argument `hessianMethod` selects where $H$ comes from. The subproblem
solver, the scaling and the bound handling are identical in all four cases.

| value | Hessian | required from the objective |
|---|---|---|
| `"gn"` | the objective's $J^{\mathsf T} J$, rebuilt each iteration | value, gradient, Hessian |
| `"bfgs"` | a Powell-damped BFGS update [@powell1978fast] | value, gradient |
| `"sr1"` | a symmetric rank-one update, possibly indefinite | value, gradient |
| `"hybrid"` | `"gn"`, then `"bfgs"` after the first soft stop | value, gradient, Hessian |

For a sum of squares the Gauss-Newton matrix $J^{\mathsf T} J$ is an adequate
curvature model wherever the residuals are small, and in dMod2 it is obtained
without additional cost: the same Jacobian yields both the gradient and the
Hessian, so a gradient and a full derivative evaluation are measurably the same
price. This is why `"gn"` is the default.

The Gauss-Newton matrix omits the term involving residuals times second
derivatives. Close to the optimum the quadratic model therefore ceases to
improve, and `ftol` or `mtol` fires while the iterate is still short of the
minimum. A quasi-Newton approximation does not carry that bias, since it infers
curvature from the steps actually taken.

## The hybrid switch

`"hybrid"` runs `"gn"` until the first of `fvalue`, `preddiff`, `step` or
`stagnation` would terminate the run, then switches once to `"bfgs"` and
continues from the last accepted $J^{\mathsf T} J$ as its seed. The `gradient`
test remains terminal, since it is a genuine optimality statement rather than a
symptom of a stalled model. The construction follows Fides
[@froehlich2022fides], which switches when the trust radius stops updating. The
`qnEval` component counts the objective evaluations spent after the switch.

```{r hybrid}
set.seed(1L)
start <- as.parvec(msParframe(pouter, n = 1, sd = 3))

single <- .fit("single", list(
  gn     = trust(obj, start, rinit = 0.1, rmax = 10, iterlim = 5000,
                 parlower = parlower, parupper = parupper),
  hybrid = trust(obj, start, rinit = 0.1, rmax = 10, iterlim = 5000,
                 parlower = parlower, parupper = parupper,
                 hessianMethod = "hybrid")))

data.frame(method = names(single),
           value  = vapply(single, `[[`, 0.0, "value"),
           stop   = vapply(single, `[[`, "",  "stopReason"),
           neval  = vapply(single, `[[`, 0L,  "neval"),
           qnEval = vapply(single, `[[`, 0L,  "qnEval"),
           row.names = NULL)
```

Over the shared starting points of the next section the hybrid spends a few per
cent more evaluations than `"gn"` and reaches a lower median value. The `grad`
column of the table there counts the starts that end at a stationary point,
which the hybrid reaches and `"gn"` does not.

# Seeding and memory of the quasi-Newton approximation

Three arguments shape the approximation itself. They are read when the run
begins in a quasi-Newton phase, that is for `"bfgs"` and `"sr1"`; `"gn"`
ignores them, and `"hybrid"` seeds its handover from the Gauss-Newton matrix
irrespective of `hessianInit`.

`hessianInit` selects the initial matrix. The default `"gn"` uses the
objective's Hessian at `parinit`. The alternative `"identity"` starts from the
identity and additionally stops `trust()` requesting a Hessian at all, which is
what a derivative scheme returning only gradients would be able to supply.

The expectation that a genuine Hessian must be the better seed does not survive
a multi-start, which is why the identity is the seed used throughout the rest of
this vignette. At a starting point three decades away from the optimum,
$J^{\mathsf T} J$ is large and severely ill-conditioned. A BFGS update is of
rank two and preserves positive definiteness, so it cannot cut such a seed down
along the directions where it is wrong by orders of magnitude, and the
approximation stays dominated by curvature that was never right; on this model
those runs exhaust the iteration cap rather than reaching a termination test.
`"sr1"` is not affected, because its update is of rank one and carries no sign
constraint, so it can overwrite bad curvature outright, and the indefinite
approximations that follow are what the Moré-Sorensen subproblem solves.
`inst/benchmarks` measures both seeds side by side.

`qnMemory` bounds the curvature that is carried forward. The default `0`
accumulates every update onto the seed, which is what the rest of this vignette
uses. A positive value rebuilds the approximation in each iteration from
$\gamma I$ together with the most recent `qnMemory` pairs, where $\gamma$ is the
Shanno-Phua scaling $y^{\mathsf T} y / s^{\mathsf T} y$ of the newest pair
[@shanno1978matrix]. This is the curvature information of L-BFGS
[@nocedal2006numerical], but assembled into the explicit matrix the subproblem
requires, so it saves no memory. It is meant for problems where carrying a dense
$n_\theta \times n_\theta$ matrix is itself the binding cost, which begins at
parameter counts an order of magnitude above anything shown here.

`qnCautious` rejects a BFGS pair whose curvature along the step is negligible,
$s^{\mathsf T} y \le$ `qnCautious` $\cdot \lVert s \rVert \lVert y \rVert$
[@li2001modified]; `qnSkipped` counts the rejections. On this model it changes
neither the outcome nor the cost.

```{r seed}
seeds <- .fit("seeds", lapply(list(
  list(hessianMethod = "bfgs"),
  list(hessianMethod = "bfgs", hessianInit = "identity"),
  list(hessianMethod = "bfgs", hessianInit = "identity", qnMemory = 10L)),
  function(a) do.call(trust, c(list(obj, start, rinit = 0.1, rmax = 10,
                                    iterlim = 5000, parlower = parlower,
                                    parupper = parupper), a))))

data.frame(seed   = c("gn", "identity", "identity"),
           memory = c(0L, 0L, 10L),
           value  = vapply(seeds, `[[`, 0.0, "value"),
           stop   = vapply(seeds, `[[`, "", "stopReason"),
           neval  = vapply(seeds, `[[`, 0L, "neval"))
```

# Multi-start behaviour

`mstrust()` runs `trust()` from many starting points. Drawing them once with
`msParframe()` and passing the resulting frame as `center` ensures that a
comparison between methods differs only in the method.

```{r runs}
set.seed(20260905)
starts <- msParframe(pouter, n = 100, sd = 3)

fitAll <- function(...)
  as.parframe(mstrust(obj, center = starts, fits = nrow(starts),
                      cores = .cores, rinit = 0.1, rmax = 10, iterlim = 5000,
                      parlower = parlower, parupper = parupper, ...))

settings <- list(
  `gn`       = list(hessianMethod = "gn"),
  `hybrid`   = list(hessianMethod = "hybrid"),
  `bfgs, id` = list(hessianMethod = "bfgs", hessianInit = "identity"),
  `sr1, id`  = list(hessianMethod = "sr1",  hessianInit = "identity"))

runs <- .fit("multistart", lapply(settings, function(a) do.call(fitAll, a)))
```

A start counts as a hit when it lands within `0.1` of the best value seen
anywhere.

```{r per-start}
best <- min(vapply(runs, function(f) min(f$value), 0.0))

summary <- function(f)
  data.frame(conv     = sum(f$converged),
             grad     = sum(f$stopReason == "gradient"),
             hits     = sum(f$value <= best + 0.1),
             neval    = sum(f$neval),
             medEval  = stats::median(f$neval),
             medValue = stats::median(f$value))

cbind(source = names(runs), do.call(rbind, lapply(runs, summary)),
      row.names = NULL)
```

The waterfall plot sorts the final objective values in ascending order. A flat
leading plateau indicates repeated arrival at the same optimum, and the height
of the first step how well that optimum is separated from its competitors.
`plotValues()` draws one collection, shaded by iteration count; the figure below
ranks the four separately and colours them by method. A circle marks a converged
fit, a triangle one that stopped on any other test.

```{r waterfall, fig.width = 7.8, fig.height = 5.5, out.width = "100%", fig.cap = "Waterfall plots of the four Hessian sources over the same 100 starting points, each ranked separately. Values above 600 are omitted."}
# One panel for several fit collections, each ranked on its own. Drawn back to
# front, so the first one listed is not buried, and without the palette's
# pale yellow.
waterfall <- function(frames, cutoff = Inf) {
  ranked <- do.call(rbind, lapply(rev(names(frames)), function(set) {
    fits <- subset(frames[[set]], value < cutoff)
    fits <- fits[order(fits$value), ]
    data.frame(index = seq_len(nrow(fits)), value = fits$value,
               converged = as.character(fits$converged), set = set)
  }))
  ranked$set <- factor(ranked$set, levels = names(frames))
  colours <- stats::setNames(dMod_palette(8, "okabe")[-5][seq_along(frames)],
                             names(frames))

  ggplot(ranked, aes(index, value, colour = set)) +
    geom_line(aes(group = set), linewidth = 0.3) +
    geom_point(aes(shape = converged), size = 1.3) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 17),
                       limits = c("TRUE", "FALSE"), labels = c("yes", "no")) +
    scale_colour_manual(values = colours) +
    labs(x = "index", y = "value", colour = NULL) +
    theme_dMod() +
    theme(legend.position = "bottom", legend.box = "vertical")
}

waterfall(runs, cutoff = 600)
```

The curves share their starting points and therefore their basins, and they
fall into two shapes. Gauss-Newton and the hybrid track each other over the
whole range, as they must, since the hybrid is Gauss-Newton until a soft stop
fires; they differ in the leading plateau, which is not flat under Gauss-Newton
because the model stops improving before the minimum is reached, whereas the
quasi-Newton phase carries those starts onto a common value. The two
identity-seeded quasi-Newton runs have a different shape altogether: they reach
the global optimum from more starting points and abandon more of them on the
poor plateau, so they explore more widely and end on a worse typical value.

The `stopReason` of the runs shows the mechanism. Gauss-Newton ends on the
model tests, which for it are legitimate. The quasi-Newton runs, which those
tests no longer terminate, end on `stagnation` or on `gradient`.

# Discussion

The four sources differ in what a single start buys. `"gn"` and the hybrid
produce the better typical run, reaching the lower median value, and the hybrid
additionally resolves the leading plateau that `"gn"` leaves ragged, for a few
per cent more evaluations. The two identity-seeded quasi-Newton sources spend
fewer evaluations on a start and reach the published optimum from more of them,
at a worse typical value. Which of those matters depends on whether starting
points or evaluations are the scarce resource.

`parscale`, the diagonal preconditioner, is off by default and stays off here.
It has nothing to correct when parameters are already on a logarithmic scale,
which is the normal case in dMod2, and the ill-conditioning of sloppy models
lies along rotated directions [@gutenkunst2007universally] that no diagonal
scaling reaches.

The measurements come from a single model with nine estimated parameters, and the advantage of a quasi-Newton method is
expected to narrow as the parameter count grows, since a dense update requires
of the order of $n_\theta$ accepted steps before it carries curvature along
every direction. The hit counts are single digit to low double digit out of 100
starts and carry a corresponding sampling error, so they do not establish an
ordering between two sources that lie close together. A ranking that carried
weight would have to run over a set of models.

# Reproducing the numbers

The fits behind the tables and figures are stored in
`dev/optimisation/results.rds` and reused, so this document renders in seconds
and reproduces exactly what it shows; `dev/render-optimisation.R --refit`
recomputes them. The script `bench_hessianSource.R` under `inst/benchmarks`
runs the same comparison outside the vignette, over both seeds and with the
per-start `stopReason` tables.

# References
