---
title: "Optimisation in dMod2"
subtitle: "Hessian sources, seeding, and the economics of a multi-start"
author: "Simon Beyer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
    keep_tex: false
    latex_engine: lualatex
    fig_caption: true
bibliography: refs.bib
csl: ieee.csl
link-citations: true
header-includes:
  - \usepackage{fontspec}
  - \usepackage{unicode-math}
  - \setmainfont{CMU Serif}
  - \setsansfont{CMU Sans Serif}
  - \setmonofont{Latin Modern Mono}
  - \setmathfont{NewCMMath-Regular.otf}
  - \usepackage{mathtools}
  - \usepackage{booktabs}
  - \usepackage{fancyvrb}
  - \usepackage{fvextra}
  - \fvset{fontsize=\small,breaklines=true,breakanywhere=true}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{microtype}
  - \usepackage{etoolbox}
  - \AtBeginEnvironment{longtable}{\small}
  - \AtBeginDocument{\let\OldTOC\tableofcontents\renewcommand{\tableofcontents}{\OldTOC\clearpage}}
  - \setcounter{secnumdepth}{2}
fontsize: 10pt
---

```{r setup, include=FALSE}
options(width = 70)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      collapse = TRUE, comment = "#>",
                      dpi = 400, fig.width = 6.5, fig.height = 3.8,
                      fig.align = "center", fig.pos = "!ht")
library(dMod2)
library(ggplot2)
.outdir <- file.path(tempdir(), "vig_optim")
dir.create(.outdir, recursive = TRUE, showWarnings = FALSE)
.cores <- max(1L, min(20L, parallel::detectCores() - 2L))

# The fits are stored next to this document and reused, so a render reproduces
# the shipped PDF exactly and takes seconds. Only what the document uses is
# kept; `dev/render-optimisation.R --refit` drops the file and fits again.
.store <- "results.rds"
.saved <- if (file.exists(.store)) readRDS(.store) else list()
.fit <- function(name, expr) {
  if (is.null(.saved[[name]])) {
    .saved[[name]] <<- expr
    saveRDS(.saved, .store)
  }
  .saved[[name]]
}
```

# Introduction

Parameter estimation for ordinary differential equation models is a non-convex
problem solved by local optimisation from many starting points. dMod2
[@kaschek2019dmod] provides `trust()`, a trust-region method, and `mstrust()`,
which runs it from many starts. This vignette describes their arguments and the
behaviour those produce.

The example throughout is the STAT5 dimerisation model of Boehm et al.
[@boehm2014identification], built as in the shipped example
`example_Boehm_JProteomeRes2014.R`: nine parameters on a base-10 logarithmic
scale, a weak prior on the six dynamic ones, and bounds of $\pm 5$. Every table
and figure below rests on one shared set of 100 starting points.

```{r problem, results = "hide"}
data(boehm)
mydataL <- as.datalist(boehm)

epo <- "1.25e-7*exp(-Epo_degradation_BaF3*time)"
reactions <- eqnlist() |>
  assignCompartment(nucpApA = "nuc", nucpApB = "nuc", nucpBpB = "nuc",
                    volume = "0.45") |>
  addReaction("2*STAT5A", "pApA", paste0("k_phos*", epo, "*STAT5A^2"),
              compartment = "cyt") |>
  addReaction("STAT5A + STAT5B", "pApB", paste0("k_phos*", epo, "*STAT5A*STAT5B"),
              compartment = "cyt") |>
  addReaction("2*STAT5B", "pBpB", paste0("k_phos*", epo, "*STAT5B^2"),
              compartment = "cyt") |>
  addReaction("pApA", "nucpApA", "k_imp_homo*pApA",     compartment = "cyt") |>
  addReaction("pApB", "nucpApB", "k_imp_hetero*pApB",   compartment = "cyt") |>
  addReaction("pBpB", "nucpBpB", "k_imp_homo*pBpB",     compartment = "cyt") |>
  addReaction("nucpApA", "2*STAT5A", "k_exp_homo*nucpApA",   compartment = "nuc") |>
  addReaction("nucpApB", "STAT5A + STAT5B", "k_exp_hetero*nucpApB",
              compartment = "nuc") |>
  addReaction("nucpBpB", "2*STAT5B", "k_exp_homo*nucpBpB",   compartment = "nuc") |>
  setCompartmentVolume(cyt = "1.4")

# The optimiser can only resolve what the integrator delivers, so the tolerances
# are set explicitly rather than left at their defaults.
myOptionsODE <- list(atol = 1e-8, rtol = 1e-6, maxattemps = 100L, maxsteps = 1e6)
model <- odemodel(reactions, modelname = "boehm_ode", compile = FALSE,
                  outdir = .outdir)
x <- Xs(model, optionsOde = myOptionsODE, optionsSens = myOptionsODE)

observables <- eqnvec(
  pSTAT5A_rel = paste0("(100*pApB + 200*pApA*specC17)/",
                       "(pApB + STAT5A*specC17 + 2*pApA*specC17)"),
  pSTAT5B_rel = paste0("-(100*pApB - 200*pBpB*(specC17 - 1))/",
                       "((STAT5B*(specC17 - 1) - pApB) + 2*pBpB*(specC17 - 1))"),
  rSTAT5A_rel = paste0("(100*pApB + 100*STAT5A*specC17 + 200*pApA*specC17)/",
                       "(2*pApB + STAT5A*specC17 + 2*pApA*specC17",
                       " - STAT5B*(specC17 - 1) - 2*pBpB*(specC17 - 1))"))
errorModels <- eqnvec(pSTAT5A_rel = "sd_pSTAT5A_rel",
                      pSTAT5B_rel = "sd_pSTAT5B_rel",
                      rSTAT5A_rel = "sd_rSTAT5A_rel")

g <- Y(observables, x, modelname = "boehm_obs", attach.input = FALSE,
       compile = FALSE, outdir = .outdir)
e <- Y(errorModels, g, modelname = "boehm_err", attach.input = FALSE,
       compile = FALSE, outdir = .outdir)

innerpars <- getParameters(model, g, e)
estimated <- c("Epo_degradation_BaF3", "k_exp_hetero", "k_exp_homo",
               "k_imp_hetero", "k_imp_homo", "k_phos",
               "sd_pSTAT5A_rel", "sd_pSTAT5B_rel", "sd_rSTAT5A_rel")
p <- eqnvec() |>
  define("x~x", x = innerpars) |>
  define("x~0", x = c("pApA", "pApB", "pBpB", "nucpApA", "nucpApB", "nucpBpB")) |>
  insert("STAT5A ~ 207.6*ratio") |>
  insert("STAT5B ~ 207.6 - 207.6*ratio") |>
  insert("ratio ~ 0.693") |>
  insert("specC17 ~ 0.107") |>
  insert("x~exp10(x)", x = estimated) |>
  P(modelname = "boehm_trafo", condition = "Boehm2014", compile = FALSE,
    outdir = .outdir)

compile(g, x, p, e, output = "boehm_vig", cores = 2)

prd       <- g*x*p
outerpars <- getParameters(prd)
pouter    <- structure(rep(-1, length(outerpars)), names = outerpars)
dyn       <- setdiff(outerpars, grep("^sd_", outerpars, value = TRUE))
obj       <- normL2(mydataL, prd, e) + constraintL2(pouter[dyn], sigma = 4)

parlower <- structure(rep(-5, length(outerpars)), names = outerpars)
parupper <- structure(rep( 5, length(outerpars)), names = outerpars)
```

```{r problem-size}
length(pouter)
```

# The trust-region core

`trust()` minimises a smooth objective by repeatedly solving the constrained
quadratic subproblem

$$\min_{\lVert p \rVert \le r} \; g^{\mathsf T} p + \tfrac12 p^{\mathsf T} H p ,$$

exactly, through a full eigendecomposition of $H$ [@more1983computing]. Box
constraints are handled by the interior scheme of Coleman and Li
[@coleman1996interior], in which the distance to the bound that the gradient
pushes towards enters the metric of the subproblem. Iterates therefore remain
strictly inside the feasible box and never land on its boundary; the `atBound`
component of the result reports which parameters a bound is holding.

The subproblem is solved exactly rather than by truncated conjugate gradients
because the quasi-Newton updates described below can produce an indefinite $H$:
the Moré-Sorensen solver treats that case directly, whereas a Steihaug-type
method can only stop at the trust-region boundary.

## The subproblem

Moré and Sorensen characterise the solution of the subproblem by a scalar. With
the eigendecomposition $H = Q \Lambda Q^{\mathsf T}$ and $\gamma = Q^{\mathsf T} g$,
a step $p(\lambda) = -Q (\Lambda + \lambda I)^{-1} \gamma$ solves it exactly when
$\lambda \ge \max(0, -\lambda_{\min})$ and either $\lambda = 0$ with
$\lVert p \rVert \le r$, or $\lambda > 0$ with $\lVert p \rVert = r$. The
second case is a root of the secular equation

$$\phi(\lambda) \;=\; \frac{1}{r} - \frac{1}{\lVert p(\lambda) \rVert}
 \;=\; \frac{1}{r} - \Big( \sum_i \frac{\gamma_i^2}{(\lambda_i + \lambda)^2} \Big)^{-1/2},$$

which is smooth and almost linear on $\lambda > -\lambda_{\min}$, so Newton's
method on $\phi$ converges in a handful of iterations. Writing it this way is
what makes the eigendecomposition worth its cost: one decomposition per
iteration prices every subsequent $\lambda$ at $O(n_\theta)$.

The interesting case is the one a positive definite model never reaches. When
$\gamma_i = 0$ for every eigenvector of $\lambda_{\min}$, no $\lambda$ drives
$\lVert p \rVert$ out to $r$, and the solution is
$p = p(-\lambda_{\min}) + \tau q_{\min}$ with $\tau$ chosen so that
$\lVert p \rVert = r$. This hard case is exactly what an indefinite SR1 model
produces, and it is why the exact solver is not a luxury here.

## The step and the radius

One iteration compares the decrease the model promised with the decrease the
objective delivered,

$$\rho_k \;=\; \frac{f(x_k) - f(x_k + p_k)}{-m_k(p_k)},
\qquad m_k(p) = g_k^{\mathsf T} p + \tfrac12 p^{\mathsf T} H_k p ,$$

and reads the radius off that ratio. A step with $\rho_k < 1/4$ is rejected and
the radius contracts to $\min(r_k/4, \lVert p_k \rVert / 4)$, so a step that
was far shorter than the radius shrinks it to its own scale rather than to a
quarter of an irrelevant bound. A step with $\rho_k > 3/4$ that used at least
nine tenths of the radius doubles it, capped at `rmax`; anything between leaves
the radius alone. Only pressing against the boundary earns an increase, since a
step that stopped short was limited by the model and not by the region.

Under `stepControl$nonmonotone` the numerator is measured against a weighted
average of past values rather than the current one [@zhang2004nonmonotone]:

$$Q_{k+1} = \eta Q_k + 1, \qquad
C_{k+1} = \frac{\eta Q_k C_k + f(x_{k+1})}{Q_{k+1}},$$

with $\rho_k$ scored against $C_k$. At $\eta = 0$ this is $C_k = f(x_k)$ and
the rule above is recovered exactly. A larger $\eta$ lets the iterate climb out
of a narrow valley at the price of accepting steps that raise the objective, so
the result reports the best iterate visited rather than the last one. The
termination tests keep reading the true objective either way: acceptance is
relaxed, convergence is not.

## Bounds

Coleman and Li turn a bound into a change of metric. With
$v_i$ the signed distance from $x_i$ to whichever bound the gradient pushes it
towards, $D = \operatorname{diag}(\lvert v \rvert^{1/2})$, and $J^v$ the
diagonal of $\partial \lvert v \rvert / \partial x$, the subproblem is solved
in the scaled variables

$$\hat g = D g, \qquad \hat B = D H D + \operatorname{diag}(g) J^v ,$$

so the model steepens as a bound is approached and the step turns away from it
on its own. The first-order optimality measure that `gtol` reads is
$\max_i \lvert v_i g_i \rvert$, which vanishes both at an interior stationary
point and at a bound the gradient pushes into, and is therefore the right
quantity for a box problem. A step that would still leave the box is truncated
to a fraction `theta.max` of the distance, reflected off the blocking faces, or
replaced by the scaled steepest-descent step, whichever the model rates best.

## Termination

Six tests can end a run, and the `stopReason` component names the one that
fired. Which of them apply depends on where the Hessian comes from.

| `stopReason` | criterion | evaluated on |
|---|---|---|
| `gradient` | $\max \lvert v \cdot g \rvert \le$ `gtol` | the current iterate |
| `fvalue` | $\lvert f - f_{\mathrm{try}} \rvert <$ `ftol` | the last accepted step |
| `preddiff` | predicted decrease $<$ `mtol` | the last accepted step |
| `step` | step norm $<$ `xtol` | the last accepted step |
| `stagnation` | five rejected steps leaving $f$ flat | recent steps |
| `radius` | radius $<$ `rmin`; reports `converged = FALSE` | the radius |

Only the first tests first-order optimality. Three of the others, `fvalue`,
`preddiff` and `step`, read the prediction of the quadratic model, so a Hessian
that misrepresents the local curvature can trip them at a point that is not
stationary. The handover below exploits that asymmetry; it also accounts
for a fit reporting `converged = TRUE` at an implausibly high objective value.

Which tests may end a run follows from the same distinction. While the model is
the objective's own $J^{\mathsf T} J$, its prediction is a curvature statement
and the three apply as given. During a quasi-Newton phase the model is an
approximation still being built, so its verdict describes the approximation
rather than the iterate; the three are then suppressed and the run ends on
`gradient`, on `stagnation`, which counts rejected steps instead of reading the
model, or on the radius. The gradient norm is the termination criterion usually
recommended for quasi-Newton methods [@nocedal2006numerical].

# The source of the model Hessian

The argument `hessianMethod` selects where $H$ comes from. The subproblem
solver, the scaling and the bound handling are identical in all four cases.

| value | Hessian | required from the objective |
|---|---|---|
| `"gn"` | the objective's $J^{\mathsf T} J$, rebuilt each iteration | value, gradient, Hessian |
| `"bfgs"` | a Powell-damped BFGS update [@powell1978fast] | value, gradient |
| `"sr1"` | a symmetric rank-one update, possibly indefinite | value, gradient |

`hessianFallback` names a second source and `fallbackLimit` how often the two
may change places. A run is a method and, optionally, a fallback; no particular
pair carries a name of its own.

For a sum of squares the Gauss-Newton matrix $J^{\mathsf T} J$ is an adequate
curvature model wherever the residuals are small, and in dMod2 it is obtained
without additional cost: the same Jacobian yields both the gradient and the
Hessian, so a gradient and a full derivative evaluation are measurably the same
price. This is why `"gn"` is the default.

The Gauss-Newton matrix omits the term involving residuals times second
derivatives. Close to the optimum the quadratic model therefore ceases to
improve, and `ftol` or `mtol` fires while the iterate is still short of the
minimum. A quasi-Newton approximation does not carry that bias, since it infers
curvature from the steps actually taken.

## BFGS and SR1

Both updates read the same secant pair, the step taken and the change in
gradient along it,

$$s_k = x_{k+1} - x_k, \qquad y_k = \nabla f(x_{k+1}) - \nabla f(x_k),$$

and both ask the new model to reproduce it, $B_{k+1} s_k = y_k$. They differ in
what they are willing to give up to do so.

**BFGS** is the rank-two update that stays positive definite, which it can only
do while $s^{\mathsf T} y > 0$. Away from a minimum that curvature condition
fails, so the pair is damped before use [@powell1978fast]: with

$$\theta = \begin{cases}
1, & s^{\mathsf T} y \ge 0.2\, s^{\mathsf T} B s, \\[2pt]
\dfrac{0.8\, s^{\mathsf T} B s}{s^{\mathsf T} B s - s^{\mathsf T} y}, & \text{otherwise,}
\end{cases}
\qquad \tilde y = \theta y + (1 - \theta) B s ,$$

the update is

$$B_{k+1} = B_k - \frac{B_k s\, s^{\mathsf T} B_k}{s^{\mathsf T} B_k s}
                 + \frac{\tilde y\, \tilde y^{\mathsf T}}{s^{\mathsf T} \tilde y}.$$

Damping interpolates the pair towards $Bs$, for which the curvature condition
holds by construction, so $B$ never loses definiteness. `qnCautious` adds the
Li-Fukushima test on top [@li2001modified]: a pair counts only when
$s^{\mathsf T} y > c \lVert s \rVert \lVert y \rVert$, which discards pairs
whose curvature is numerical noise rather than information. `qnSkipped` counts
what it rejected.

**SR1** is the rank-one update that satisfies the secant equation without
insisting on definiteness. With $w = y - B_k s$,

$$B_{k+1} = B_k + \frac{w\, w^{\mathsf T}}{w^{\mathsf T} s},$$

which is the only symmetric rank-one matrix that does so, and is therefore
forced rather than chosen. The price is the denominator: $w^{\mathsf T} s$ can
be arbitrarily small, so the update is skipped when
$\lvert w^{\mathsf T} s \rvert \le 10^{-8} \lVert s \rVert \lVert w \rVert$.
SR1 converges to the true Hessian, indefinite regions included, which is
useless to a line search and exactly what a trust region wants: the negative
curvature it reports is real, and the Moré-Sorensen solver can step along it.

That is also why `qnRejected` is on for SR1. A rejected trial point still
supplies a valid $(s, y)$, since curvature along a step does not depend on
whether the step was worth taking, and the trial gradient has been paid for
already [@nocedal2006numerical]. BFGS cannot use it: its positive definiteness
argument rests on the step having been accepted.

## The handover

A fallback run stays on `"gn"` until the first of `fvalue`, `preddiff`, `step`
or `stagnation` would terminate it, then switches to the fallback source and
continues from the last accepted $J^{\mathsf T} J$ as its seed. The `gradient`
test remains terminal, since it is a genuine optimality statement rather than a
symptom of a stalled model. The construction follows Fides
[@froehlich2022fides], which switches when the trust radius stops updating. The
`qnEval` component counts the objective evaluations spent after the switch.

```{r handover}
set.seed(1L)
start <- as.parvec(msParframe(pouter, n = 1, sd = 3))

single <- .fit("single", list(
  gn     = trust(obj, start, rinit = 0.1, rmax = 10, iterlim = 5000,
                 parlower = parlower, parupper = parupper),
  fallback = trust(obj, start, rinit = 0.1, rmax = 10, iterlim = 5000,
                 parlower = parlower, parupper = parupper,
                 hessianFallback = "bfgs")))

data.frame(method = names(single),
           value  = vapply(single, `[[`, 0.0, "value"),
           stop   = vapply(single, `[[`, "",  "stopReason"),
           neval  = vapply(single, `[[`, 0L,  "neval"),
           qnEval = vapply(single, `[[`, 0L,  "qnEval"),
           row.names = NULL)
```

Over the shared starting points of the next section a fallback run spends a few
per cent more evaluations than `"gn"`. The `grad` column of the table there
counts the starts that end at a stationary point, which a fallback reaches and
`"gn"` does not.

# Seed and memory of the approximation

Three arguments shape the approximation itself. They are read when the run
begins in a quasi-Newton phase, that is for `"bfgs"` and `"sr1"`; `"gn"`
ignores them, and a fallback seeds its handover from the Gauss-Newton matrix
irrespective of `hessianInit`.

`hessianInit` selects the initial matrix. The default `"gn"` uses the
objective's Hessian at `parinit`. The alternative `"identity"` starts from the
identity and additionally stops `trust()` requesting a Hessian at all, which is
what a derivative scheme returning only gradients would be able to supply.

The expectation that a genuine Hessian must be the better seed does not survive
a multi-start, which is why the identity is the seed used throughout the rest of
this vignette. At a starting point three decades away from the optimum,
$J^{\mathsf T} J$ is large and severely ill-conditioned. A BFGS update is of
rank two and preserves positive definiteness, so it cannot cut such a seed down
along the directions where it is wrong by orders of magnitude, and the
approximation stays dominated by curvature that was never right; on this model
those runs exhaust the iteration cap rather than reaching a termination test.
`"sr1"` is not affected, because its update is of rank one and carries no sign
constraint, so it can overwrite bad curvature outright, and the indefinite
approximations that follow are what the Moré-Sorensen subproblem solves.
`inst/benchmarks` measures both seeds side by side.

`qnMemory` bounds the curvature that is carried forward. The default `0`
accumulates every update onto the seed, which is what the rest of this vignette
uses. A positive value rebuilds the approximation in each iteration from
$\gamma I$ together with the most recent `qnMemory` pairs, where $\gamma$ is the
Shanno-Phua scaling $y^{\mathsf T} y / s^{\mathsf T} y$ of the newest pair
[@shanno1978matrix]. This is the curvature information of L-BFGS
[@nocedal2006numerical], but assembled into the explicit matrix the subproblem
requires, so it saves no memory. It is meant for problems where carrying a dense
$n_\theta \times n_\theta$ matrix is itself the binding cost, which begins at
parameter counts an order of magnitude above anything shown here.

`qnCautious` rejects a BFGS pair whose curvature along the step is negligible,
$s^{\mathsf T} y \le$ `qnCautious` $\cdot \lVert s \rVert \lVert y \rVert$
[@li2001modified]; `qnSkipped` counts the rejections. On this model it changes
neither the outcome nor the cost.

```{r seed}
seeds <- .fit("seeds", lapply(list(
  list(hessianMethod = "bfgs"),
  list(hessianMethod = "bfgs", qnControl = list(hessianInit = "identity")),
  list(hessianMethod = "bfgs",
       qnControl = list(hessianInit = "identity", qnMemory = 10L))),
  function(a) do.call(trust, c(list(obj, start, rinit = 0.1, rmax = 10,
                                    iterlim = 5000, parlower = parlower,
                                    parupper = parupper), a))))

data.frame(seed   = c("gn", "identity", "identity"),
           memory = c(0L, 0L, 10L),
           value  = vapply(seeds, `[[`, 0.0, "value"),
           stop   = vapply(seeds, `[[`, "", "stopReason"),
           neval  = vapply(seeds, `[[`, 0L, "neval"))
```

# Multi-start on Boehm

`mstrust()` runs `trust()` from many starting points. Drawing them once with
`msParframe()` and passing the resulting frame as `center` ensures that a
comparison between methods differs only in the method.

```{r runs}
set.seed(20260905)
starts <- msParframe(pouter, n = 100, sd = 3)

fitAll <- function(...)
  as.parframe(mstrust(obj, center = starts, fits = nrow(starts),
                      cores = .cores, rinit = 0.1, rmax = 10, iterlim = 5000,
                      parlower = parlower, parupper = parupper, ...))

source(system.file("benchmarks", "hessianSourceSettings.R", package = "dMod2"))
settings <- hessianSourceSettings[c("gn", "gn_bfgs", "gn_sr1", "bfgs_id", "sr1_id")]

runs <- .fit("multistart", lapply(settings, function(a) do.call(fitAll, a)))
names(runs) <- hessianSourceLabels[names(settings)]
```

A start counts as a hit when it lands within `0.1` of the best value seen
anywhere.

```{r per-start}
best <- min(vapply(runs, function(f) min(f$value), 0.0))

summary <- function(f) {
  hits <- sum(f$value <= best + 0.1)
  data.frame(conv       = sum(f$converged),
             grad       = sum(f$stopReason == "gradient"),
             hits       = hits,
             neval      = sum(f$neval),
             medEval    = stats::median(f$neval),
             # The price of one start that reaches the optimum, as reported by
             # the Bachmann run.
             evalPerHit = if (hits > 0) round(sum(f$neval) / hits) else NA_real_)
}

cbind(source = names(runs), do.call(rbind, lapply(runs, summary)),
      row.names = NULL)
```

The waterfall plot sorts the final objective values in ascending order. A flat
leading plateau indicates repeated arrival at the same optimum, and the height
of the first step how well that optimum is separated from its competitors.
`plotValues()` draws one collection, shaded by iteration count; the figure below
ranks the five separately and colours them by method. A circle marks a converged
fit, a triangle one that stopped on any other test.

```{r waterfall, fig.width = 7.8, fig.height = 5.5, out.width = "100%", fig.cap = "Waterfall plots of the five Hessian sources over the same 100 starting points, each ranked separately. Values above 600 are omitted."}
# One panel for several fit collections, each ranked on its own. Drawn back to
# front, so the first one listed is not buried, and without the palette's
# pale yellow.
waterfall <- function(frames, cutoff = Inf) {
  ranked <- do.call(rbind, lapply(rev(names(frames)), function(set) {
    fits <- subset(frames[[set]], value < cutoff)
    fits <- fits[order(fits$value), ]
    data.frame(index = seq_len(nrow(fits)), value = fits$value,
               converged = as.character(fits$converged), set = set)
  }))
  ranked$set <- factor(ranked$set, levels = names(frames))
  colours <- stats::setNames(dMod_palette(8, "okabe")[-5][seq_along(frames)],
                             names(frames))

  ggplot(ranked, aes(index, value, colour = set)) +
    geom_line(aes(group = set), linewidth = 0.3) +
    geom_point(aes(shape = converged), size = 1.3) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 17),
                       limits = c("TRUE", "FALSE"), labels = c("yes", "no")) +
    scale_colour_manual(values = colours) +
    labs(x = "index", y = "value", colour = NULL) +
    theme_dMod() +
    theme(legend.position = "bottom", legend.box = "vertical")
}

waterfall(runs, cutoff = 600)
```

The curves share their starting points and therefore their basins, and they
fall into two shapes. Gauss-Newton and the two fallback runs track each other
over the whole range, as they must, since a fallback is Gauss-Newton until a
soft stop fires; they differ in the leading plateau, which is not flat under
Gauss-Newton because the model stops improving before the minimum is reached,
whereas the quasi-Newton phase carries those starts onto a common value. The
identity-seeded runs have a different shape altogether, with a longer leading
plateau: they reach the global optimum from more starting points and abandon
more of them further down. Under `"bfgs"` that trade costs the typical value,
under `"sr1"` it does not.

The `stopReason` of the runs shows the mechanism. Gauss-Newton ends on the
model tests, which for it are legitimate. The quasi-Newton runs, which those
tests no longer terminate, end on `stagnation` or on `gradient`.

## Boehm without the prior

The prior above keeps a start from running into a bound, and it shapes the
comparison. The same 200 starts were therefore also run on the likelihood
alone, with the bounds widened to $\pm 8$. The published optimum is
$-2 \log L = 276.44$ [@hass2019benchmark], and a hit ends within 0.1 of it.

| source | best | reaching the optimum | median evaluations | evaluations per optimum |
|---|---|---|---|---|
| `"gn"` | 276.396 | 28 of 200 | 98 | 4236 |
| `"gn"` to `"sr1"` | 276.396 | 28 of 200 | 133 | 4452 |
| `"sr1"`, identity | 276.396 | 32 of 200 | 219 | 4190 |
| `"sr1"`, gn seed | 276.396 | **40 of 200** | 226 | 2440 |
| `"sr1"` to `"gn"` | 276.396 | 35 of 200 | 172 | **2411** |

```{r boehm-noprior-waterfall, fig.width = 7.8, fig.height = 4.0, out.width = "100%", fig.cap = "Waterfall plots of five Hessian sources on the Boehm likelihood without a prior, over the same 200 starting points, each ranked separately. Values above 600 are omitted."}
noprior <- readRDS("boehm_noprior.rds")
waterfall(split(noprior[c("value", "converged")], noprior$source), cutoff = 600)
```

Removing the prior costs every source: each still reaches the published
optimum, but from fewer starts and at several times the evaluations per one.
The ordering changes. `"sr1"` now reaches the optimum from the most starts and
for the fewest evaluations each, and the reversed handover is level with it at
a fifth fewer evaluations per start.

The reversed handover also repairs how a run ends. Alone, `"sr1"` stagnates on
156 of 200 starts; handing back to `"gn"` turns that into 159 ends on `fvalue`
and 11 on stagnation, and cuts runs exhausting the iteration cap from 13 to 4.
The forward handover does none of this and matches `"gn"` on every count, as on
Bachmann.

# Bachmann, 113 parameters

A dense quasi-Newton update needs of the order of $n_\theta$ accepted steps
before it carries curvature along every direction, so its advantage should
narrow as the problem grows. Bachmann's JAK2-STAT5 model
[@bachmann2011division] puts a number on that: 113 estimated parameters, 25
states, 20 observables, 36 conditions. The fits come from a SLURM cluster
through `dMod2::distributedComputing()`, one job per source, each an array of
50 tasks running four fits.

The setup needs one deliberate choice. A multi-start over 113 parameters needs
a weak prior, or starts run into the bounds instead of converging, but one
centre for every parameter is wrong where the published values span twelve
decades: at $\sigma = 4$ it costs 27 at the published optimum and moves it.
Every parameter further than a decade from the default centre is therefore
moved onto the order of magnitude of its published value, which leaves the
centre within one decade everywhere and the same $\sigma = 4$ costing 1.04. The
objective keeps its published minimum, at -479.26, and a hit ends within 0.1 of
it. Starts are drawn around that centre and redrawn where they leave the box.

| source | best | reaching the optimum | median evaluations | evaluations per optimum |
|---|---|---|---|---|
| `"gn"` | -479.264 | 83 of 200 | 141 | **638** |
| `"gn"` to `"sr1"` | -479.264 | 83 of 200 | 156 | 678 |
| `"sr1"`, identity | -479.264 | 79 of 200 | 500 | 1292 |
| `"sr1"`, gn seed | -479.264 | 65 of 200 | 592 | 1926 |
| `"sr1"` to `"gn"` | -479.264 | 79 of 200 | 490 | 1275 |

```{r bachmann-waterfall, fig.width = 7.8, fig.height = 4.0, out.width = "100%", fig.cap = "Waterfall plots of five Hessian sources on Bachmann over the same 200 starting points, each ranked separately. Starts ending above -400 are omitted."}
# Cluster fits, so they sit beside the document as data rather than in the
# refit cache: a --refit cannot recompute them.
bachmann <- readRDS("bachmann.rds")
waterfall(split(bachmann[c("value", "converged")], bachmann$source), cutoff = -400)
```

Every source reaches the same best value, and the counts at 0.1 and at 1.0 from
the optimum are identical, so a start either finds the published minimum or
ends far from it. No source finds an optimum another misses. What separates
them is price: Gauss-Newton reaches the optimum for a third of what the
cheapest quasi-Newton run pays, reversing the ordering seen at nine
parameters.

Two differences are structural rather than statistical.

An identity seed beats a Gauss-Newton seed for `"sr1"`, 79 starts against 65.
Seeding an approximation that is about to be rebuilt from scratch with a matrix
from a point it is about to leave carries the wrong curvature into the first
steps, and SR1 has no damping to discard it.

The handovers do what their construction says and no more. Forward, the model
tests never fire again and 197 of 200 runs end on stagnation, at the values
`"gn"` reaches alone. Reversed, the run ends on `fvalue` 140 times instead,
because the Gauss-Newton model is a legitimate stopping test again. Neither
changes which optimum is found: matched start by start, `"gn"` and the forward
handover differ by $3 \times 10^{-6}$. By the time a soft stop fires the basin
is chosen, and the handover only decides how the run ends inside it.

# Discussion

The five sources differ in what a single start buys. `"gn"` and the two
fallback runs reach the published optimum from the same number of starts; a
fallback changes only how a run ends, moving stops off the model tests and onto
the gradient, and an `"sr1"` fallback moves more of them than a `"bfgs"` one.

The source that separates is `"sr1"` seeded from the identity. It reaches the
published optimum from several times as many starts as `"gn"` and at a fraction
of the evaluations each one costs.

`parscale`, the diagonal preconditioner, is off by default and stays off here,
and the quasi-Newton seed is not scaled either. Neither has anything to correct
when parameters are already on a logarithmic scale, which is the normal case in
dMod2, and the ill-conditioning of sloppy models lies along rotated directions
[@gutenkunst2007universally] that no diagonal scaling reaches. Scaling the seed
is worse than idle: it reads one curvature from the first accepted pair of a
run and applies it to every direction, which a spectrum spread over decades
does not support.

Three settings are not a set of models, but they are enough to separate what
travels from what does not.

The ordering of the sources does not travel. Gauss-Newton reaches the optimum
for a third of what the cheapest quasi-Newton run pays on Bachmann; on Boehm
`"sr1"` is cheapest, by an order of magnitude with the prior and by a factor of
two without it. The direction is the one the theory gives: where the residuals
at the optimum are small, $J^{\mathsf T} J$ is close to the true Hessian and a
source that infers curvature from steps pays for information it does not need.
Where they are large the omitted term is not, the Gauss-Newton model stops
improving early, and its own tests fire far from the minimum.

Three things do travel. No source reaches an optimum another misses; every
comparison here separates on price, not on which minimum is found. A handover
does not change which basin a run ends in, because by the time a soft stop
fires the basin has been chosen: matched start by start on Bachmann, `"gn"` and
the forward handover differ by $3 \times 10^{-6}$. And a quasi-Newton phase
costs of the order of $n_\theta$ accepted steps before it carries curvature
everywhere, which is why its price per optimum grows with the parameter count.

The two handovers therefore buy different things, and neither buys an optimum.
The forward one changes only where inside a basin a run stops. The reversed one
restores a legitimate stopping test: on Boehm without a prior it turns 156
stagnations into 159 ends on `fvalue` and cuts runs hitting the iteration cap
from 13 to 4.

# Reproducing the numbers

The nine-parameter fits behind the tables and figures are stored in
`dev/optimisation/results.rds` and reused, so this document renders in seconds
and reproduces exactly what it shows; `dev/render-optimisation.R --refit`
recomputes them.

The two heavy comparisons are too expensive for a render and sit beside the
document as data instead, one file each, with the script that produced them:

- `boehm_noprior.rds` comes from `bench_hessianSource.R` under `inst/benchmarks`
  with the environment variable `BOEHM_NOPRIOR` set, and takes minutes on one
  machine.
- `bachmann.rds` comes from `bachmann_hessianSource.R` under `dev/optimisation`,
  first with `BACHMANN_SUBMIT` and then with `BACHMANN_COLLECT`. It needs a
  SLURM cluster: five job arrays of 50 tasks, four fits each.

`bench_hessianSource.R` without that variable is the nine-parameter comparison
of the earlier sections, over the whole catalogue rather than the subset shown
here and with the per-start `stopReason` tables.

# References
